### 2021-01-05 thesis/thesis.tex
move paper into folder as thesis, so I can write up all that I want to write

% Papers should be formatted according to the two-column ACM proceedings
% format.  Each paper should have no more than 12 pages, excluding
% bibliography, in 10pt font. There is no limit on the page count for
% references. Each reference must list all authors of the paper (do not use et
% al). The citations should be in numeric style, e.g., [52]. Submissions should
% be in PDF format and printable on US Letter and A4 sized paper. These
% requirements are all the same as in the previous year.

\def\paperversiondraft{draft}
\def\paperversionblind{blind}
\def\paperversioncamera{camera}

% If no special paper-version is requested, compile in draft mode
\ifx\paperversion\paperversionblind
\else
  \ifx\paperversion\paperversioncamera
  \else
     \def\paperversion{draft}
  \fi
\fi

\def\grammarlyon{on}

\ifx\grammarly\grammarlyon
\def\review{}
\else
\def\review{review,}
\fi

\ifx\paperversion\paperversiondraft
  \documentclass[sigplan,\review anonymous]{acmart}
\fi

\ifx\paperversion\paperversionblind
  \documentclass[sigplan,\review anonymous]{acmart}
\fi

\ifx\paperversion\paperversioncamera
  \documentclass[sigplan]{acmart}\settopmatter{}
\fi

% Most PL conferences are edited by conference-publishing.com. Follow their
% advice to add the following packages.
%
% The first enables the use of UTF-8 as character encoding, which is the
% standard nowadays. The second ensures the use of font encodings that support
% accented characters etc. (Why should I use this?). The mictotype package
% enables certain features 'to­wards ty­po­graph­i­cal per­fec­tion
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}

\usepackage{xargs}
\usepackage{lipsum}
\usepackage[textsize=tiny]{todonotes}
\usepackage{xparse}
\usepackage{xifthen, xstring}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{minted}
\usemintedstyle{bw}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue}
\usepackage{mathpartir}
\usepackage{dirtytalk}
\usepackage[ligature, inference]{semantic}

\usepackage{listings}
\usepackage{xcolor}

%New colors defined below
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

%Code listing style named "mystyle"
% https://www.overleaf.com/project/5fc73ba8b2793a82677493d5
% https://tex.stackexchange.com/questions/200208/how-can-i-reduce-the-line-spacing-in-a-listing
\lstdefinestyle{mystyle}{
  commentstyle=\color{gray},
  keywordstyle=\bfseries,
  numberstyle=\ttfamily\footnotesize\color{codegray},
  basicstyle=\linespread{0.8}\ttfamily,%,\footnotesize,
  breakatwhitespace=false,
  breaklines=true,
  captionpos=b,
  keepspaces=true,
  numbers=left,
  columns=flexible,
  numbersep=5pt,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2
}

%"mystyle" code listing set
\lstset{style=mystyle}

% Use to keep figure in same section
% \usepackage[section]{placeins}

\makeatletter
\font\uwavefont=lasyb10 scaled 652

\newcommand\colorwave[1][blue]{\bgroup\markoverwith{\lower3\p@\hbox{\uwavefont\textcolor{#1}{\char58}}}\ULon}
\makeatother

\ifx\paperversion\paperversiondraft
\newcommand\createtodoauthor[2]{%
\def\tmpdefault{emptystring}
\expandafter\newcommand\csname #1\endcsname[2][\tmpdefault]{\def\tmp{##1}\ifthenelse{\equal{\tmp}{\tmpdefault}}
   {\todo[linecolor=#2,backgroundcolor=#2,bordercolor=#2]{\textbf{#1:} ##2}}
   {\ifthenelse{\equal{##2}{}}{\colorwave[#2]{##1}\xspace}{ \todo[linecolor=#2,backgroundcolor=#2,bordercolor=#2]{\textbf{#1:} ##2}\colorwave[#2]{##1}}}}}
\else
\newcommand\createtodoauthor[2]{%
\def\tmpdefault{emptystring}
\expandafter\newcommand\csname #1\endcsname[2][\tmpdefault]{\def\tmp{##1}\ifthenelse{\equal{\tmp}{\tmpdefault}}
   {}
   {\ifthenelse{\equal{##2}{}}{##1\xspace}{ ##1}}}}
\fi

% Broaden margins to make room for todo notes
\ifx\paperversion\paperversiondraft
  \paperwidth=\dimexpr \paperwidth + 6cm\relax
  \oddsidemargin=\dimexpr\oddsidemargin + 3cm\relax
  \evensidemargin=\dimexpr\evensidemargin + 3cm\relax
  \marginparwidth=\dimexpr \marginparwidth + 3cm\relax
  \setlength{\marginparwidth}{4.6cm}
\fi

% We use the following color scheme
% 
% This scheme is both print-friendly and colorblind safe for
% up to four colors (including the red tones makes it not
% colorblind safe any more)
%
% https://colorbrewer2.org/#type=qualitative&scheme=Paired&n=4

\definecolor{pairedOneLightBlue}{HTML}{a6cee3}
\definecolor{pairedTwoDarkBlue}{HTML}{1f78b4}
\definecolor{pairedThreeLightGreen}{HTML}{b2df8a}
\definecolor{pairedFourDarkGreen}{HTML}{33a02c}
\definecolor{pairedFiveLightRed}{HTML}{fb9a99}
\definecolor{pairedSixDarkRed}{HTML}{e31a1c}

\createtodoauthor{grosser}{pairedOneLightBlue}
% \createtodoauthor{authorTwo}{pairedTwoDarkBlue}
\createtodoauthor{sid}{pairedThreeLightGreen}
\createtodoauthor{authorFour}{pairedFourDarkGreen}
\createtodoauthor{authorFive}{pairedFiveLightRed}
\createtodoauthor{authorSix}{pairedSixDarkRed}

%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format should change 'sigplan' to
%% 'acmsmall'.

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{marvosym}

% [Operational semantics](https://tex.stackexchange.com/questions/271147/how-to-align-the-premises-in-a-semantic)
\renewcommand{\inference}[3][]{%
  \[\begin{array}[b]{@{}lc}
      \\
      \begin{array}[b]{l} #2 \end{array}
      \smash{\raisebox{-0.6\normalbaselineskip}{\scriptsize}} \\
      % \cline{2-2}
      \cmidrule[0.4pt]{1-1}
      \begin{array}[t]{l} #3 \end{array}
      \\
      \\
  \end{array}\]
}

% https://tex.stackexchange.com/questions/198462/the-lightning-symbol-in-an-equation-environment
%\newcommand{\lightning}{\mbox{\Lightning}}
\newcommand{\val}{\texttt{value} }
\newcommand{\unk}{\texttt{unk} }
\newcommand{\absent}{\texttt{absent} }
\newcommand{\case}{\texttt{case} }
\newcommand{\CONS}{\texttt{cons} }
\newcommand{\lzap}{\texttt{lz.ap} }
\newcommand{\lzthunk}{\texttt{lz.thunk} }
\newcommand{\construct}{\texttt{lz.construct} }
\newcommand{\stdcall}{\texttt{std.call} }
\newcommand{\std}{\texttt{std} }
\newcommand{\nofib}{\texttt{nofib} }
\newcommand{\Int}{\texttt{Int} }

\newcommand{\cpp}[1]{\mintinline{cpp}{#1}}
\newcommand{\raw}[1]{\mintinline{text}{#1}}
\newcommand{\denotation}[1]{\left\llbracket {\raw{#1}} \right\rrbracket}

\newcommand{\mlir}{\texttt{MLIR} }
\newcommand{\lz}{\texttt{lz} }
\newcommand{\lzforce}{\texttt{lz.force}}
\newcommand{\lzconstruct}{\texttt{lz.construct}}
\newcommand{\GHC}{\texttt{GHC} }
\newcommand{\ghc}{\GHC}
\newcommand{\Core}{\core}
\newcommand{\core}{\texttt{Core} }

% syntactic elements of our language
\newcommand{\EXPR}{\texttt{EXPR}}
\newcommand{\PHI}{\texttt{PHI}}
\newcommand{\BB}{\texttt{BB}}
\newcommand{\BBID}{\texttt{BBID}}
\newcommand{\INST}{\texttt{INST}}
\newcommand{\TERM}{\texttt{TERM}}
\newcommand{\LABEL}{\texttt{LABEL}}
\newcommand{\FLAG}{\texttt{FLAG}}
\newcommand{\ASSIGN}{\texttt{ASSIGN}}
\newcommand{\VID}{\texttt{VID}}

% transition relations
\newcommand{\assignarrow}{\xrightarrow{~\texttt{asgn}~}}
\newcommand{\phiarrow}{\xrightarrow{phi}}
\newcommand{\controlflowarrow}{\xrightarrow{~\texttt{ctrl}~}}

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2017}
\acmMonth{1}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\else\makeatother
%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
\setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

\begin{document}

%% Title information
\title[Short Title]{Catching fire: Undefined behaviour for equational reasoning of lazy languages with strictness annotations}
                                        %% when present, will be used in
                                        %% header instead of Full Title.
\subtitle{Subtitle}                     %% \subtitle is optional

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{First1 Last1}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}
}
\email{first2.last2@inst2b.org}         %% \email is recommended

\begin{abstract}
% An abstract should consist of six main sentences:
%  1. Introduction. In one sentence, what’s the topic?
%  2. State the problem you tackle.
%  3. Summarize (in one sentence) why nobody else has adequately answered the research question yet.
%  4. Explain, in one sentence, how you tackled the research question.
%  5. In one sentence, how did you go about doing the research that follows from your big idea.
%  6. As a single sentence, what’s the key impact of your research?

% 1
We propose a new style of semantics
for non-strict languages that provide access to many optimizations that are
currently unsound for lazy languages.
% 2 
Lazy languages, while providing equational reasoning, are unable to reason
effectively about \emph{strictness annotations}. On the other hand, most strict
languages are impure, and thus unable to provide equational reasoning due to
the presence of side effects.
% 3
Research on optimizing non-strict languages has focused on optimizing away
non-strictness with demands, and efficient lowering of strictness onto modern
hardware.
% 4
We take a different approach: we propose to use undefined behaviour judiciously
while defining the semantics of our language to enable many key optimizations that
are possible in the strict world, that are not possible in the non-strict world.
We model \emph{divergent computations as undefined behaviour}, giving the
compiler far more freedom to optimize and reorder a mix of strict and non-strict
code.
% 5 
We first consider the various deficiencies of introducing an explicit "bang pattern"
or "strictness annotation" as is done within the Glasgow Haskell compiler today.
Then we demonstrate how setting divergent computations to be undefined behaviour 
cleanly enables the compiler to optimize these programs.
% 6
This modelling of divergence as undefined behaviour enables us to write an
\emph{optimizing} compiler that can reason about a mix of strictness and non
strictness. Furthermore, it enables us to modularly add non-strictness into
an existing strict language. We demonstrate this with a prototype optimizing
functional language called Lizzy. Lizzy uses MLIR as a backend. We introduce
a new dialect \lz in MLIR for modularly expressing non-strictness.  As a benchmark,
we consider a shootout between GHC's optimization of vector and matrix
computations, and \lz's ability to optimize the same computations.
% (http://www.easterbrook.ca/steve/2010/01/how-to-write-a-scientific-abstract-in-six-easy-steps/)
\end{abstract}

% Only add ACM notes and keywords in camera ready version
% Drop citations and footnotes in draft and blind mode.
\ifx\paperversion\paperversioncamera
%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords is optional

\else
\settopmatter{printacmref=false} % Removes citation information below abstract
\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
\fi

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle
\ifx\grammarly\grammarlyon 
\onecolumn 
\else 
\fi

\section{Introduction}

% borrowed from : https://arxiv.org/abs/2003.00532 (Uday's note on MLIR)
Lazy languages such as Haskell
offer many benefits, chief of which is the ability to reason
equationally about programs. For performance, considerations,
the prototypical lazy language, Haskell, provides ``bang patterns'' to
mark strictness. Herein lies the trouble; these ``bang patterns'' do not
permit equational reasoning for the user. Worse, they form a sequence point
in the otherwise pure semantics, taking away most of the optimization freedom
from the compiler. We propose to remedy the situation by making judicious use
of \emph{undefined behaviour semantics}. While undefined behaviour is often
reviled in discourse, it is in fact a major reason why C and C++ can be optimized
as well as they can; defining certain rare or impossible conditions as undefined
behaviour provides the compiler a great deal of freedom by being able to ignore
these cases. We choose to mark \emph{divergence as undefined behaviour}. This
solves the \emph{correctness problem} in reordering strict computations; The
only thing our compiler (Lizzy) will reason about is performance when reordering
strict computations. Furthermore, this is not too insane a proposal; After all,
the C standard also defines side-effect free infinite loops as undefined behaviour.
Next, we show a variety of examples where this simple extension to the language
semantics permits far easier reasoning about mixing strictness and non-strictness.
We then put this to the test, by extending MLIR with a new non-strict IR, called
\lz. Note that the undefined behaviour semantics are critical in order to
reason effectively about the mix of strict and non strict behaviour we wish
to model within MLIR. Finally, to validate our approach, we consider a series
of benchmarks where our programming language, Lizzy outperforms similar programs
in Haskell compiled by the Glasgow Haskell Compiler.

% \begin{figure}
% % Link to figure
% %
% % https://docs.google.com/drawings/d/1juKp43D3rLC-luBQPwQZ_wCnDK2S_6C1k6USV0wKE0g/edit?usp=sharing
% \includegraphics[width=\columnwidth]{images/overview.pdf}
% \caption{Our key idea visualized}
% \end{figure}

% \grosser{Always state your contributions explicitly: (A) this makes it
% easy for the reader to understand what novelty is presented, and (B)
% these contributions help you to focus. In particular, the objectie of
% the remaining paper should be to support the claims stated here.
% }
Our contributions are:                                          
\begin{itemize}
\item A formal semantics for bang patterns in the Haskell language.
\item A study of where bang patterns inhibit GHC's ability to reason about strictness.
\item A novel yet simple extension to a lazy language semantics that remove this
      inability of GHC (and lazy languages in general) to reason about strictness.
\item An implementation of a prototype lazy language (Lizzy) that chooses this
      approach pervasively.
\item A new dialect (\lz) for MLIR to encode these semantics, and provide modular
      non-strictness.
\end{itemize}

\section{Strictness Annotations: The Bad And The Ugly}

We first motivate why strictness annotations inhibit the compiler
from being able to optimize programs. Consider the program in
\autoref{fig:lazy-loop}, where \texttt{out} declares
a binding \raw{x = loop} and proceeds to not use it. This allows GHC
to optimize out the program into \raw{out = 42}. On the
other hand, consider  \autoref{fig:strict-loop} where \raw{out}
declares  \emph{strict} binding \raw{!x  = loop} and proceeds to not
use it.

\begin{lstlisting}[language=haskell,caption=unoptimized \raw{out} and optimized \raw{outOptimized}]
{-# LANGUAGE BangPatterns #-}
loop, out, out' :: Int
loop = loop;
out = let x = loop in 42
outOptimized = 42
main :: IO ()
main = do print out; print out'
\end{lstlisting}

\begin{lstlisting}[language=haskell, caption=unoptimized \raw{outStrict} with an incorrect optimization in \raw{outStrictOptimized}]
{-# LANGUAGE BangPatterns #-}
loop, out, out' :: Int
loop = loop;
outStrict = let !x = loop in 42
outStrictOptimized = 42 -- INCORRECT TRANSFORM
main :: IO ()
main = do print out; print out'
\end{lstlisting}
% \caption{An example of how strictness annotations break equational reasoning}

\begin{lstlisting}[language=haskell, caption=Original haskell program]
{-# LANGUAGE BangPatterns #-}
out b = let large = unknown 10 in 
  if False then large else 42
out' b = let !large = unknown  10 in 
  if False then large else 42
main :: IO ()
main = do print out; print out'
\end{lstlisting}

\begin{lstlisting}[language=Haskell, caption=Correct optimization that is incorrect with a bang pattern]
{-# LANGUAGE BangPatterns #-}
loop, out, out' :: Int
loop = loop;
out b= if b then 
  let large = unknown 10 in large else 42
-- | incorrect. If `large` had been divergent, 
-- then `out` no longer
-- exhibits divergence
out' b = if b then 
  let !large = unknown 10 in large else 42

main :: IO ()
main = do print out; print out'
\end{lstlisting}
% \caption{An example of how strictness annotations disallow floating a \texttt{let}
%          down into the use-site, inhibiting optimization}

\section{Undefined Behaviour For Divergence?}

As we have shown, the existence of bang patterns obviates equational reasoning.
We are now in a bind: we can choose to forego bang patterns, thereby
regaining equational reasoning with a loss of performance. This is not a 
very reasonable suggestion, as bang patterns are critical for controlling
evaluation ordering, and ensuring that we do not build up thunks.

We first propose a simple, outrageous, solution:  Ban all divergence.
We define divergence to be \emph{undefined behaviour}. In essence, this
allows the compiler to assume that \emph{no value} is divergent, for if a
value were divergent, then \emph{anything} can happen. We analyze the
merits and deficiencies of this proposal, and then provide a more
refined version of the above solution.

% \begin{figure}[ht]
% \begin{minted}{hs}
% {-# LANGUAGE BangPatterns #-}
% loop, out, out' :: Int
% outStrict = let !x = loop in 42
% -- ==If x were not divergent==, 
% --   it is safe to set out' to 42.
% -- ==If x were divergent==
% --   It is undefined behaviour. Anything can
% --   happen, including evaluating out' to 42.
% outStrictOptimized = 42 
% main :: IO ()
% main = do print out; print out'
% \end{minted}
% \caption{An example of how defining divergence as UB allows continued
%          equational reasoning with bang patterns.}
% \label{fig:strict-loop-ub}
% \end{figure}

Consider the program in \autoref{fig:strict-loop}. Here, according
to our semantics, we define \emph{divergence as undefined behaviour}.
Interpreted according to these semantics, here's how we legally
allow the transformation \raw{outStrictOptimized = 42}:

\begin{itemize}
    \item Since \raw{x} is forced, we must evaluate \raw{loop} into weak head normal form.
    \item \raw{loop} does not have a weak head normal form. This is a divergent computation.
    \item Since we have divergence, \emph{anything} can happen [we can refine $\bot$ into a
    value that is greater than $\bot$].
    \item We choose to refine $\bot$ into $42$ as $\bot \sqsubset 42$.
\end{itemize}

%\begin{figure}
%\begin{minted}{hs}
\begin{lstlisting}[language=haskell, caption=An example of how defining divergence as UB is too agressive]
{-# LANGUAGE BangPatterns #-}
-- | never evaluates to a value.
main = putStrLn "42"; main
mainOptimized = return () 
 -- | legal if divergence is UB.
\end{lstlisting}
%\end{minted}
%\label{fig:io-loop}
%\caption{An example of how defining divergence as UB is too agressive}
%\end{figure}

On the other hand, this definition is \emph{far too agressive}. Consider
the program in \autoref{fig:io-loop}. Here, main prints $42$ forever,
never returning the final \raw{IO ()} that it is expected to. Thus,
we can \emph{optimize this program} into \raw{mainOptimized = return ()}
if divergence is undefined behaviour. This is not what we want! We wish
to allow programs which \emph{make progess without halting}, while
still being able to exploit undefined behaviour. Thus, we must change our
semantics to adapt to this need.

\section{Progressive operational semantics}

Naively speaking, one maybe tempted to summarize our approach as
"make divergence undefined". While this is appealing, this is both too
strong, and too naive. For example, such a definition would allow us to
optimize away programs that \emph{progress} without \emph{terminating}.
A tiny example of such a program, which prints ``42'' forever is
given in \autoref{fig:print-forever}

% \begin{figure}[ht]
% \begin{minted}{hs}
\begin{lstlisting}[language=haskell, caption=A program that makes progress without terminating.  We wish to preserve such program semantics]
action :: IO (); action = putStrLn "42";
main :: IO (); main = action >> main
\end{lstlisting}
% \end{minted}
% \label{fig:print-forever}
% \caption{A program that makes progress without terminating.  We wish to preserve such program semantics}
% \end{figure}

% \begin{figure}[ht]

%\begin{subfigure}[b]{0.48\textwidth}
%\begin{minted}{hs}
\begin{lstlisting}[language=haskell]
{-# LANGUAGE BangPatterns #-}
loop :: Int; loop = loop
ex1 = let x = loop in 42
ex2 = let !x = loop in 42
ex3 = let !x = loop in x
-- | New program
{-# LANGUAGE BangPatterns #-}
loop :: Int; loop = loop
ex1 = 42
ex2 = 42
ex3 = let !x = loop in x
ex3' = loop -- equivalent semantics
\end{lstlisting}
% \end{minted}
% \subcaption{Maxmially transformed programs with our proposed semantics}
% \end{subfigure}
% \caption{Programs that exhibit a range of behviours, clarifying what values we do and do not eliminate}
% \end{figure}

We call our semantics \emph{progressive}, which tries to make as much progress
as possible before getting stuck at a stuck term.
\section{Formal Semantics Of Bang Patterns in Haskell}

There is currently no formal definition of bang patterns in the Haskell language.
The semantics are defined based on how the Glasgow Haskell Compiler chooses
to manipulate these strictness annotations. So wherever we have previously argued
that it would be ``incorrect'' for GHC to perform these transforms, we have been
arguing from the position of:

\begin{itemize}
\item This is what GHC does today, and that is the current state of definition for bang patterns.
\item Were GHC to make the above transformation, we believe it would violate the
      intuitions that many people have about how GHC \emph{ought} to transform code.
      \sid{I asked bgamari who works on GHC full-time, he said that GHC does not perform the above transform most likely.
           Similarly, other friends (carter, davean) who commit into GHC also said likely not.}
\end{itemize}

To begin our discussion, we first formalize the semantics of bang patterns.
\textbf{TODO: fill in semantics here.}
\section{\lz: Its Syntax, Type system, semantics}
\subsection{A high level overview of \mlir}

\begin{figure*}
\begin{minted}{text}
toplevel := func | global
func := <fn-name> formal-param-list -> <ret-type> <region>
region := list [bb]
bb := <bb-name> formal-param-list : list inst; terminator-inst
inst := retval "=" <op-name> arg-list
terminator-inst := std.return(<name>) | br <name> | condbr <name> <bb1> <param-list> <bb2> <param-list>
\end{minted}
\caption{The \std dialect syntax}

\end{figure*}

\begin{figure*}
\vspace{3em}

\begin{tabular}{l l}
\begin{minipage}[t][1cm][b]{0.5\textwidth}
\inference[assign-cst]
{
}
{
  (\texttt{\%vid =}~ \texttt{std.constant } const, R) ~\assignarrow~R[\texttt{\%vid} \mapsto const]
}
\end{minipage}

&

\begin{minipage}[t][1cm][b]{0.5\textwidth}
\inference[assign-fncall]
{
  R[x_1] = v_1\quad R[x_2] = v_2\quad \dots\quad R[x_n] = v_n \\
  \denotation{f}(v_1, v_2, \dots, v_n) = y \\
}
{
  (\texttt{\%vid =}~ \texttt{f(}x_1, x_2, \dots, x_n \texttt{)}, R) ~\assignarrow~R[\texttt{\%vid} \mapsto y]
}
\end{minipage}
\end{tabular}

% =========================

\vspace{3em}
\begin{tabular}{l l}
\begin{minipage}[t][1cm][b]{0.5\textwidth}
\inference[cbr-left]
{
    % P[pc] = \texttt{cond\_br \%c $\caret$ bb1(...) \^bb2(...)} \quad
    P[pc] = \raw{condbr c ^bb1(xs) ^bb2(ys)} \\
    P[\raw{^bb1}] =  \raw{^bb1(ls)} \\
    R[c] \ne 0
}
{
    (pc, R) ~\controlflowarrow~(pc', R[ls \mapsto R[xs])
}

\end{minipage}
&
\begin{minipage}[t][1cm][b]{0.5\textwidth}

\inference[cbr-right]
{
    P[pc] = \texttt{br}~ c~ l(v_1)~ r(v_2) \quad
    P[r] = pc' \quad
    R[c] = 0
}
{
    (pc, R) ~\controlflowarrow~(pc', R[r.input \mapsto R[v_2])
}
\end{minipage}
\end{tabular}

% =========================

\inference[br]
{
    P[pc] = \texttt{br}~ n(v) \quad
    P[n] = pc'
}
{
    (pc, R) ~\controlflowarrow~(pc', R[n.input \mapsto R[v])
}

\caption{The operational semantics of the \std dialect}
\end{figure*}

\subsection{A high level overview of \lz}
We extend the basic \std dialect modestly, by adding primitives that can 
describe non-strictness. We first introduce a primitive known as \lzthunk,
which is responsible for creating non-strict thunks. 
To represent lazy function application, we introduce a new operation called \lzap,
which is the lazy sibling of \stdcall. This creates a thunk, which when evaluated,
invokes the function with the given arguments. To force
a thunk, we introduce \lzforce, which receives a thunk as input and returns
the forced value of the thunk as output. In effect, we have an encoding of
administrative normal form as (ANF) an MLIR dialect.

To represent algebraic data types, which enable reasoning of control flow 
and are the primary mode of abstraction in most functional languages, as
well as newer imperative languages such as Rust, we introduce the \construct
and \case operators, which are the introduction and elimination forms of
structured data in the dialect.

\begin{figure*}[hbt!]
\begin{minted}{text}
thunk(x): T -> Thunk<T>
ap(f, v1, v2, ..., vn): ((T1,...Tn) -> R) x T1 x ... Tn -> Thunk<R>
force(x): Thunk<T> -> T
construct(ConsName, v1, ..., vn): Symbol x T1 x T2 ... x Tn -> ADT<T>
case(x: T) of ... : ADT<T> -> R
\end{minted}
\caption{Syntax of \lz extensions}
\end{figure*}

\begin{figure*}
\vspace{3em}
\begin{tabular}{l l}
\begin{minipage}[t][1cm][b]{0.5\textwidth}
\inference[asgn-constructor]
{
   P[pc] = \\
    \quad \texttt{\%vid = lz.construct(@ConsName, \%x1, ..., \%xn)} \\
   P[\texttt{\%x1}] = v_1, \quad \dots, \quad P[\texttt{\%xn}] = v_n \\
}
{
  R[\texttt{\%vid} \mapsto \texttt{constructor(@ConsName, $v_1, \dots, v_n$)}]
}
\end{minipage}
&
\begin{minipage}[t][1cm][b]{0.5\textwidth}
\inference[asgn-thunk]
{
   P[pc] = \texttt{\%vid = lz.ap(\%fref, \%x1, ..., \%xn)} \\
   P[\texttt{\%f1}] = \texttt{ref(@fn)} \\
   P[\texttt{\%x1}] = v_1, \quad \dots, \quad P[\texttt{\%xn}] = v_n \\
}
{
  R[\texttt{\%vid} \mapsto \texttt{thunk(@fn, $v_1, \dots, v_n$)}]
}
\end{minipage}
\end{tabular}

% ================

\vspace{5em}
\begin{tabular}{l l}
\begin{minipage}[t][1cm][b]{0.5\textwidth}
\inference[asgn-case]
{
   P[pc] = \texttt{\%vid = lz.case(\%x, @Cons\textsubscript{1}, r\textsubscript{1},}\dots 
                    \texttt{, @Cons\textsubscript{N}, r \textsubscript{N})} \\
   \texttt{@Cons}_i = \texttt{@XCons} \\ 
   P[\texttt{\%x}] = \texttt{construct(@XCons, $v_1, \dots, v_n$)} \\
   r_i = \raw{ { ^entry(arg1, ..., argn): ... } }
}
{
  % decide on how to represent region execution
  (pc, R) \controlflowarrow (P[r_i], R[\texttt{arg}_i \mapsto v_i])
}
\end{minipage}
&
\begin{minipage}[t][1cm][b]{0.5\textwidth}
\inference[asgn-force]
{
   p[pc] = \texttt{\%vid = lz.force(\%thnk)} \\
   p[\texttt{\%thnk}] = \texttt{thunk(@fn, $v_1, \dots, v_n$)} \\
   \denotation{@fn}(v_1, v_2, \dots, v_n) = y
}
{
  r[\texttt{\%vid} \mapsto y]
}
\end{minipage}
\end{tabular}
\caption{Operational semantics of \lz  extensions}
\end{figure*}
\textbf{TODO: this is broken, because it doesn't scope properly. We need to describe \lzthunk
as building a thunk \emph{in memory/on the heap}, and the value stored in the register file
is the pointer to this thunk}.

\subsection{Purifying \lzforce}

Let us for a moment assume that divergence is not undefined behaviour. Now consider
the code in \autoref{fig:unused-divergent-variable-lz}. The variable called
\texttt{\%loopv} is unused. However, because divergence \texttt{\%loopv} is
divergent, it is incorrect to eliminate the variable definition of 
\texttt{\%loopv}, for the divergence of \texttt{loop} occurs when \texttt{\%loopt} is forced;
That is, when the instruction \texttt{\%loopv = lz.force(\%loopv)} is executed.
This implies that the instruction \lzforce is side-effecting.

If we wished to regain purity of \lzforce and leverage the power of
SSA, then we must be allowed to eliminate a call to \lzforce if the
value is unused. This is exactly the same requirement as being able to 
equationally reason with bangs in a \texttt{let} binding. In that use-case,
we wished for undefined behaviour semantics for ease-of-reasoning. Here, we wish
for undefined behaviour semantics for ease-of-optimization.

\begin{lstlisting}[language=c++, caption=The SSA encoding of a lazy program with divergence in the \lz dialect.  Note the unused divergent variable \texttt{\%loopv}]
// loop = loop
lz.func @loop () -> !lz.adt<@Box> {
  %outt = lz.ap(@loop) : !lz.thunk<lz.adt<@Box>> 
  %outv = lz.force(@loop) : !lz.adt<@Box>
  lz.return(%outv) : !lz.value
}

// main = let !x = loop in 42
lz.func @main() -> !lz.adt<@Int> {
  %loopt = lz.ap(@loop) 
    : !lz.thunk<lz.adt<@Box>>
  %loopv = lz.force(%loopv) // <- unused %loopv
    : !lz.adt<@Box>
  %i = std.constant 42 : i64
  %retv = lz.construct(@Int, i)
  lz.return(%retv)
}
\end{lstlisting}
% \caption{The SSA encoding of a lazy program with divergence in the \lz dialect.
%          Note the unused divergent variable \texttt{\%loopv}}
% \label{fig:unused-divergent-variable-lz}
% \end{figure}

%\begin{figure}[hbt]
%\begin{minted}{text}
\begin{lstlisting}[language=c++, caption=Under the assumption that divergence is UB we can eliminate the call to \lzforce. This regains SSA semantics.]
// loop = loop
lz.func @loop () -> !lz.adt<@Box> {
  %outt = lz.ap(@loop) : !lz.thunk<lz.adt<@Box>> 
  %outv = lz.force(@loop) : !lz.adt<@Box>
  lz.return(%outv) : !lz.value
}

// main = let !x = loop in 42
lz.func @main() -> !lz.adt<@Int> {
  // %loopt = lz.ap(@loop) 
  //  : !lz.thunk<lz.adt<@Box>>
  // %loopv = lz.force(%loopv) 
  //   : !lz.adt<@Box> 
  %i = std.constant 42 : i64
  %retv = lz.construct(@Int, i)
  lz.return(%retv)
}
\end{lstlisting}
% \caption{Under the assumption that divergence is UB, we can eliminate the call to
%          \lzforce. Thus, this assumption allows us to regain SSA semantics.}
% \label{fig:unused-divergent-variable-lz-eliminated}
% \end{figure}

\subsection{Purifying \lzconstruct}

Consider the program in \autoref{fig:unused-construct}. For us to eliminate
the call to \texttt{lz.construct}, we need to be sure that memory allocation
has no side effects. In our case, since we assume that allocations are not
visible to the user in terms of their implementation details (eg. The address
of the pointer of the block of memory), we can safely remove the value
\texttt{\%unused = lz.construct(@Box)}  as \texttt{\%unused} is unused.

On the flipside, consider a program where we want to \emph{copy} the 
instruction \texttt{lz.construct}. If we were not guaranteed the existence
of a garbage collector, then this would not be pure; we would have created a
chunk of memory that is never freed. \textbf{TODO: Is there a good example where we 
want to copy a value for optimization? Something like, copy an SSA value both
into a loop and outside or something?}

\begin{lstlisting}[caption=An unused called to \lzconstruct]
lz.func() -> !hask.adt<@Box> {
  %unused = lz.construct(@Box)
  %one = std.constant 1 : i64
  lz.ret(%one) : i64
}
\end{lstlisting}

\section{Aiding the performance programmer with explicit forcing}

\begin{lstlisting}[language=haskell, caption=Haskell program with two datatypes  \texttt{Lazy} and \texttt{Strict}
with the same shape of data but different strictness annotations.]
{-# LANGUAGE BangPatterns #-}
{-# LANGUAGE UnboxedTuples #-}
module Foo(boxLazyAndStrict) where
-- | A large datatype
data Box = Box Int Int Int Int Int
data Lazy = MkLazy Box
data Strict = MkStrict !Box

data BoxPair = MkBoxPair !Box !Box

boxLazyAndStrict :: Lazy 
  -> Strict 
  -> BoxPair
boxLazyAndStrict lz str =
  case lz of
   MkLazy lbox  ->
       case str of
         MkStrict strbox ->
           MkBoxPair lbox strbox
\end{lstlisting}

\begin{lstlisting}[language=haskell,caption=\ghc \core corresponding to the  above Haskell program  where there is no indication
         that \texttt{strbox} is a value while \texttt{lbox} is a thunk]
boxLazyAndStrict :: Lazy -> Strict -> BoxPair
 ...
 Tmpl= \ (lz_aub :: Lazy)
         (str_auc :: Strict) ->
         case lz_aub of { MkLazy lbox_aud ->
         case str_auc of { MkStrict strbox_aue ->
         case lbox_aud of dt_Xve [Occ=Once]
         { Box _ _ _ _ _  ->
         Foo.MkBoxPair dt_Xve strbox_aue
         }}}}]
\end{lstlisting}

In Haskell, Core and STG, all of GHC's intermediate representations, there is
no explicit description of the act of forcing a thunk. The \texttt{case} construct
is responsible for \emph{both} forcing and case analysis. This is a fine state
of affairs when we are in a language where values are forced right before they
are used --- that is, the language is \emph{truly lazy}. However, Haskell currently
supports bang patterns in defining data constructors, which violates this
assumption. We begin with an example of a simple program with two datatypes,
called \texttt{Lazy} and \texttt{Strict}, where the both contain the lazy data
of type \texttt{Box}. \texttt{Strict}, to its name, is actually strict
in its \texttt{Box} field. However, this is completely opaque on reading the
Core: We need to infer from the facts:
\begin{itemize}
\item because \raw{lbox_aud} is \texttt{case}'d on
\item The result of the aforementioned case is called \raw{dt_Xve}.
\item The case is followed by the pointless match \raw{(Box _ _ _ _ _)}.
\item Finally, the \raw{(Foo.MkBoxPair dt_Xve ...)} is built.
\item Thus, the only reason \raw{dt_Xve} exists is to represent the forced \raw{lbox_aud}.
\item So, the value \raw{lbox_aud} is forced into the value \raw{dt_Xve}
\end{itemize}

These infelicities, while slight, add up rapidly when attempting to optimize
more complicated Haskell programs. On the other hand, consider our encoding
in \lz. Since we separate the concept of \emph{forcing} a value from \emph{pattern matching},
we can represent the force necessary as a \texttt{lz.force}, without needlessly
introducing a pattern match. This improves the readability of the source, and
more directly expresses the semantics we are interested in.  While GHC Core
is an amazing intermediate representation for analysis and transformation, it
is sub-par for \emph{consumption by the end-user}. Though the source language has
gotten more complex, \core has changed very little; This is a source of
strength (as the optimizer needs to only reasons about a very minimal language).
It is also a source of obfuscation, as many overlapping ideas are represented
by the same construct: in this case, that of forcing, and that of pattern
matching.

\begin{lstlisting}[language=llvm, caption=\lz dialect encoding of the above program where there is a clear indication of the thunk being forced]
lz.func(%lz: !lz.adt<@Lazy>, 
        %str: !lz.adt<Strict>) {
  %lboxt = lz.extractvalue(%lz, 0)
     : !lz.thunk<!lz.adt<@Box>>
  %strboxv = lz.extractvalue(%str, 1)
     : !lz.adt<@Box>
  %lboxv = lz.force(%lboxt) # FORCE
    : !lz.adt<@Box>
  %retval = lz.construct(@MkBoxPair, 
                         %lboxv, %strboxv)
    : !lz.adt<@BoxPair>
  lz.return(%retval)
}
\end{lstlisting}

\section{Worker wrapper by local rewrites: A first stab}
In this section, we consider a small example program, which we will explain
how to perform a classical transform (the worker-wrapper transform) purely
by using local rewrites. Traditionally, such a transformation is performed
by first performing a \emph{demand analysis} whose results are used
to drive the rewrite.

We explore how this transformation can be achieved \emph{without} the need
for demand analysis by performing \emph{local rewrites}. This provides
both a simpler manifestation of the algorithm, as well as a potentially 
faster implementation, as the MLIR infrastructure is capable of performing
rewrites in parallel.
% \begin{figure}[htb!]
%\begin{minted}[linenos]{cpp}
{\footnotesize
\begin{lstlisting}[caption=initial source code]
SimpleInt f(Thunk<SimpleInt> i) {
    SimpleInt icons = force(i);
    int ihash = casedefault(icons);
    if (ihash <= 0) {
        return SimpleInt(42);
    } else {
        int prev = ihash - 1;
        SimpleInt siprev = SimpleInt(prev);
        Thunk<SimpleInt> siprev_t = thunkify(siprev);
        SimpleInt f_prev_v = apStrict(f, siprev_t);
        return f_prev_v;
    }
}

int main() {
    printf("%d\n", f(thunkify(SimpleInt(1))).v);
}
\end{lstlisting}
}
%\end{minted}
%\caption{Step 0: The source code we optimize using worker wrapper}
%\end{figure}

%\begin{figure}[htb!]
%\begin{minted}{cpp}
{\footnotesize
\begin{lstlisting}[caption=Step 1: outlining everything after the initial \lzforce into \texttt{f2}]
SimpleInt f2(SimpleInt icons) {
    int ihash = casedefault(icons);
    if (ihash <= 0) {
        return SimpleInt(42);
    } else {
        int prev = ihash - 1;
        SimpleInt siprev = SimpleInt(prev);
        Thunk<SimpleInt> siprev_t = thunkify(siprev);
        SimpleInt f_prev_v = apStrict(f, siprev_t);
        return f_prev_v;
    }
}

SimpleInt f(Thunk<SimpleInt> i) {
    SimpleInt icons = force(i);
    SimpleInt ret = apStrict(f2, icons);
    return ret;
}
\end{lstlisting}
}
%\caption{Step 1: outlining everything after the initial \lzforce into \texttt{f2}}
%\end{figure}

%\begin{figure}[htb!]
%\begin{minted}{cpp}
\begin{lstlisting}[language=c++, caption=Step 2: Removing the strict application]
SimpleInt f2(SimpleInt icons) {
    int ihash = casedefault(icons);
    if (ihash <= 0) {
        return SimpleInt(42);
    } else {
        int prev = ihash - 1;
        SimpleInt siprev = SimpleInt(prev);
        // Thunk<SimpleInt> siprev_t = thunkify(siprev);
        // SimpleInt f_prev_v = apStrict(f, siprev_t);
        Thunk<SimpleInt> f_prev_v = apStrict(f2, siprev);
        return f_prev_v;
    }
}

SimpleInt f(Thunk<SimpleInt> i) {
    SimpleInt icons = force(i);
    SimpleInt ret = apStrict(f2, icons);
    return ret;
}
\end{lstlisting}
%\end{minted}
%\caption{Step 2: Removing the strict application}
%\end{figure}

%\begin{figure}[htb!]
{\footnotesize
\begin{lstlisting}[language=c++, caption=Step 3: outline everything after casedefault into f3]
SimpleInt f3(int icons) {
    if (ihash <= 0) {
        return SimpleInt(42);
    } else {
        int prev = ihash - 1;
        SimpleInt siprev = SimpleInt(prev);
        Thunk<SimpleInt> f_prev_v = apStrict(f2, siprev);
        return f_prev_v;
    }
}

SimpleInt f2(SimpleInt icons) {
    int ihash = casedefault(icons);
    SimpleInt ret = apStrict(f3, ihash);
    return ret;
}

SimpleInt f(Thunk<SimpleInt> i) {
    SimpleInt icons = force(i);
    SimpleInt ret = apStrict(f2, icons);
    return ret;
}
\end{lstlisting}
}
%\caption{ Step 3: outline everything after casedefault(icons) into f3}
%\end{figure}

%\begin{figure}[htb!]
{\footnotesize
\begin{lstlisting}[language=c++, caption=replace call \texttt{f2(SimpleInt(prev))} to \texttt{f3(prev)}]
SimpleInt f3(int icons) {
    if (ihash <= 0) {
        return SimpleInt(42);
    } else {
        int prev = ihash - 1;
        Thunk<SimpleInt> f_prev_v = apStrict(f3, prev);
        return f_prev_v;
    }
}
SimpleInt f2(SimpleInt icons) {
    int ihash = casedefault(icons);
    SimpleInt ret = apStrict(f3, ihash);
    return ret;
}
SimpleInt f(Thunk<SimpleInt> i) {
    SimpleInt icons = force(i);
    SimpleInt ret = apStrict(f2, icons);
    return ret;
}
\end{lstlisting}
}
%\caption{Step 4: replace call f2(SimpleInt(prev)) to f3(prev)}
%\end{figure}

%\begin{figure}[htb!]
%% ! {\footnotesize
%% ! \begin{lstlisting}[language=c++]
%% ! SimpleInt f3(int icons) {
%% !     if (ihash <= 0) {
%% !         return SimpleInt(42);
%% !     } else {
%% !         int prev = ihash - 1;
%% !         Thunk<SimpleInt> f_prev_v = apStrict(f3, prev);
%% !         return f_prev_v;
%% !     }
%% ! }
%% ! SimpleInt f(Thunk<SimpleInt> i) {
%% !     SimpleInt icons = force(i);
%% !     int ihash = casedefault(icons);
%% !     SimpleInt ret = apStrict(f3, ihash);
%% !     return ret;
%% ! }
%% ! \end{lstlisting}
%% ! }
%\caption{Step 5: Inline f2 into f. This gives us the worker(f3) and wrapper(f)}
%\end{figure}

\section{Worker wrapper by local rewrites: Non-tail-calls}

Currently implementing the constructor simplification

\section{Worker wrapper by local rewrites: Sum types}

\section{Worker wrapper by local rewrites: Sum types + non-tail-calls}

\section{Demand analysis by rewrites: The full algorithm}

\begin{figure*}
\vspace{3em}
{\footnotesize
\begin{tabular}{l l}
\begin{minipage}[t][1cm][b]{0.5\textwidth}
% \begin{subfigure}[b]{0.48\textwidth}

\inference[ForceOfKnownAp]{
\texttt{\%x = ap(\%f, \%v1, ..., \%vn): !lz.thunk<T>} \\
\texttt{\%y = force(\%x): T}
}{
\texttt{\%y = \%f(\%v1, ..., \%vn): T}
}
\subcaption{force of a known function application: remove laziness}
\end{minipage}

&

\begin{minipage}[t][1cm][b]{0.5\textwidth}
\inference[ForceOfThunkify]{
\texttt{\%x = thunkify(\%v) : !lz.thunk<T>} \\
\texttt{\%y = force(\%x): T}
}{
\texttt{\%y = \%x}
}
\subcaption{force of a thunk: remove laziness}
\end{minipage}
\end{tabular}
}

% =========================
\vspace{20em}

\begin{tabular}{l l}
\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize

\inference{
% foo
\texttt{func f(\%x1, ... \%txi: thunk<T>, ..., \%xn) \{} \\
\texttt{  \%xi = force(\%txi)} \\
\texttt{  \%zi = ... ; \%tzi = thunkify(\%zi)} \\
\texttt{  f(\%y1, ..., \%tzi, ..., \%yn)} \\
\texttt{\}}
}{
\texttt{func f\_strict\_i(\%x1, ... \%xi: T, ..., \%xn) \{} \\
\texttt{   \%zi = ...} \\
\texttt{   f\_strict\_i(\%y1, ..., \%zi, ..., \%yn)} \\
\texttt{\}} \\
\texttt{func f(\%x1, ... \%txi: thunk<T>, ..., \%xn) \{} \\
\texttt{   \%xi = force(\%txi)} \\
\texttt{   \%zi = ...} \\
\texttt{   f\_strict\_i(\%y1, \%y2, ..., \%zi, ..., \%yn)} \\
\texttt{\}}
}
} % end footnotesize
\subcaption{outlining recursive call that is immediately forced}
\end{minipage}
&
\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize

\inference{
\texttt{func f(\%x1, ... \%wrapxi: @ADT, ..., \%xn) \{} \\
\texttt{   \%xi = case (\%wrapxi) [\@Wrapper -> \string^ entry(\%v) \{ return \%v \}]} \\
% \texttt{   \%zi = ...}
% \texttt{   \%wrapzi = construct(@Wrapper, \%zi)}
% \texttt{   f(\%y1, ..., \%wrapzi, ..., \%yn)}
\texttt{\}}
}{
\texttt{func f\_work\_i(\%x1, ... \%xi: T, ..., \%xn) \{} \\
\texttt{   \%zi = ...} \\
\texttt{   f\_strict\_i(\%y1, \%y2, ..., \%zi, ..., \%yn)} \\
\texttt{\}} \\
\texttt{func f(\%x1, ... \%txi: thunk<T>, ..., \%xn) \{} \\
\texttt{   \%xi = case (\%wrapxi) [\@Wrapper -> \string^ entry(\%v) \{ return \%v \}]} \\
\texttt{   \%zi = ...} \\
\texttt{   f\_work\_i(\%y1, ..., \%zi, ..., \%yn)} \\
\texttt{\}}
} % end inference
} % end footnotesize
\subcaption{outlining of recursion of a monovariant wrapper}
\end{minipage}
\end{tabular}

\vspace{10em}

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
\texttt{\%x = constructor(@Constructor, \%v1, ..., \%vm)} \\
\texttt{\%y = case \%x} \\
\texttt{      [\@C1 -> \{\string^entry1(\%z11, ..., \%z1n1): ...  \}]} \\
\texttt{      [...]} \\
\texttt{      [\@Ci -> \{\string^entry1(\%zi1, ..., \%zini): ...  \}]} \\
}{
\texttt{Inline \string^entryi with \%zik = \%vk}
}
} % end footnotesize
\subcaption{Case of known constructor: remove indirection}
\end{minipage}

\vspace{10em}

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
FOO
}{
BAR
}
} % end footnotesize
\subcaption{Inline a function call}
\end{minipage}
\caption{Local rewrites performed to eliminate laziness (1)}
\end{figure*}

\begin{figure*}

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
FOO
}{
BAR
}
} % end footnotesize
\subcaption{Outline recursive call of constructor that is immediately unwrapped}
\end{minipage}
% OutlineRecursiveApEagerOfConstructorPattern

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
FOO
}{
BAR
}
} % end footnotesize
\subcaption{Outline pattern matching branches on a function input}
% OutlineCaseOfFnInput
\end{minipage}

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
FOO
}{
BAR
}
} % end footnotesize
\subcaption{Outline return of constructor}
% outline return of constructor
\end{minipage}

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
FOO
}{
BAR
}
} % end footnotesize
\subcaption{Peel constructor from case}
% peel constructor from case
\end{minipage}

%% % \begin{minted}{text}
%% % Inlining
%% % \end{minted}
\caption{Local rewrites performed to eliminate laziness (2)}
\end{figure*}

\section{Evaluation}

As a baseline, we compare our performance on \nofib test-suite, which is the
test suite that GHC is tested with and performance improvements to GHC 
are reported on. We only consider a representative subset of programs from \nofib;
The full test suite also extensively tests Haskell's semantics of parallelism, 
software transactional memory, and other features that are orthogonal to the
problem of representing and optimizing non strictness.

\section{Related Work}

\subsection{Demand analysis}
\cite{wadler1987projections} ``Projections for demand analysis'' performs
demand analysis computation by modelling demands as projections on the semantic
domain. We use their broad framework, adapted to our setting.

\cite{grafcall}, \cite{breitner2014call} extend the demand analysis with
finer grained information, such as Call Arity analysis.

\subsection{GHC's intermediate representation}

The Glasgow Haskell compiler \cite{jones1993glasgow} is a mature optimizing
compiler for Haskell, whose plugin infrastructure we hook into to  develop a test
set for \lz, and whose demand analysis we use as a benchmark to measure against.

\subsection{Intel Haskell research compiler IR}
The Intel labs Haskell Research compiler \cite{liu2013intel} models Core in a
strict ANF language. The compiler performs demand analysis and abstract
simplification on ANF. The demand analysis is performed using traditional
abstract interpretation techniques. Later, this demand information is used
to interpret the program and perform abstract simplification. (TODO: figure out
details of this step).

It then compiles to an intermediate representation called MIL,
which is a loosely typed CFG based, SSA-lite intermediate representation.  They
represent laziness as heap values, and manipulate the heap. Thus, their
representation and usage of lazy values reasons with memory, instead of
providing value semantics.
MIL also does not have a notion of nested regions. Therefore, MIL extends
the traditional control flow controls with more finer-grained information,
called as \texttt{cut} and \texttt{interproc}.

\subsection{Mixing dataflow analysis and transformation}

This is based on the theory of compositional dataflow analysis
and transformations, \cite{lerner2002composing} 
which proves that we can interleave dataflow analyses and transformations
safely.

Hoopl \cite{ramsey2010hoopl} is the dataflow analysis and transformation library
within GHC. However, GHC does not use this for performing dataflow analysis
over Core, as the Hoopl library is designed to work with a CFG based
intermediate representation, while GHC Core is a typed functional, expression
oriented intermediate representation. While Hoopl is used for certain simplifications
within C-- in GHC, it is not used extensively due to  poor performance characteristics
exhibited by the implementation. (TODO: bench Hoopl on contrived examples)

\subsection{Alternative encodings of laziness}

GRIN  \cite{boquist1996grin} is an alternative intermediate representation
for lazy and strict functional programming languages which explicitly represents
heap manipulation and case analysis. However, it is not SSA based.

\subsection{Theoretical justification for our encoding}

Our encoding is based on call-by-push-value \cite{levy2012call}, which
breaks down call-by-value and call-by-name paradigms into simple primitives. In
our dialect \lz, we expose these primitives to enable a clean mixture of
call by name and call by value. This enjoys the many properties that
are proven to hold for call-by-push-value (TODO: which ones?!)

\subsection{Technology substrate}

MLIR \cite{lattner2020mlir} is a compiler framework for building reusable and
extensible compiler infrastructure.  MLIR aims to address software
fragmentation, improve compilation for heterogeneous hardware, significantly
reduce the cost of building domain specific compilers, and aid in connecting
existing compilers together.

\section{Conclusion}

We provide a baseline implementation of non-strictness for the MLIR framework
which with very minimal extensions, allows us to fully express non strict semantics
while being compatible with the rest of MLIR's strict dialects.
As an experience report, we explore MLIR's strengths and weaknesses at representing
non-strictness within an SSA based framework. We also provide a baseline implementation
of demand analysis, as inspired by Wadler and Hugh's ``projections for demand
analysis''. We evaluate our example programs against the nofib benchmark test suite.
We also provide provocative examples where the potent combination of MLIR's 
loop optimization infrastructure along with our modest extensions allows us to
beat native haskell's performance.

\section{TODO}

\subsection{Performance benchmarking}
\subsection{Correct encoding for mutual recursion}
\subsection{Correct encoding for partially applied functions/closures}

We currently have a \texttt{lz.lambda} but it's basically untested, because
we didn't have many things that used closures. We need to make sure this
works properly, and encodes data correctly.

\subsubsection{Lambdas in GRIN}
GRIN is also a low level IR, so it's useful to recall how this is encoded
within GRIN. To quite the GRIN thesis:

\say{
end. Using hbcc we get well optimised
code in a low level functional style, comparable to for example the Core lan-
guage [PJ96] used by the Glasgow Haskell compiler.  The code is lambda lifted,
i.e., has only super combinators, and most high level Haskell constructions,
like overloading, have been transformed away.
}

\subsubsection{Implementing lambda lifting}
Recall that optimal lambda lifting is $O(n^2)$. I don't know the
algorithm; I'd have to study it before I implement it.

\subsection{Stretch: rewrite based unification}
\subsection{Stretch: rewrite based tabled typeclass generation}
\subsection{Stretch: As static as possible garbage collection}

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography

%% why does this not show?
\bibliography{references}

%% Appendix
\appendix
\section{Appendix}

Text of appendix

\end{document}

### 2021-01-06 hask-opt/Interpreter.cpp
Actually make sure the test round-trips, and that we eliminate all forces
-- Testing: 28 tests, 8 workers --
PASS: HASK_OPT :: Lowering/lower-args.mlir (1 of 28)
PASS: HASK_OPT :: Lowering/lower-ap.mlir (2 of 28)
PASS: HASK_OPT :: Lowering/lower-ap-ret-memref.mlir (3 of 28)
PASS: HASK_OPT :: Lowering/lower-ap-capture-memref.mlir (4 of 28)
PASS: HASK_OPT :: Lowering/lower-apeager.mlir (5 of 28)
PASS: HASK_OPT :: Lowering/lower-ap-and-force.mlir (6 of 28)
PASS: HASK_OPT :: Lowering/lower-add.mlir (7 of 28)
PASS: HASK_OPT :: Lowering/lower-ap-and-force-2.mlir (8 of 28)
PASS: HASK_OPT :: Lowering/

@@ -462,9 +462,17 @@ struct Interpreter {
         args.push_back(env.lookup(call.getLoc(), operand));
       }
       // TODO: generalize to more than 1 result
-      assert(call.getNumResults() == 1);
-      env.addNew(call.getResult(0),
-                 *interpretFunction(call.getCallee().str(), args));
+      // assert(call.getNumResults() == 1);
+
+      Optional<InterpValue> res = interpretFunction(call.getCallee().str(), args);
+
+      if (call.getNumResults() == 1) {
+          assert(res && "call has a result but function returned no result!");
+          env.addNew(call.getResult(0), *res);
+      } else {
+          assert(!res && "call has no results but function returned a result!");
+      }
+                 
       return;
     }
 
@@ -752,11 +760,29 @@ struct Interpreter {
     }
   }
 
+
+  void interpretPrimopPrintInt(ArrayRef<InterpValue> args) {
+    llvm::errs().changeColor(llvm::raw_fd_ostream::GREEN);
+    llvm::errs() << "--interpreting primop:|printInt|--\n";
+    llvm::errs().resetColor();
+    
+    assert(args.size() == 1 && "printInt expects single argument");
+    InterpValue v = args[0];
+    assert(v.type == InterpValueType::I64 && "printInt expects int argument");
+    // vvv is this a hack? Maybe.
+    llvm::outs() << v.i() << "\n";
+  };
+
   llvm::Optional<InterpValue> interpretFunction(std::string funcname,
                                                 ArrayRef<InterpValue> args) {
     llvm::errs().changeColor(llvm::raw_fd_ostream::GREEN);
     llvm::errs() << "--interpreting function:|" << funcname.c_str() << "|--\n";
     llvm::errs().resetColor();
+
+    if (funcname == "printInt") {
+        interpretPrimopPrintInt(args);
+        return {};
+    }
 
     // functions are isolated from above; create a fresh environment
     if (FuncOp haskfn = module.lookupSymbol<FuncOp>(funcname)) {

### 2021-01-06 hask-opt/Runtime.cpp

@@ -2,7 +2,6 @@
 #include <stdlib.h>
 
 extern "C" {
-
 #define DEBUG_LOG                                                              \
   if (1) {                                                                     \
     DEBUG_INDENT();                                                            \
@@ -11,7 +10,7 @@ extern "C" {
 
 #ifndef NODEBUG
 
-[[maybe_unused]] static int DEBUG_STACK_DEPTH = 0;
+static int DEBUG_STACK_DEPTH = 0;
 void DEBUG_INDENT();
 
 void DEBUG_PUSH_STACK();
@@ -133,7 +132,7 @@ void *__attribute__((used, always_inline)) evalClosure(void *closure_voidptr) {
     FnTwoArgs f = (FnTwoArgs)(c->fn);
     ret = f(c->args[0], c->args[1]);
   } else {
-    assert(false && "unhandled function arity");
+    assert(0 && "unhandled function arity");
   }
 #ifndef NODEBUG
   DEBUG_POP_STACK();
@@ -193,9 +192,9 @@ void *extractConstructorArg(void *__restrict__ cptr, int i) {
   return v;
 }
 
-bool isConstructorTagEq(void *__restrict__ cptr, const char *tag) {
+int isConstructorTagEq(void *__restrict__ cptr, const char *tag) {
   Constructor *c = (Constructor *)cptr;
-  const bool eq = !strcmp(c->tag, tag);
+  const int eq = !strcmp(c->tag, tag);
 #ifndef NODEBUG
   DEBUG_LOG;
   fprintf(stderr, "(%p:%s, %s) -> %d\n", cptr, c->tag, tag, eq);
@@ -230,5 +229,12 @@ void printConstructor(void *v, const char *fmt) {
   fmt = printConstructorGo(v, fmt);
   assert(fmt[0] == 0 && "did not consume format string entirely!");
   printf("\n");
+}
+
+I64Memref *unboxI64Memref(void *v) {
+  I64Memref *mem = (I64Memref *)malloc(sizeof(I64Memref));
+  mem->p = mem->q = mem->s = mem->t = 0;
+  mem->r = 42;
+  return mem;
 }
 } // end extern C

### 2021-01-06 include/Runtime.h

@@ -6,11 +6,16 @@
 extern "C" {
 
 static const int MAX_CLOSURE_ARGS = 10;
-struct Closure {
+typedef struct {
+  void *p, *q;
+  int r;
+  void *s, *t;
+} I64Memref;
+typedef struct {
   int n;
   void *fn;
   void *args[MAX_CLOSURE_ARGS];
-};
+} Closure;
 
 char *getPronouncableNum(size_t N);
 char *getPronouncablePtr(void *N);
@@ -22,6 +27,8 @@ void *__attribute__((used)) mkClosure_capture0_args0(void *fn);
 void *identity(void *v);
 void *__attribute__((used)) mkClosure_thunkify(void *v);
 
+void *__attribute__((used)) boxMemref(void *v);
+
 typedef void *(*FnZeroArgs)();
 typedef void *(*FnOneArg)(void *);
 typedef void *(*FnTwoArgs)(void *, void *);
@@ -29,17 +36,17 @@ typedef void *(*FnTwoArgs)(void *, void *);
 void *__attribute__((used)) evalClosure(void *closure_voidptr);
 
 static const int MAX_CONSTRUCTOR_ARGS = 2;
-struct Constructor {
+typedef struct {
   const char *tag; // inefficient!
   int n;
   void *args[MAX_CONSTRUCTOR_ARGS];
-};
+} Constructor;
 
 void *__attribute__((used)) mkConstructor0(const char *tag);
 void *__attribute__((used)) mkConstructor1(const char *tag, void *a);
 void *__attribute__((used)) mkConstructor2(const char *tag, void *a, void *b);
 void *extractConstructorArg(void *cptr, int i);
-bool isConstructorTagEq(void *cptr, const char *tag);
+int isConstructorTagEq(void *cptr, const char *tag);
 void printInt(int i);
 void printConstructor(void *v, const char *fmt);
 } // end extern C

### 2021-01-06 make-runtime.sh

@@ -1,9 +1,0 @@
-#!/usr/bin/env bash
-set -e
-set -o xtrace
-clang++ -DNODEBUG -O2 -S -emit-llvm -I ./include/ ./hask-opt/Runtime.cpp -o runtime.ll
-sed -i "s/attributes #3 = {/attributes #3 = {norecurse alwaysinline argmemonly /g" runtime.ll
-sed -i "s/attributes #2 = {/attributes #2 = {norecurse alwaysinline argmemonly /g" runtime.ll
-sed -i "s/attributes #6 = {/attributes #6 = {norecurse alwaysinline argmemonly /g" runtime.ll
-sed -i "s/attributes #7 = {/attributes #7 = {norecurse alwaysinline argmemonly /g" runtime.ll
-

### 2021-01-06 run-optimized.sh

@@ -1,10 +1,10 @@
 #!/usr/bin/env bash
 set -e
 clang++ -DNODEBUG -O2 -S -emit-llvm -I /home/bollu/work/mlir/lz/include/ /home/bollu/work/mlir/lz/hask-opt/Runtime.cpp -o runtime.ll
-sed -i "s/attributes #3 = {/attributes #3 = {norecurse alwaysinline argmemonly /g" runtime.ll
-sed -i "s/attributes #2 = {/attributes #2 = {norecurse alwaysinline argmemonly /g" runtime.ll
-sed -i "s/attributes #6 = {/attributes #6 = {norecurse alwaysinline argmemonly /g" runtime.ll
-sed -i "s/attributes #7 = {/attributes #7 = {norecurse alwaysinline argmemonly /g" runtime.ll
+# sed -i "s/attributes #3 = {/attributes #3 = {norecurse alwaysinline argmemonly /g" runtime.ll
+# sed -i "s/attributes #2 = {/attributes #2 = {norecurse alwaysinline argmemonly /g" runtime.ll
+# sed -i "s/attributes #6 = {/attributes #6 = {norecurse alwaysinline argmemonly /g" runtime.ll
+# sed -i "s/attributes #7 = {/attributes #7 = {norecurse alwaysinline argmemonly /g" runtime.ll
 
 hask-opt $@ --lz-worker-wrapper --lz-lower --convert-scf-to-std --ptr-lower -o /tmp/out.mlir
 mlir-translate /tmp/out.mlir --mlir-to-llvmir -o /tmp/out.ll

### 2021-01-06 run-optimized.sh
OK, we do get squeaky clean LLVM

@@ -6,7 +6,7 @@ clang++ -DNODEBUG -O2 -S -emit-llvm -I /home/bollu/work/mlir/lz/include/ /home/b
 # sed -i "s/attributes #6 = {/attributes #6 = {norecurse alwaysinline argmemonly /g" runtime.ll
 # sed -i "s/attributes #7 = {/attributes #7 = {norecurse alwaysinline argmemonly /g" runtime.ll
 
-hask-opt $@ --lz-worker-wrapper --lz-lower --convert-scf-to-std --ptr-lower -o /tmp/out.mlir
+hask-opt $@ --lz-worker-wrapper --affine-loop-fusion --canonicalize --lz-lower --convert-scf-to-std --ptr-lower -o /tmp/out.mlir
 mlir-translate /tmp/out.mlir --mlir-to-llvmir -o /tmp/out.ll
 llvm-link /home/bollu/work/mlir/lz/runtime.ll /tmp/out.ll -S -o /tmp/out-linked.ll
 opt -always-inline -O3 -S /tmp/out-linked.ll -o /tmp/out-linked-o3.ll

### 2021-01-06 test/bench-vector/Tridiag.hs
Benchmark delta between C++ and haskell. C++ numbers seem too good to be true.
[I] /home/bollu/work/mlir/lz/test/bench-vector > ./tridiagc.out
Run on (8 X 3400 MHz CPU s)
2021-01-06 18:48:05
***WARNING*** CPU scaling is enabled, the benchmark real time measurements may be noisy and will incur extra overhead.
***WARNING*** Library was built as DEBUG. Timings may be affected.
--------------------------------------------------
Benchmark           Time           CPU Iterations
--------------------------------------------------
BM_tridiag         26 us         26 us      27865

[I] /home/bollu/work

@@ -1,0 +1,37 @@
+import Gauge.Main
+import Data.Vector.Unboxed as V
+import Data.Word
+import System.Random.Stateful
+
+tridiag :: (Vector Double, Vector Double, Vector Double, Vector Double)
+            -> Vector Double
+{-# NOINLINE tridiag #-}
+tridiag (as,bs,cs,ds) = V.prescanr' (\(c,d) x' -> d - c*x') 0
+                      $ V.prescanl' modify (0,0)
+                      $ V.zip (V.zip as bs) (V.zip cs ds)
+    where
+      modify (c',d') ((a,b),(c,d)) = 
+                   let id = 1 / (b - c'*a)
+                   in
+                   id `seq` (c*id, (d-d'*a)*id)
+
+
+
+useSize :: Int
+useSize = 2000000
+
+useSeed :: Int
+useSeed = 42
+
+indexFindThreshold :: Double
+indexFindThreshold = 2e-5
+
+main :: IO ()
+main = do
+  gen <- newIOGenM (mkStdGen useSeed)
+  let randomVector l = V.replicateM l (uniformDoublePositive01M gen)
+  as <- randomVector useSize
+  bs <- randomVector useSize
+  cs <- randomVector useSize
+  ds <- randomVector useSize
+  defaultMain $ [bench "tridiag" $ whnf tridiag (as,bs,cs,ds)]

### 2021-01-06 test/bench-vector/tridiag.cpp

@@ -1,0 +1,65 @@
+#include <benchmark/benchmark.h>
+#include <stdlib.h>
+typedef struct { double ds[2]; } pair;
+typedef struct { double ds[4]; } quad;
+
+void tridiag(const double *__restrict__ as, 
+        const double * __restrict__ bs, 
+        const double *__restrict__ cs, 
+        const double * __restrict__ ds, 
+        double * __restrict__ out, 
+        const int len) {
+    pair *scanl = new pair[len];
+    // state of prescanl
+    double state_scanl[2];
+    for(int i = 0; i < len; ++i) {
+        // c' = stated
+        // d' = stated
+        scanl[i].ds[0] = state_scanl[0];
+        scanl[i].ds[1] = state_scanl[1];
+        double id = 1/ (bs[i] - state_scanl[0]*as[i]);
+
+        double newd[2];
+        newd[0] =  cs[i] * id; 
+        newd[1] =  (ds[i] - state_scanl[1] * as[i]) * id;
+        state_scanl[0] = newd[0];
+        state_scanl[1] = newd[1];
+    }
+
+    double state_scanr = 0;
+    for(int i = 0; i < len; ++i) {
+        out[i] = state_scanr;
+        state_scanr = scanl[i].ds[1] - scanl[i].ds[0]*state_scanr;
+    }
+}
+
+
+static const int useSize = 2000000;
+static const int useSeed = 42;
+double *newArray(int len, unsigned short xsubi[3]) {
+    double *xs= new double[len];
+    for(int i = 0; i  < len;++i) {
+        xs[i] = erand48(xsubi);
+    }
+    return xs;
+}
+void BM_tridiag(benchmark::State& state) {
+    unsigned short xsubi[3];
+    xsubi[0] = xsubi[1] = 0;
+    xsubi[2] = useSeed;
+
+    double *as = newArray(useSize, xsubi);
+    double *bs = newArray(useSize, xsubi);
+    double *cs = newArray(useSize, xsubi);
+    double *ds = newArray(useSize, xsubi);
+    double *out = new double[useSize];
+
+    for (auto _ : state) {
+        tridiag(as, bs, cs, ds, out, useSize);
+    }
+}
+
+BENCHMARK(BM_tridiag)->Unit(benchmark::kMicrosecond);
+// BENCHMARK(BM_tridiag)->Unit(benchmark::kMillisecond);
+// Run the benchmark
+BENCHMARK_MAIN();

### 2021-01-06 lib/GRIN/GRINTransforms.cpp
[WIP] making tridiag in MLIR correct. Currently it crashes in parsing a case somewhere.
==20217==ERROR: AddressSanitizer: SEGV on unknown address 0x00007a00001a (pc 0x7f7e1f6d9147 bp 0x000000000000 sp 0x7fffa7093fc0 T0)
==20217==The signal is caused by a WRITE memory access.
    #0 0x7f7e1f6d9147 in mlir::IROperand<mlir::OpOperand, mlir::detail::OpaqueValue>::insertIntoCurrent() /home/bollu/work/mlir/llvm-project/mlir/include/mlir/IR/UseDefLists.h:203
    #1 0x7f7e1f6d9147 in mlir::IROperand<mlir::OpOperand, mlir::detail::OpaqueValue>::IROperand(mlir::Operation*, mlir::detail::OpaqueValue) /ho

@@ -1,0 +1,6 @@
+// https://github.com/grin-compiler/grin
+// EvaluatedCaseElimination
+// TrivialCaseOptimization
+// SparseCaseOptimization
+// UpdateElimination
+

### 2021-01-06 lib/Hask/HaskOps.cpp
[WIP] continue encoding. Writing MLIR by hand is quite painful

@@ -209,7 +209,12 @@ ParseResult CaseOp::parse(OpAsmParser &parser, OperationState &result) {
   assert(altRegions.size() > 0);
 
   HaskReturnOp retFirst =
-      cast<HaskReturnOp>(altRegions[0]->getBlocks().front().getTerminator());
+      dyn_cast<HaskReturnOp>(altRegions[0]->getBlocks().front().getTerminator());
+  if (!retFirst) {
+      assert(false && "expected caseop to end in lz.return");
+      return failure();
+
+  }
   for (int i = 1; i < (int)altRegions.size(); ++i) {
     HaskReturnOp ret =
         cast<HaskReturnOp>(altRegions[i]->getBlocks().front().getTerminator());

### 2021-01-07 include/Pointer/PointerDialect.h
[WIP] lowering; This is so fucking stupid, I need to teach MLIR to lower templates

@@ -61,6 +61,20 @@ public:
   void print(OpAsmPrinter &p);
 };
 
+
+// %ptr = doubletoptr %i
+class DoubleToPtrOp
+    : public Op<DoubleToPtrOp, OpTrait::OneResult, OpTrait::OneOperand> {
+public:
+  using Op::Op;
+  static StringRef getOperationName() { return "ptr.double2ptr"; };
+  // static ParseResult parse(OpAsmParser &parser, OperationState &result);
+  static void build(mlir::OpBuilder &builder, mlir::OperationState &state,
+                    Value vint);
+  void print(OpAsmPrinter &p);
+};
+
+
 // %ptr = ptrtoint %i
 class PtrToIntOp
     : public Op<PtrToIntOp, OpTrait::OneResult, OpTrait::OneOperand> {

### 2021-01-07 lib/Hask/HaskLowering.cpp

@@ -108,6 +108,13 @@ public:
       return mlir::FunctionType::get(fnty.getContext(), argtys, rettys);
     });
 
+
+    addConversion([&](mlir::MemRefType memref) -> Type {
+        return mlir::MemRefType::get(memref.getShape(),
+                                   this->convertType(memref.getElementType()),
+                                   memref.getAffineMaps());
+    });
+
     // lz.value -> !ptr.void
     /*
     addTargetMaterialization([&](OpBuilder &rewriter, ptr::VoidPtrType resultty,
@@ -197,7 +204,11 @@ public:
           builder.create<ptr::MemrefToVoidPtrOp>(builder.getUnknownLoc(), src);
       return op;
     }
-
+    if (src.getType().isa<FloatType>()) {
+      ptr::DoubleToPtrOp op =
+          builder.create<ptr::DoubleToPtrOp>(builder.getUnknownLoc(), src);
+      return op;
+    }
     llvm::errs() << "ERROR: unknown type: |" << src.getType() << "|\n";
     assert(false && "unknown type to convert to void pointer");
   };
@@ -885,99 +896,6 @@ public:
   }
 };
 
-/*
-class I64ToI8PtrConversionPattern : public ConversionPattern {
-public:
-  explicit I64ToI8PtrConversionPattern(TypeConverter &tc)
-      : ConversionPattern(1, tc, MatchAnyOpTypeTag()) {}
-
-  LogicalResult
-  matchAndRewrite(Operation *op, ArrayRef<Value> operands,
-                  ConversionPatternRewriter &rewriter) const override {
-    bool changed = false;
-    for (int i = 0; i < (int)op->getNumResults(); ++i) {
-      LLVM::LLVMIntegerType opResultTy =
-          op->getResult(i).getType().dyn_cast<mlir::LLVM::LLVMIntegerType>();
-      if (!opResultTy) {
-        continue;
-      }
-
-      // TODO: how do I check that the use site wants an I64Ptr?
-      return failure();
-    }
-
-    return changed ? success() : failure();
-  }
-};
-*/
-
-// Keep the old version of the class around because it has details about
-// function pointers.
-// vvvvvvvv
-// class ApOpConversionPattern : public ConversionPattern {
-// public:
-//   explicit ApOpConversionPattern(TypeConverter &tc, MLIRContext *context)
[…]

### 2021-01-07 lib/Pointer/PointerDialect.cpp

@@ -63,6 +63,7 @@ PtrDialect::PtrDialect(mlir::MLIRContext *context)
   addOperations<IntToPtrOp, PtrToIntOp, PtrStringOp, FnToVoidPtrOp, PtrUndefOp>();
   // addOperations<PtrToHaskValueOp, HaskValueToPtrOp>();
   addOperations<PtrToMemrefOp>();
+  addOperations<DoubleToPtrOp>();
   addOperations<MemrefToVoidPtrOp>();
   addTypes<VoidPtrType, CharPtrType>();
 
@@ -110,6 +111,26 @@ void IntToPtrOp::build(mlir::OpBuilder &builder, mlir::OperationState &state,
 void IntToPtrOp::print(OpAsmPrinter &p) {
   p.printGenericOp(this->getOperation());
 };
+
+// === DOUBLE TO PTR ===
+// === DOUBLE TO PTR ===
+// === DOUBLE TO PTR ===
+// === DOUBLE TO PTR ===
+// === DOUBLE TO PTR ===
+
+void DoubleToPtrOp::build(mlir::OpBuilder &builder, mlir::OperationState &state,
+                       Value v) {
+  FloatType ft = v.getType().dyn_cast<FloatType>();
+  assert(ft && "expected argument to be of float type");
+  state.addOperands(v);
+  state.addTypes(VoidPtrType::get(builder.getContext()));
+};
+
+void DoubleToPtrOp::print(OpAsmPrinter &p) {
+  p.printGenericOp(this->getOperation());
+};
+
+
 
 // === FNPTR TO VOID PTR ===
 // === FNPTR TO VOID PTR ===

### 2021-01-07 lib/Hask/HaskLowering.cpp
get stuff lowered to ptr

@@ -1020,6 +1020,24 @@ struct AllocOpLowering : public mlir::ConversionPattern {
   }
 };
 
+struct StoreOpLowering : public mlir::ConversionPattern {
+  explicit StoreOpLowering(TypeConverter &tc,
+                           MLIRContext *context)
+      : ConversionPattern(mlir::StoreOp::getOperationName(), 1, tc, context) {}
+  LogicalResult
+  matchAndRewrite(Operation *op, ArrayRef<Value> rands,
+                  ConversionPatternRewriter &rewriter) const override {
+    using namespace mlir::LLVM;
+    mlir::StoreOp store = cast<mlir::StoreOp>(op);
+    rewriter.setInsertionPointAfter(store);
+    llvm::errs() << "rands.size()|" << rands.size() << "|\n";
+    assert(rands.size() >= 3 && "store needs memref, value to store, and store indeces");
+    rewriter.replaceOpWithNewOp<mlir::StoreOp>(store, rands[0], rands[1], rands.drop_front(2));
+    return success();
+  }
+};
+
+
 
 
 
@@ -1122,9 +1140,15 @@ struct LowerHaskPass : public Pass {
           return isTypeLegal(op.getType());
     });
 
+    target.addDynamicallyLegalOp<StoreOp>([](mlir::StoreOp store) {
+      return isTypeLegal(store.getMemRefType()) &&
+          isTypeLegal(store.getValueToStore().getType());
+    });
 
 
-      HaskTypeConverter typeConverter(&getContext());
+
+
+    HaskTypeConverter typeConverter(&getContext());
     mlir::OwningRewritePatternList patterns;
 
     // OK why is it not able to legalize func? x(
@@ -1151,6 +1175,8 @@ struct LowerHaskPass : public Pass {
 
     // Memref ops? is this really how I'm supposed to do this?
     patterns.insert<AllocOpLowering>(typeConverter, &getContext());
+    patterns.insert<StoreOpLowering>(typeConverter, &getContext());
+
     ::llvm::DebugFlag = true;
 
     // applyPartialConversion | applyFullConversion

### 2021-01-07 lib/Pointer/PointerDialect.cpp
lower to LLVM

@@ -514,6 +514,21 @@ public:
   }
 };
 
+struct DoubleToPtrOpLowering : public ConversionPattern {
+public:
+  explicit DoubleToPtrOpLowering(TypeConverter &tc, MLIRContext *context)
+      : ConversionPattern(DoubleToPtrOp::getOperationName(), 1, tc, context) {}
+
+  LogicalResult
+  matchAndRewrite(Operation *rator, ArrayRef<Value> rands,
+                  ConversionPatternRewriter &rewriter) const override {
+    Value i = rands[0];
+    rewriter.replaceOpWithNewOp<LLVM::BitcastOp>(
+        rator, getInt8PtrTy(rator->getContext()), i);
+    return success();
+  }
+};
+
 
 FunctionType convertFunctionType(FunctionType fnty, TypeConverter &tc,
                                  OpBuilder &builder) {
@@ -748,6 +763,9 @@ struct LowerPointerPass : public Pass {
 
     patterns.insert<StringOpLowering>(typeConverter, &getContext());
     patterns.insert<Int2PtrOpLowering>(typeConverter, &getContext());
+    patterns.insert<DoubleToPtrOpLowering>(typeConverter, &getContext());
+
+
     patterns.insert<Ptr2IntOpLowering>(typeConverter, &getContext());
     // vvv yuge hack.
     patterns.insert<PtrUndefOpLowering>(typeConverter, &getContext());

### 2021-01-07 lib/Hask/HaskLowering.cpp
fix lowering of call instructions' return results

@@ -361,8 +361,11 @@ public:
   matchAndRewrite(Operation *operation, ArrayRef<Value> operands,
                   ConversionPatternRewriter &rewriter) const override {
     auto call = cast<CallOp>(operation);
+    SmallVector<Type, 4> resultTypes;
+    typeConverter->convertTypes(call.getResultTypes(), resultTypes);
+
     rewriter.replaceOpWithNewOp<CallOp>(operation, call.getCallee(),
-                                        call.getResultTypes(), operands);
+                                        resultTypes, operands);
     return success();
   }
 };
@@ -1130,6 +1133,12 @@ struct LowerHaskPass : public Pass {
         target.addDynamicallyLegalOp<CallOp>([](CallOp call) {
       for (Value arg : call.getOperands()) {
         if (!isTypeLegal(arg.getType())) {
+          return false;
+        }
+      }
+
+      for(Type t : call.getResultTypes()) {
+        if (!isTypeLegal(t)) {
           return false;
         }
       }

### 2021-01-07 lib/Hask/HaskLowering.cpp
unable to legalize affine.yield(!lz.value) -> scf.yield(!lz.value) -> scf.yield(void*)
Legalizing operation : 'affine.yield'(0x60c00001b580) {
  "affine.yield"(%6) : (!lz.value) -> ()

  * Fold {
  } -> FAILURE : unable to fold

  * Pattern : 'affine.yield -> ()' {
    ** Insert  : 'scf.yield'(0x60c00001ba00)
    ** Replace : 'affine.yield'(0x60c00001b580)

    //===-------------------------------------------===//
    Legalizing operation : 'scf.yield'(0x60c00001ba00) {
      "scf.yield"(%6) : (!lz.value) -> ()

      * Fold {
      } -> FAILURE : unable to fold

      * Pattern : 'scf.yield -

@@ -339,6 +339,22 @@ public:
   }
 };
 
+class ScfYieldOpLowering : public ConversionPattern {
+public:
+  explicit ScfYieldOpLowering(TypeConverter &tc, MLIRContext *context)
+      : ConversionPattern(scf::YieldOp::getOperationName(), 1, tc, context) {}
+
+  LogicalResult
+  matchAndRewrite(Operation *op, ArrayRef<Value> rands,
+                  ConversionPatternRewriter &rewriter) const override {
+    assert(false);
+    rewriter.replaceOpWithNewOp<scf::YieldOp>(op, rands);
+    assert(false);
+    return success();
+  }
+};
+
+
 bool isI8Ptr(Type ty) {
   auto ptr = ty.dyn_cast<LLVM::LLVMPointerType>();
   if (!ptr) {
@@ -1154,6 +1170,14 @@ struct LowerHaskPass : public Pass {
           isTypeLegal(store.getValueToStore().getType());
     });
 
+    target.addDynamicallyLegalOp<mlir::scf::YieldOp>([](scf::YieldOp yield) {
+      for(Type t : yield.getOperandTypes()) {
+        if (!isTypeLegal(t)) { return false; }
+      }
+      return true;
+    });
+
+
 
 
 
@@ -1176,6 +1200,8 @@ struct LowerHaskPass : public Pass {
     patterns.insert<FuncOpLowering>(typeConverter, &getContext());
     patterns.insert<ReturnOpLowering>(typeConverter, &getContext());
     patterns.insert<CallOpLowering>(typeConverter, &getContext());
+    patterns.insert<ScfYieldOpLowering>(typeConverter, &getContext());
+
 
     patterns.insert<ApOpConversionPattern>(typeConverter, &getContext());
     patterns.insert<ApEagerOpConversionPattern>(typeConverter, &getContext());
@@ -1185,6 +1211,7 @@ struct LowerHaskPass : public Pass {
     // Memref ops? is this really how I'm supposed to do this?
     patterns.insert<AllocOpLowering>(typeConverter, &getContext());
     patterns.insert<StoreOpLowering>(typeConverter, &getContext());
+
 
     ::llvm::DebugFlag = true;
 

### 2021-01-08 lib/Hask/HaskLowering.cpp
tried a bunch of things, nothing works. Checkpoint

@@ -1172,6 +1172,8 @@ struct LowerHaskPass : public Pass {
 
     target.addDynamicallyLegalOp<mlir::scf::YieldOp>([](scf::YieldOp yield) {
       for(Type t : yield.getOperandTypes()) {
+        llvm::errs() << "scfOperandType: |" << t << "|\n";
+        // assert(false);
         if (!isTypeLegal(t)) { return false; }
       }
       return true;

### 2021-01-19 paper/thesis.tex
add an actual thesis that is separate from the paper

--- 2021-01-05 thesis/thesis.tex
@@
 
+\section{Stuff that doesn't work}
+\begin{itemize}
+\item circular references: SSA by default cannot handle circular references.
+      We need to use the "relaxed SSA" that allows graphs. It's unclear how well
+      this adapts.
+\item A garbage collector. The usual story here, we don't really make a fair
+      comparison.
+\item Any realistic program exhibiting "difficult" laziness. I feel what would
+      be interesting would be to implement, say, GRIN, along with a sophisticated
+      pointer analysis using the MLIR analysis-as-rewrite story here.
+\item A proof of correctness of the naive demand analysis I wrote.
+\end{itemize}
+
+\section{Stuff that does work}
+\begin{itemize}
+\item An encoding of laziness that works for the simple stuff I've tried
+\item A simple version of "demand analysis" that eliminates first order laziness.
+\item Some toy examples where we generate better IR than GHC because we can
+     expose the "real work" to LLVM/MLIR in a way that they understand, by
+     eliminating the noise that is laziness.
+\end{itemize}
+
+\section{What I feel would make a good paper/thesis}
+\begin{itemize}
+\item Our story of performing worker/wrapper by outlining/inlining is
+\emph{intriguing}. I don't think it's \emph{strong}. So, to remedy that, we can:
+\item (1) Make our laziness encoding bulletproof, connect to CBPV for example,
+          handle circular references, etc.
+\item (2) Make a strong case that MLIR is an interesting target for this breed of language.
+          So, perform unification in MLIR, typeclass resolution in MLIR, perform some
+          analysis/optimisation. Ths way, we show an end-to-end prototype of something that
+          sorta works, instead of one thing that sorta works (?)
+\end{itemize}
+
 \section{Strictness Annotations: The Bad And The Ugly}
@@
 
+\subsection{Stretch: Separate dialect for pattern matching}
+\subsection{Stretch: rewrite based demand analysis}
 \subsection{Stretch: rewrite based unification}
-\subsection{Stretch: rewrite based tabled typeclass generation}
+We introduce a dialect for performing unification.
+
+\subsection{Stretch: rewrite based tabled typeclass unification}
 \subsection{Stretch: As static as possible garbage collection}

### 2021-01-19 thesis/main.tex
add an actual thesis that is separate from the paper

% - Week 1 is on the website: https://math216.wordpress.com/2020/06/27/discussion-for-this-week-starting-june-27-2020/
% - Week 2:https://math216.wordpress.com/2020/07/06/readings-and-problem-set-for-after-the-second-pseudolecture/
% - Week 3: https://math216.wordpress.com/2020/07/13/reading-and-problems-after-the-third-pseudolecture/
% Link to rising sea: https://math.stanford.edu/~vakil/216blog/FOAGnov1817public.pdf
% https://agittoc.zulipchat.com/
\documentclass{book}

%%% % https://tex.stackexchange.com/a/97128
\usepackage[sc,osf]{mathpazo}   % With old-style figures and real smallcaps.
%%% \linespread{1.025}              % Palatino leads a little more leading
%%% % Euler for math and numbers
%%% \usepackage[euler-digits,small]{eulervm}

% \usepackage{cmbright}

\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{minted}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{cancel}
\usepackage{mathtools}

% https://tex.stackexchange.com/questions/469588/strike-out-an-arrow-with-a-small-oblique-segment-like-with-nrightarrow
% \newcommand*\Neg[2][0mu]{\Neginternal{#1}{\negslash}{#2}}
% \newcommand*\sNeg[2][0mu]{\Neginternal{#1}{\snegslash}{#2}}
% \newcommand*\negslash[1]{\m@th#1\not\mathrel{\phantom{=}}}
% \newcommand*\snegslash[1]{\rotatebox[origin=c]{60}{$\m@th#1-$}}

\newcommand{\Hom}{\operatorname{Hom}}
\newcommand{\Tr}{\operatorname{Tr}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\F}{\ensuremath{\mathcal{F}}}
\newcommand{\G}{\ensuremath{\mathcal{G}}}
\newcommand{\N}{\ensuremath{\mathbb{N}}}
\newcommand{\Z}{\ensuremath{\mathbb{Z}}}
\newcommand{\T}{\ensuremath{\mathbb{T}}}
\newcommand{\Q}{\ensuremath{\mathbb{Q}}}
\newcommand{\C}{\ensuremath{\mathbb{C}}}
\newcommand{\Cstar}{\ensuremath{\mathbb{C}^*}}
\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\CP}{\ensuremath{\mathbb{CP}}}
\renewcommand{\P}{\ensuremath{\mathbb{P}}}
\newcommand{\A}{\ensuremath{\mathbb{A}}}
\renewcommand{\O}{\ensuremath{\mathcal{O}}}
\newcommand{\RP}{\ensuremath{\mathbb{RP}}}
\newcommand{\Spec}{\operatorname{Spec}}
\newcommand{\spec}{\operatorname{Spec}}
\newcommand{\m}{\mathfrak{m}}
\newcommand{\p}{\mathfrak{p}}
\newcommand{\q}{\mathfrak{q}}
\newcommand{\mSpec}{\m\operatorname{Spec}}
\newcommand{\mspec}{\m\operatorname{Spec}}
\newcommand{\frakp}{\ensuremath{\mathfrak{p}}}
\newcommand{\fraka}{\ensuremath{\mathfrak{a}}}
\newcommand{\coker}{\operatorname{coker}}
\newcommand{\colim}{\operatorname{colim}}
\newcommand{\Span}{\operatorname{span}}
\newcommand{\spn}{\Span}
\newcommand{\inv}{{\ensuremath{-1}}}
\newcommand{\im}{\operatorname{Im}}
\newcommand{\piinv}{\ensuremath{\pi^\inv}}
\renewcommand{\k}{\mathbb{k}} % field 
\newcommand{\pinv}{\piinv}
\newcommand{\piinverse}{\piinv}
\newcommand{\supp}{\operatorname{Support}}
\newcommand{\Set}{\ensuremath{\mathbf{Set}}}
\newcommand{\rad}{\sqrt} % radical
\newcommand{\ann}{\operatorname{Annhilator}} % annhilator
\newcommand{\osum}{\oplus} % I confuse it often enough
\newcommand{\tensor}{\otimes} % I confuse it often enough
\renewcommand{\c}{\complement} % got tiring
\newcommand{\geqzero}{\ensuremath{\geq 0}}
\newcommand{\gezero}{\gezero}
\newcommand{\gtzero}{\ensuremath{> 0}}
\newcommand{\sym}{\operatorname{Sym}}
\newcommand{\Sym}{\sym}
\newcommand{\nil}{\operatorname{NilRadical}}
\newcommand{\Nil}{\nil}
\newcommand{\grade}{\ensuremath{\bullet}}

\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
\DeclarePairedDelimiter\fracpart{\{}{\}}
\theoremstyle{definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{example}[theorem]{Example}
\newtheorem{nonexample}[theorem]{Non Example}
\newtheorem{aside}[theorem]{Aside}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{note}[theorem]{Note}
\newtheorem{slogan}[theorem]{Slogan}
\newtheorem{Proposition}[theorem]{Proposition}

\begin{document}
\tableofcontents

% https://qchu.wordpress.com/2009/08/30/the-orthogonality-relations-for-representations-of-finite-groups/

\chapter{Repr Theory of finite groups: Benjamin Steinberg: Ch 3}
\begin{theorem} Semisimplicity of representations of finite groups:
Every representation can be written as a direct sum of irreducible representations.
\end{theorem}
\begin{proof}
Let $\rho : G \rightarrow GL(V)$ be a representation. If $\rho$ is irreducible,
we are done. Otherwise, let $W \subsetneq V$ be a non-trivial $G$ invariant 
subspace of $V$. Now, let $\langle \cdot, \cdot \rangle$ be any
inner product (which definitely exists. Take a basis and create an inner product
from that). From this, Induce an averaged inner product:

\begin{align*}
\langle v_1, v_2 \rangle_G \equiv 
\sum_{g \in G}  \langle \rho(g) v_1, \rho(g) v_2 \rangle
\end{align*}

Under this inner product, $\rho(h)$ is unitary for all $h \in H$. This is
proved as follows:

\begin{align*}
&\langle \rho(h) v_1, \rho(h) v_2 \rangle_H = \sum_{g \in G}  \langle (\rho(g)(\rho(h) v_1)), \rho(g)(\rho(h)(v_2)) \rangle \\
&= \sum_{g \in G}  \langle \rho(gh) (v_1), \rho(gh)(v_2) \rangle \\
&= \sum_{g = g' h^{-1} \in G}  \langle \rho(gh) (v_1), \rho(gh)(v_2) \rangle \\
&= \sum_{g = g' h^{-1} \in G}  \langle \rho(g' h^{-1} h) (v_1), \rho(g' h^{-1} h)(v_2) \rangle \\
&= \sum_{g = g' h^{-1} \in G}  \langle \rho(g') (v_1), \rho(g')(v_2) \rangle \\
&= \langle v_1 , v_2 \rangle_G
\end{align*}

Thus, we have $W$ as a $G$ invariant subspace of the unitary $\rho(g)$. So we
can define the orthogonal subspace $W^\perp \equiv \{ v \in V : \langle v, W \rangle_G = 0 \}$. That is:
$W^\perp \equiv \{ v \in V : \forall w \in W, \langle v, w \rangle_G = 0\}$.
This is intuitively invariant , because if it were not, then we would have some
element $g \in G$ that would pull elements of $W^\perp$ into $W$: $W^\perp \xmapsto{g} W$.
But this would mean that $W \xmapsto{g^{-1}} W^\perp$. This is impossible because $W$
is a $G$-invariant subspace.

Formally, let $w^\perp \in W^\perp$. Now note that if $\rho(g) w^\perp \in W$,
then there exists a $w \in W$ such that $\langle \rho(g)w^{\perp}, w \rangle \neq 0$.
This gives us the computation:

\begin{align*}
&\langle \rho(g)w^{\perp}, w \rangle \neq 0 \\ 
&=\langle \rho(g)w^{\perp}, \rho(g) \rho(g^{-1}) w \rangle \neq 0 \\ 
&\text{$\rho(g)$ is unitary, so can be removed from inner product computation} \\
&= \langle w^{\perp}, \rho(g^{-1}) w \rangle \neq 0 \\ 
&\text{(Let $w' = \rho(g^{-1})w \in W$ as $W$ is $G$-invariant)} \\
&= \langle w^{\perp}, w'\rangle \neq 0 \\ 
\end{align*}

But this means that $w^{\perp}$ is not orthogonal to $W$, a contradiction.

\end{proof}

\section{3.3: Equivalent 1D representations are equal}

Let $p, q : G \rightarrow \C^*$. Let there be an isomorphism $T: \C^*  \rightarrow \C^*$
such that $p = Tq T^{-1}$.

Write with respect to a basis. $p(x) = \lambda_p x, q(x) = \lambda_q x, T(x) = \lambda_Tx $.
This gives us the equation:

\begin{align*}
&p = T q T^\inv \\
&p(x) = T(q(T^\inv(x))) \\
&\lambda_p(x) = \lambda_t(\lambda_q(1/\lambda_t))x \\
&\lambda_p(x) = \lambda_q(x) \\
&\lambda_p = \lambda_q
\end{align*}

\section{3.5: average of representations is a projection}
Let $p: G \rightarrow GL(V)$ be a representation. Define the fixed
subspace $V^G = \{ v \in V : \forall g \in G, p_g(v)  = v \}$.

First we see that this subspace is $G$ invariant. Let $v \in V^G$. $p_g(v) = v$
(by definition of $V^G$), hence $p_g(v) \in V^G$.

Define $P: V \rightarrow V; P(v) \equiv 1/|G|\sum_{g \in G}p_g(v)$.
We need to show that $P : V \rightarrow V^G$ (That is, the image lies entirely in $V^G$).

Consider $v \in V$. Now compute:

\begin{align*}
&p_h(P(v)) = p_h(1/|G|\sum_{g \in G} p_g(v)) \\
& = 1/|G|\sum_{g \in G} p_h(p_g(v)) \\
& = 1/|G|\sum_{g \in G} p_{hg}(v) \\
& \text{($g \mapsto hg$ is a bijection. $\sum_{g \in G} p_{hg}$ visits each element)} \\
& = 1/|G|\sum_{g \in G} p_{g}(v) = P(v)\\
\end{align*}

$p_h(P(v)) = P(v)$ hence $\forall v \in V, P(v) \in V^G$.

Next we are to show that $P: V \rightarrow V^G$ is a projection: $P \circ P = P$.
See that the image of $P$ is entirely in $V^G$. Now we need to show that $P(v \in V^G) = v$:
\begin{align*}
&P(v \in V^G) = 1/|G| \sum_{g \in G} p_h(v) \\
&\text{(since $v \in V^G$, $v$ is fixed by all $p_h$)} \\
&P(v \in V^G) = 1/|G| \sum_{g \in G} v = 1/|G| (|G| v) = v
\end{align*}

Hence, $P \circ P = P$.

Now, we can show that $dim(V^G)$ is the rank of $P$. In the eigenbasis of $P$,
we will have a $+1$ for each basis vector in the image $V^G$ and a zero 
everywhere else. Thus the rank of $P$ is the dimension of $V^G$.

Similarly, by the eigenbasis argument, the trace of $P$ will add up all the $1$s
in the eigenbasis, which is equal to the dimension of $V^G$.

Thus we conclude:

$$
dim(V^G) = rank(P) = Tr(P) = 1/|G| \sum_g Tr(p_g)
$$

\section{3.6: conjugates don't have to be equivalent}
Let $p: G \rightarrow GL(C)$ be a representation. Set $s(v) = \overline{p(v)}$.
Show that $p, s$ do not have to be equivalent.

Let $G \equiv \mathbb Z/3Z$. Let $p(1) \equiv e^{i\pi/3}$. We get $s(1) = \overline{p(1)} = \overline{e^{i \pi/3}} = e^{i 2 \pi/3}$.
These are not equivalent, because they are 1D representations and are not
isomorphic (recall that equivalent 1D representations are isomorphic).

\section{3.7: bijection between unitary 1D representations of $\Z$ and elements of $\T$}

Recall that $\T \equiv \{ e^{ir} : r \in \R \}$.

Consider a 1D representation of $\Z$, $p: \Z \rightarrow \Cstar$.
We must have that $p(z) p^{-1}(z) = 1$. Since $p$ is unitary, we have that $p^{-1} = p^\dagger = p^*$.
Thus we get $p(z)p^*(z) = 1$, or $|p(z)|^2 = 1$. Hence $p(z) \in \T$. Biject
to $T$ considering the image of $1 \mapsto p(1) \in T$, since the representation of $\Z$
is determined by the image of $1$ under $p$.

\section{3.8: Eigenvectors}
(TODO).

\chapter{Repr Theory of finite groups: Benjamin Steinberg: Ch 4}

Point of interest that I had never noticed: the space of representation morphisms/intertwinings
themselves form a vector space!

\begin{theorem}
Irreducible representations (irreps) of an abelian $G$ have degree 1 
\end{theorem}
\begin{proof}
Let $\phi: G \rightarrow V$ where $G$ is abelian.
For some $h \in G$. Define $T \equiv \phi_h$. Now note that:

$$ T \phi_g = \phi_h \phi_g = \phi{hg} = \phi_g \phi_h = \phi_g T $$. So we have
a $T$ that is an intertwining of $\phi$ with $\phi$: $T_h \in \Hom(\phi, \phi)$.
Thus, by schur's lemma, we must have that $T_h = \lambda_h I$. Now let $v \in V$.
Consider some element $kv \in \C v$: $T_h(kv) = \lambda_h (kv) \in \C v$.
Hence, $T_h(\C v) \in \C v$.

Now we can see that $\C v$ is a $G$ invariant subspace, since the choice of $h$
was arbitrary.

But since the representation $\phi$ is irreducible, we must have that $\C v$
equals the whole space $V$, and thus $\dim V = 1$

\end{proof}

\begin{theorem}
representations of abelian groups are simultaneously diagonalizable.
\end{theorem}
\begin{proof}
Let $\phi: G \rightarrow GL(V)$. Since all operators $\phi_g, \phi_h$ commute
pairwise, they are all simultaneously diagonalizable with each other. Hence
there exists some invertible transform $T$ such that $T \phi_g T^{-1}$ is
diagonal for all $g \in G$ ($T$ is invariant of $G$).
\end{proof}

\begin{definition}
Group algebra: The inner product space of functions $L(G) \equiv \mathbb C^G = \{ f: G \rightarrow \C \}$,
with inner product given by $\langle f | f' \rangle \equiv  1/|G|\sum_{g \in G} f(g) \overline{f'(g)}$,
and algebra given by convolution: $(f f')(g) \equiv \sum_{hh' = g}f(h_1)f'(h')$.
\end{definition}

\begin{lemma}
All linear maps can be averaged to get intertwinings. Let $p: G \rightarrow GL(V)$
and $r: G \rightarrow GL(W)$. Let $T \in \Hom(V, W)$ be a linear map. Now
define $T^\sharp: V \rightarrow W;  T^\sharp \equiv 1/|G| \sum_{g \in G} p_g; T; r_{g^-1}$.
We will show:
\begin{enumerate}
\item $T^\sharp \in \Hom(p, r)$ is an intertwining for any linear map.
\item $T \in \Hom_G(p, r)$ then $T^\sharp = T$.
\item $( \cdot )^\sharp : \Hom(V, W) \rightarrow  \Hom_G(p, r)$ is linear.
\item $( \cdot )^\sharp : \Hom(V, W) \rightarrow  \Hom_G(p, r)$ is onto.
\end{enumerate}

Thus, the map $( \cdot)^\sharp$ is a projection from the space $\Hom(V, W)$ to the
subspace $\Hom_G(p, r)$.
\end{lemma}
\begin{proof}
We verify that $T^\sharp$ is an intertwining by direct computation. We need
to check that $p_g; T^\sharp = T^\sharp; r_g$:

\begin{align*}
&x; T^\sharp; r_g =  \\
&x; 1/|G| \sum_{x \in G} (p_g; T; r^{g^{-1}}); r_g \\
&1/|G| \sum_{x \in G} (x; p_g; T; r^{e}) \\
&1/|G| \sum_{x \in G} (x; p_g; T) \\
\end{align*}

To prove (2), if $T \in \Hom(p, r)$, then consider:

\begin{align*}
&T^\sharp(x) = 1/|G|\sum_{g \in G} r_{g^{-1}} (T(p_g (x))) \\
& = 1/|G|\sum_{g \in G} r_{g^{-1}} (r_g(T (x))) \\
& = 1/|G|\sum_{g \in G} r_{e}(T (x))) \\
& = 1/|G|\sum_{g \in G} (T (x))) \\
& = T(x)
\end{align*}

(3) To establish linearity, we compute once again:

\begin{align*}
&(c_1T_1 + c_2 T_2)^\sharp(x) = 1/|G|\sum_{g \in G} r_{g^{-1}} (c_1 T_1 + c_2 T_2) p_g \\
&(c_1T_1 + c_2 T_2)^\sharp(x) = \\
& \quad c_1 \cdot 1/|G|\sum_{g \in G} r_{g^{-1}}  T_1 \rho_g + 
 c_2 \cdot 1/|G|\sum_{g \in G} r_{g^{-1}} T_2 \rho_g \\
&= c_1 T_1^\sharp + c_2 T_2^\sharp
\end{align*}

(3) immediately follows. The map is onto because $\Hom(p, r) \subseteq \Hom(V, W)$.
Hence, every element $T \in \Hom(p, r)$ is mapped onto itself by $T$ itself:
every $T \in \Hom(p, r) \in Hom(V, W)$, we have that $T^\sharp = T \in \Hom(p, r) \in \Hom(V, W)$.

Alternatively, note that $\Hom(p, r)$ is a subspace of $\Hom(V, W)$, and that 
$\sharp$ is a projection. Since projections are always onto their \emph{image}, we are done.
\end{proof}

\begin{theorem}
We know exactly what $T^\sharp$ looks like.
If $p \in GL(V)$, $r \in GL(W)$, $T \in \Hom(V, W)$:
\begin{enumerate}
\item If $p \neq r$ then $T^\sharp = 0$
\item If $p = r$ then $T^\sharp = \frac{\Tr(t)}{\deg(p)} I$
\end{enumerate}
\end{theorem}
\begin{proof}
If $p \neq r$ then $\Hom(p, r) = 0$ and hence $T^\sharp \in \Hom(p, r) = 0$.

If $p = r$ then by Schur's lemma $T^\sharp = \lambda I$ for some $\lambda in \C$. Recall
the argument: Let $T$ have full eigenspectrum. Then, note that the eigenspaces
of different eigenvalues are orthogonal. But that's impossible if the
representation is irreducible. So we have only a single eigenspace, corresponding
to the single eigenvalue $\lambda$.  Thus we must have that $T^\sharp = \lambda I$.
We want to know precisely what $\lambda$ is. Since $T \in \hom(p, r)$, we have
that $V = W$ and $T \in \Hom(V, V)$. Thus,

\begin{align*}
&\Tr(T^\sharp) = \lambda \Tr(I_{V, V}) \\
& \Tr(T^\sharp) = \lambda \dim V = \lambda \deg \phi \\
& \lambda = \Tr(T^\sharp)/deg \phi
\end{align*}

We can now relate $\Tr(T^\sharp)$  to $\Tr(T)$ using the definition of $T^\sharp$,
and remember that $p = r$:
\begin{align*}
&\Tr(T^\sharp) = 1/|G| \sum_{g \in G} Tr(r_{g^{-1}} T p_g) \\
&= 1/|G| \sum_{g \in G}\Tr(p_{g^{-1}} T p_g) \\
&= 1/|G| \sum_{g \in G} \Tr(p_g p_{g^{-1}} T) \\
& = 1/|G| \sum_{g \in G} \Tr(p_{g g^{-1}} T) = 1/|G| \sum_{g \in G}\Tr(p_e T) \\
&= 1/|G| \sum_{g \in G} \Tr(I T) \\
&= \Tr(T)
\end{align*}

Hence, $T^\sharp = \lambda I = \frac{\Tr(T^\sharp)}{\deg \phi}I = \frac{\Tr(T)}{\deg \phi}I$
\end{proof}

\begin{theorem}
Let $p: G \rightarrow U_n(\C)$, $r: G \rightarrow U_m(\C)$ be unitary representations
of $G$. Let $A^x_y \equiv \delta^x_k \delta^i_y \in M_{mn}(\C)$. Then we will
show that ${A^\sharp}^l_j = \langle p^i_j, r^k_l \rangle$.
\end{theorem}
\begin{proof}
Since $r$ is unitary, $r_g^{-1} = r_g^\dagger$. Thus, $r[l][k]^{-1} = \overline{r[k][l](g)}$.
This gives us the computation:

\end{proof}

\section{Characters and class functions}

Let $\phi: G \rightarrow GL(V)$ be a representation. Define $\chi_\phi(g) \equiv \Tr(\phi(g))$.
Now $\chi_\phi: G \rightarrow \C^*$ is a 1 dimensional representation.

To be able to compute anything with characters, we need to choose a basis to
compute a trace. So we will from now on pick matrix representations, where
we will identify $GL(V)$ with $U_n(V)$ (space of unitary matrices) with an          
appropriate choice of basis.

\chapter{Borcherds: Orthogonality relations}

\begin{itemize}
\item Link: \url{https://www.youtube.com/watch?v=BixUw1oHBec}
\end{itemize}

\begin{theorem} Schur's Lemma: If $V, W$ are irreducible then $\Hom_G(V, W)$ is 1D
if $V = W$ and $0D$ if $V \neq W$.
\end{theorem}
\begin{proof}
Let $T \in \Hom_G(V, W)$. The kernel of $T$ is a $G$ invariant subspace of $V$. since $V$
is irreducible either $\ker T = V$ (the map is zero) or $\ker T = 0$ the map is
injective. Similarly, $\Im(T)$ is $G$ invariant subspace  of $W$. Either $\Im(T) = 0$
and the map is zero, or $\Im(T) = W$ and the map is surjective. Hence the map
is either 0, or injective and surjective (thus bijective).

If the map is zero, then $\Hom_G(V, W) = 0$ since $T$ was arbitrary.
If the map is bijective, then every map in $\Hom_G(V, W)$ is invertible. This 
makes $\Hom_G(V, W)$ a division algebra over $\C$. The complex numbers are
algebraically closed, and hence it must be isomorphic to $\C$ (why? This seems
like a very interesting argument).

This tells us that $\sum_g \chi_i(g) \chi_j(g) = 0 : i \neq j$, and
$\sum_g \chi_i(g) \chi_j(g) = |G| : i = j$. This is because $\Hom_G(V_j, V_i)$ is
$1$ or $0$ depending on whether $i = j$ or not.
\end{proof}

\chapter{Clebsh gordon theory}
Given two irreducible representations $r: G \rightarrow GL(V)$, $s: G \rightarrow GL(W)$,
their tensor product $r \tensor s: G \rightarrow GL(V \tensor W)$ need not
be irreducible. Decomposing $r \tensor s$ into irreducible representations is
the Clebsh Gordon problem.

\chapter{Fully Worked out theory for $S_3$}

\chapter{Borcherds: Schur indicator}

See that the usual averaging inner product is sesquilinear: it's linear
in the left, and conjugate linear in the right.

We want to understand the space of $G$ invariant \textbf{bilinear} forms: $V \tensor V \rightarrow \C$.

\end{document}

### 2021-01-19 hask-opt/hask-opt.cpp
Updated the MLIR version. Lowering tests fail :( Failed Tests (10):   FRONTEND :: io.rs   FRONTEND :: update.rs   HASK_OPT :: Lowering/lower-ap-and-force-2.mlir   HASK_OPT :: Lowering/lower-fn-ref.mlir   HASK_OPT :: Lowering/lower-indirect-call.mlir   HASK_OPT :: Lowering/lower-lz-and-affine.mlir   HASK_OPT :: Lowering/lower-lz-and-memref.mlir   HASK_OPT :: Lowering/lower-memref-of-lz-value.mlir   HASK_OPT :: Lowering/lower-tridiag-prescanl.mlir   HASK_OPT :: Lowering/lower-worker-wrapper-maybe-non-tail-recursive.mlir
Testing Time: 9.45s
  Passed    : 50
  Unresolved: 44
  Failed    : 10

@@ -86,7 +86,7 @@ int main(int argc, char **argv) {
   registry.insert<mlir::standalone::HaskDialect>();
   registry.insert<mlir::grin::GRINDialect>();
   registry.insert<mlir::ptr::PtrDialect>();
-  registry.insert<mlir::unification::UnificationDialect>();
+  registry.insert<mlir::unif::UnificationDialect>();
 
   // registry.insert<mlir::StandardOpsDialect>();
   // registry.insert<mlir::AffineDialect>();

### 2021-01-19 include/GRIN/GRINOps.h

@@ -59,7 +59,7 @@ public:
   static StringRef getOperationName() { return "grn.box"; };
   static ParseResult parse(OpAsmParser &parser, OperationState &result);
   std::string getDataConstructorName() {
-    return this->getAttrOfType<FlatSymbolRefAttr>("value").getValue().str();
+    return this->getOperation()->getAttrOfType<FlatSymbolRefAttr>("value").getValue().str();
   }
   void print(OpAsmPrinter &p);
 };

### 2021-01-19 include/Hask/HaskOps.h

@@ -310,7 +310,7 @@ public:
   static const char *getDataConstructorAttrKey() { return "dataconstructor"; }
   static const char *getDataTypeAttrKey() { return "datatype"; }
   StringRef getDataConstructorName() {
-    return getAttrOfType<FlatSymbolRefAttr>(getDataConstructorAttrKey())
+    return getOperation()->getAttrOfType<FlatSymbolRefAttr>(getDataConstructorAttrKey())
         .getValue();
   }
 

### 2021-01-19 include/Unification/UnificationDialect.h

@@ -5,7 +5,7 @@
 #include <llvm/ADT/ArrayRef.h>
 
 namespace mlir {
-namespace unification {
+namespace unif {
 
 class UnificationDialect : public mlir::Dialect {
 public:
@@ -13,25 +13,31 @@ public:
   mlir::Type parseType(mlir::DialectAsmParser &parser) const override;
   void printType(mlir::Type type,
                  mlir::DialectAsmPrinter &printer) const override;
-  static llvm::StringRef getDialectNamespace() { return "unification"; }
+  static llvm::StringRef getDialectNamespace() { return "unif"; }
 };
 
-class UnificationType : public Type {
+class UnifType : public Type {
 public:
   /// Inherit base constructors.
   using Type::Type;
 
   /// Support for UnificationLikeTypeTraits.
   using Type::getAsOpaquePointer;
-  static UnificationType getFromOpaquePointer(const void *ptr) {
-    return UnificationType(static_cast<ImplType *>(const_cast<void *>(ptr)));
+  static UnifType getFromOpaquePointer(const void *ptr) {
+    return UnifType(static_cast<ImplType *>(const_cast<void *>(ptr)));
   }
   /// Support for isa/cast.
   static bool classof(Type type);
   UnificationDialect &getDialect();
 };
 
+class UnifNodeType : public mlir::Type::TypeBase<UnifNodeType, UnifType, TypeStorage> {
+public:
+  using Base::Base;
+  static UnifNodeType get(MLIRContext *context) { return Base::get(context); }
+};
+
 void registerLowerUnificationPass();
 
-}; // namespace unification
+}; // namespace unif
 }; // namespace mlir

### 2021-01-19 include/Unification/UnificationOps.h

@@ -26,6 +26,45 @@
 namespace mlir {
 namespace unif {
 
+class UnifRootOp
+    : public Op<UnifRootOp, OpTrait::OneResult, OpTrait::ZeroOperands> {
+public:
+  using Op::Op;
+  static StringRef getOperationName() { return "unif.root"; };
+  static ParseResult parse(OpAsmParser &parser, OperationState &result);
+  void print(OpAsmPrinter &p);
+};
 
-} // namespace grin
+class UnifConstructorOp
+    : public Op<UnifRootOp, OpTrait::OneResult, OpTrait::VariadicOperands> {
+public:
+  using Op::Op;
+  static StringRef getOperationName() { return "unif.constructor"; };
+  StringRef getConstructorName() {
+    return this->getOperation()
+        ->getAttrOfType<StringAttr>(getConstructorNameKey())
+        .getValue();
+  };
+  static StringRef getConstructorNameKey() { return "value"; }
+  static ParseResult parse(OpAsmParser &parser, OperationState &result);
+  void print(OpAsmPrinter &p);
+};
+
+class UnifVarOp
+    : public Op<UnifVarOp, OpTrait::OneResult, OpTrait::OneOperand> {
+public:
+  using Op::Op;
+  static StringRef getOperationName() { return "unif.var"; };
+  StringRef getName() {
+    return this->getOperation()
+        ->getAttrOfType<StringAttr>(getNameKey())
+        .getValue();
+  };
+  Value getParent() { return this->getOperand(); }
+  static StringRef getNameKey() { return "value"; }
+  static ParseResult parse(OpAsmParser &parser, OperationState &result);
+  void print(OpAsmPrinter &p);
+};
+
+} // namespace unif
 } // namespace mlir

### 2021-01-19 lib/Hask/HaskLowering.cpp

@@ -312,8 +312,8 @@ public:
     // The problem is that I can't copy *all* attributes, because Type
     // is also an attribute x(.
 
-    if (Attribute visibility = funcOp.getAttr("sym_visibility")) {
-      newFuncOp.setAttr("sym_visibility", visibility);
+    if (Attribute visibility = funcOp->getAttr("sym_visibility")) {
+      newFuncOp->setAttr("sym_visibility", visibility);
     }
     // newFuncOp.setAttrs(funcOp.getAttrs());
     rewriter.inlineRegionBefore(funcOp.getBody(), newFuncOp.getBody(),
@@ -379,11 +379,11 @@ bool isI8Ptr(Type ty) {
   if (!ptr) {
     return false;
   }
-  auto elem = ptr.getElementType().dyn_cast<LLVM::LLVMIntegerType>();
+  auto elem = ptr.getElementType().dyn_cast<IntegerType>();
   if (!elem) {
     return false;
   }
-  return elem.getBitWidth() == 8;
+  return elem.getWidth() == 8;
 }
 
 // I don't understand why I need this.
@@ -606,7 +606,7 @@ public:
   // fill the region `out` with the ith RHS of the caseop.
   void genCaseAltRHS(Region *out, CaseOp caseop, Value scrutinee, int i,
                      ConversionPatternRewriter &rewriter) const {
-    ModuleOp mod = caseop.getParentOfType<ModuleOp>();
+    ModuleOp mod = caseop->getParentOfType<ModuleOp>();
     assert(out->args_empty());
     assert(out->getBlocks().size() == 1);
     llvm::SmallVector<Value, 4> rhsVals;
@@ -633,7 +633,7 @@ public:
   scf::IfOp genCaseAlt(mlir::standalone::CaseOp caseop, Value scrutinee, int i,
                        const std::vector<int> &order,
                        ConversionPatternRewriter &rewriter) const {
-    ModuleOp mod = caseop.getParentOfType<ModuleOp>();
+    ModuleOp mod = caseop->getParentOfType<ModuleOp>();
 
     // check if equal
     const bool hasNext = (i + 1 < (int)order.size());
@@ -710,7 +710,7 @@ public:
     rewriter.replaceOp(caseop, caseladder.getResults());
 
     llvm::errs() << "\nvvvvvvcase op module [after inline]vvvvvv\n";
[…]

### 2021-01-19 lib/Hask/WorkerWrapperPass.cpp

@@ -87,7 +87,7 @@ struct ForceOfKnownApPattern : public mlir::OpRewritePattern<ForceOp> {
     if (!apfnname) {
       return failure();
     }
-    ModuleOp mod = force.getParentOfType<ModuleOp>();
+    ModuleOp mod = force->getParentOfType<ModuleOp>();
     FuncOp fn = mod.lookupSymbol<FuncOp>(apfnname);
     if (!fn) {
       return failure();
@@ -130,7 +130,7 @@ struct OutlineUknownForcePattern : public mlir::OpRewritePattern<ForceOp> {
       // region that is dominated by the BB that `force` lives in.
       // For now, approximate.
       if (force.getOperation()->getBlock() !=
-          &force.getParentRegion()->front()) {
+          &force->getParentRegion()->front()) {
         assert(false && "force not in entry BB");
         return failure();
       }
@@ -138,9 +138,9 @@ struct OutlineUknownForcePattern : public mlir::OpRewritePattern<ForceOp> {
       // is this going to break *completely*? or only partially?
       std::unique_ptr<Region> r = std::make_unique<Region>();
       // create a hask func op.
-      FuncOp parentfn = force.getParentOfType<FuncOp>();
+      FuncOp parentfn = force->getParentOfType<FuncOp>();
 
-      ModuleOp module = parentfn.getParentOfType<ModuleOp>();
+      ModuleOp module = parentfn->getParentOfType<ModuleOp>();
       rewriter.setInsertionPointToEnd(&module.getBodyRegion().front());
 
       FuncOp outlinedFn = rewriter.create<FuncOp>(
@@ -165,7 +165,7 @@ struct ForceOfThunkifyPattern : public mlir::OpRewritePattern<ForceOp> {
   mlir::LogicalResult
   matchAndRewrite(ForceOp force,
                   mlir::PatternRewriter &rewriter) const override {
-    // HaskFuncOp fn = force.getParentOfType<HaskFuncOp>();
+    // HaskFuncOp fn = force->getParentOfType<HaskFuncOp>();
     ThunkifyOp thunkify = force.getOperand().getDefiningOp<ThunkifyOp>();
     if (!thunkify) {
       return failure();
[…]

### 2021-01-19 lib/Pointer/PointerDialect.cpp

@@ -36,19 +36,13 @@
 using namespace mlir;
 using namespace mlir::ptr;
 
-LLVM::LLVMIntegerType getInt8Ty(MLIRContext *ctx) {
-  return LLVM::LLVMIntegerType::get(ctx, 8);
-}
+IntegerType getInt8Ty(MLIRContext *ctx) { return IntegerType::get(ctx, 8); }
 
-LLVM::LLVMIntegerType getInt64Ty(MLIRContext *ctx) {
-  return LLVM::LLVMIntegerType::get(ctx, 8);
-}
-
+IntegerType getInt64Ty(MLIRContext *ctx) { return IntegerType::get(ctx, 64); }
 
 LLVM::LLVMPointerType getInt8PtrTy(MLIRContext *ctx) {
   return LLVM::LLVMPointerType::get(getInt8Ty(ctx));
 }
-
 
 bool PtrType::classof(Type type) {
   return llvm::isa<PtrDialect>(type.getDialect());
@@ -121,7 +115,7 @@ void IntToPtrOp::print(OpAsmPrinter &p) {
 // === DOUBLE TO PTR ===
 
 void DoubleToPtrOp::build(mlir::OpBuilder &builder, mlir::OperationState &state,
-                       Value v) {
+                          Value v) {
   FloatType ft = v.getType().dyn_cast<FloatType>();
   assert(ft && "expected argument to be of float type");
   state.addOperands(v);
@@ -131,8 +125,6 @@ void DoubleToPtrOp::build(mlir::OpBuilder &builder, mlir::OperationState &state,
 void DoubleToPtrOp::print(OpAsmPrinter &p) {
   p.printGenericOp(this->getOperation());
 };
-
-
 
 // === FNPTR TO VOID PTR ===
 // === FNPTR TO VOID PTR ===
@@ -157,8 +149,8 @@ void FnToVoidPtrOp::print(OpAsmPrinter &p) {
 // === MEMREF TO VOID PTR ===
 // === MEMREF TO VOID PTR ===
 
-void MemrefToVoidPtrOp::build(mlir::OpBuilder &builder, mlir::OperationState &state,
-                          Value v) {
+void MemrefToVoidPtrOp::build(mlir::OpBuilder &builder,
+                              mlir::OperationState &state, Value v) {
   assert(v.getType().isa<MemRefType>());
   state.addOperands(v);
[…]

### 2021-01-19 lib/Unification/UnificationDialect.cpp

@@ -33,13 +33,13 @@
 #include "mlir/Transforms/DialectConversion.h"
 
 using namespace mlir;
-using namespace mlir::unification;
+using namespace mlir::unif;
 
-bool UnificationType::classof(Type type) {
+bool UnifType::classof(Type type) {
   return llvm::isa<UnificationDialect>(type.getDialect());
 }
 
-UnificationDialect &UnificationType::getDialect() {
+UnificationDialect &UnifType::getDialect() {
   return static_cast<UnificationDialect &>(Type::getDialect());
 }
 UnificationDialect::UnificationDialect(mlir::MLIRContext *context)
@@ -47,16 +47,17 @@ UnificationDialect::UnificationDialect(mlir::MLIRContext *context)
 }
 
 mlir::Type UnificationDialect::parseType(mlir::DialectAsmParser &parser) const {
-    assert(false);
+    if (parser.parseOptionalKeyword("node")) {
+        return parser.getBuilder().getType<UnifNodeType>();
+    }
+    assert(false && "unknown type for unification dialect");
 }
 
 void UnificationDialect::printType(mlir::Type type, mlir::DialectAsmPrinter &p) const {
-    assert(false);
+    if (type.isa<UnifNodeType>()) {
+        p << "node";
+        return;
+    }
+    assert(false && "unknown type for unification dialect");
 }
-
-// === INT TO PTR ===
-// === INT TO PTR ===
-// === INT TO PTR ===
-// === INT TO PTR ===
-// === INT TO PTR ===
 

### 2021-01-19 lib/Unification/UnificationOps.cpp

@@ -1,4 +1,4 @@
-#include "Pointer/PointerOps.h"
+#include "Unification/UnificationOps.h"
 #include "mlir/IR/Attributes.h"
 #include "mlir/IR/BlockAndValueMapping.h"
 #include "mlir/IR/Builders.h"
@@ -15,8 +15,93 @@
 #include "mlir/Dialect/SCF/SCF.h"
 #include "mlir/Dialect/StandardOps/IR/Ops.h"
 
-#define DEBUG_TYPE "ptr-ops"
+#define DEBUG_TYPE "unification-ops"
 #include "llvm/Support/Debug.h"
 
 using namespace mlir;
-using namespace mlir::ptr;
+using namespace mlir::unif;
+
+// ROOT OP
+// ROOT OP
+// ROOT OP
+// ROOT OP
+// ROOT OP
+// ROOT OP
+
+ParseResult UnifRootOp::parse(OpAsmParser &parser, OperationState &result) {
+  result.addTypes(UnifNodeType::get(parser.getBuilder().getContext()));
+  return success();
+}
+
+// CONSTRUCTOR OP
+// CONSTRUCTOR OP
+// CONSTRUCTOR OP
+// CONSTRUCTOR OP
+// CONSTRUCTOR OP
+// CONSTRUCTOR OP
+
+ParseResult UnifConstructorOp::parse(OpAsmParser &parser,
+                                     OperationState &result) {
+  if (failed(parser.parseLParen())) {
+    return failure();
+  }
+  StringAttr name;
+  if (failed(parser.parseAttribute<StringAttr>(name))) {
+    return failure();
+  }
+  result.addAttribute(UnifConstructorOp::getConstructorNameKey(), name);
+
+  // ")"
+  if (failed(parser.parseOptionalRParen())) {
+    while (1) {
+      // <arg>
+      OpAsmParser::OperandType arg;
+      parser.parseOperand(arg);
+      parser.resolveOperand(arg,
+                            UnifNodeType::get(parser.getBuilder().getContext()),
+                            result.operands);
+      // ")" | "," <rest>
+      if (succeeded(parser.parseOptionalRParen())) {
+        break;
+      } else if (failed(parser.parseComma())) {
+        return failure();
+      }
+    }
+  }
+  result.addTypes(unif::UnifNodeType::get(parser.getBuilder().getContext()));
+  return success();
+};
+
+void UnifConstructorOp::print(OpAsmPrinter &p) {
[…]

### 2021-01-19 test/bench-vector/Spectral/test.cpp

@@ -61,6 +61,7 @@ double *test(double *us, int len) {
     return out;
 }
 
+#define NDEBUG
 #ifdef NDEBUG
 
 void BM_test(benchmark::State& state) {
@@ -68,7 +69,7 @@ void BM_test(benchmark::State& state) {
     double *as = readArray<double>("as.bin", &len);
 
     for (auto _ : state) {
-        test(as, len);
+        benchmark::DoNotOptimize(test(as, len));
     }
 
     double *out = test(as, len);

### 2021-01-19 lib/Hask/HaskLowering.cpp
Fixed bug; seemed to be an incorrect call that Asan learnt to catch now?
Failed Tests (2):
  HASK_OPT :: Lowering/lower-memref-of-lz-value.mlir
  HASK_OPT :: Lowering/lower-tridiag-prescanl.mlir

@@ -256,16 +256,14 @@ public:
   matchAndRewrite(Operation *rator, ArrayRef<Value> rands,
                   ConversionPatternRewriter &rewriter) const override {
     auto constant = cast<ConstantOp>(rator);
-
     FunctionType fnty = constant.getResult().getType().dyn_cast<FunctionType>();
-    assert(fnty);
+    assert(fnty && "expected function");
 
-//    FunctionType outty = convertFunctionType(fnty, *typeConverter, rewriter);
-
-    FunctionType outty = typeConverter->convertType(outty).dyn_cast<FunctionType>();
-
-    assert(fnty && "was asked to lower a constant op that's not a reference to "
+    FunctionType outty = typeConverter->convertType(fnty).dyn_cast<FunctionType>();
+    assert(outty && "was asked to lower a constant op that's not a reference to "
                    "a function?!");
+
+
 
     // Anything whose legality is checked with `isDynamicallyLegal` needs to be
     // `rewrite.erase`d and then `rewriter.create`d. It seems like you can't

### 2021-01-19 include/Unification/UnificationOps.h
add unification nodes

@@ -36,7 +36,7 @@ public:
 };
 
 class UnifConstructorOp
-    : public Op<UnifRootOp, OpTrait::OneResult, OpTrait::VariadicOperands> {
+    : public Op<UnifConstructorOp, OpTrait::OneResult, OpTrait::VariadicOperands> {
 public:
   using Op::Op;
   static StringRef getOperationName() { return "unif.constructor"; };
@@ -62,6 +62,15 @@ public:
   };
   Value getParent() { return this->getOperand(); }
   static StringRef getNameKey() { return "value"; }
+  static ParseResult parse(OpAsmParser &parser, OperationState &result);
+  void print(OpAsmPrinter &p);
+};
+
+class UnifUnifyOp
+    : public Op<UnifUnifyOp, OpTrait::OneResult, OpTrait::NOperands<2>::Impl> {
+public:
+  using Op::Op;
+  static StringRef getOperationName() { return "unif.unify"; };
   static ParseResult parse(OpAsmParser &parser, OperationState &result);
   void print(OpAsmPrinter &p);
 };

### 2021-01-19 lib/Unification/UnificationDialect.cpp

@@ -19,6 +19,7 @@
 #include "mlir/Support/LogicalResult.h"
 
 // dilect lowering
+#include "Unification/UnificationOps.h"
 #include "mlir/Pass/PassRegistry.h"
 #include "mlir/Transforms/DialectConversion.h"
 // https://github.com/llvm/llvm-project/blob/80d7ac3bc7c04975fd444e9f2806e4db224f2416/mlir/examples/toy/Ch6/toyc.cpp
@@ -44,10 +45,16 @@ UnificationDialect &UnifType::getDialect() {
 }
 UnificationDialect::UnificationDialect(mlir::MLIRContext *context)
     : Dialect(getDialectNamespace(), context, TypeID::get<UnificationDialect>()) {
+
+  addTypes<UnifNodeType>();
+  addOperations<UnifRootOp>();
+  addOperations<UnifConstructorOp>();
+  addOperations<UnifVarOp>();
+  addOperations<UnifUnifyOp>();
 }
 
 mlir::Type UnificationDialect::parseType(mlir::DialectAsmParser &parser) const {
-    if (parser.parseOptionalKeyword("node")) {
+    if (succeeded(parser.parseOptionalKeyword("node"))) {
         return parser.getBuilder().getType<UnifNodeType>();
     }
     assert(false && "unknown type for unification dialect");

### 2021-01-19 lib/Unification/UnificationOps.cpp

@@ -33,6 +33,11 @@ ParseResult UnifRootOp::parse(OpAsmParser &parser, OperationState &result) {
   return success();
 }
 
+void UnifRootOp::print(OpAsmPrinter &p) {
+  p.printGenericOp(this->getOperation());
+  return;
+};
+
 // CONSTRUCTOR OP
 // CONSTRUCTOR OP
 // CONSTRUCTOR OP
@@ -54,17 +59,19 @@ ParseResult UnifConstructorOp::parse(OpAsmParser &parser,
   // ")"
   if (failed(parser.parseOptionalRParen())) {
     while (1) {
+      // ","
+      if (failed(parser.parseComma())) {
+        return failure();
+      }
       // <arg>
       OpAsmParser::OperandType arg;
       parser.parseOperand(arg);
       parser.resolveOperand(arg,
                             UnifNodeType::get(parser.getBuilder().getContext()),
                             result.operands);
-      // ")" | "," <rest>
+      // ")"?
       if (succeeded(parser.parseOptionalRParen())) {
         break;
-      } else if (failed(parser.parseComma())) {
-        return failure();
       }
     }
   }
@@ -86,13 +93,17 @@ void UnifConstructorOp::print(OpAsmPrinter &p) {
 
 ParseResult UnifVarOp::parse(OpAsmParser &parser, OperationState &result) {
   StringAttr name;
-  if (failed(parser.parseAttribute<StringAttr>(name))) {
+  if (parser.parseLParen() || parser.parseAttribute<StringAttr>(name)) {
     return failure();
   }
+
   result.addAttribute(UnifConstructorOp::getConstructorNameKey(), name);
 
   OpAsmParser::OperandType parent;
-  parser.parseOperand(parent);
+  if (parser.parseComma() || parser.parseOperand(parent) ||
+      parser.parseRParen()) {
+    return failure();
+  }
   parser.resolveOperand(parent,
                         UnifNodeType::get(parser.getBuilder().getContext()),
                         result.operands);
@@ -102,6 +113,34 @@ ParseResult UnifVarOp::parse(OpAsmParser &parser, OperationState &result) {
 };
 
 void UnifVarOp::print(OpAsmPrinter &p) {
+  p.printGenericOp(this->getOperation());
+  return;
+};
+
+// UNIFY OP
+// UNIFY OP
+// UNIFY OP
+// UNIFY OP
+// UNIFY OP
+// UNIFY OP
+
[…]

### 2021-01-21 paper/thesis.tex
nuke large sections that are outdated

@@
 code.
-% 5 
-We first consider the various deficiencies of introducing an explicit "bang pattern"
-or "strictness annotation" as is done within the Glasgow Haskell compiler today.
-Then we demonstrate how setting divergent computations to be undefined behaviour 
-cleanly enables the compiler to optimize these programs.
-% 6
-This modelling of divergence as undefined behaviour enables us to write an
-\emph{optimizing} compiler that can reason about a mix of strictness and non
-strictness. Furthermore, it enables us to modularly add non-strictness into
-an existing strict language. We demonstrate this with a prototype optimizing
-functional language called Lizzy. Lizzy uses MLIR as a backend. We introduce
-a new dialect \lz in MLIR for modularly expressing non-strictness.  As a benchmark,
-we consider a shootout between GHC's optimization of vector and matrix
-computations, and \lz's ability to optimize the same computations.
-% (http://www.easterbrook.ca/steve/2010/01/how-to-write-a-scientific-abstract-in-six-easy-steps/)
 \end{abstract}
@@
 \begin{itemize}
-\item A formal semantics for bang patterns in the Haskell language.
-\item A study of where bang patterns inhibit GHC's ability to reason about strictness.
-\item A novel yet simple extension to a lazy language semantics that remove this
-      inability of GHC (and lazy languages in general) to reason about strictness.
-\item An implementation of a prototype lazy language (Lizzy) that chooses this
-      approach pervasively.
 \item A new dialect (\lz) for MLIR to encode these semantics, and provide modular
       non-strictness.
+\item A translator from STG to \lz that exhibits the ability to 
+      eliminate laziness and leverage MLIR's loop optimisation  framework.
+\item A rewrite driven strategy for worker/wrapper that interleaves 
+     worker/wrapper analysis along with worker/wrapper rewriting, by phrasing
+     the worker/wrapper problem as an outlining/inlining problem.
 \end{itemize}
@@
 
-\section{Strictness Annotations: The Bad And The Ugly}
-
-We first motivate why strictness annotations inhibit the compiler
-from being able to optimize programs. Consider the program in
-\autoref{fig:lazy-loop}, where \texttt{out} declares
-a binding \raw{x = loop} and proceeds to not use it. This allows GHC
-to optimize out the program into \raw{out = 42}. On the
-other hand, consider  \autoref{fig:strict-loop} where \raw{out}
-declares  \emph{strict} binding \raw{!x  = loop} and proceeds to not
-use it.
-
-\begin{lstlisting}[language=haskell,caption=unoptimized \raw{out} and optimized \raw{outOptimized}]
-{-# LANGUAGE BangPatterns #-}
-loop, out, out' :: Int
-loop = loop;
-out = let x = loop in 42
-outOptimized = 42
-main :: IO ()
-main = do print out; print out'
-\end{lstlisting}
-
-\begin{lstlisting}[language=haskell, caption=unoptimized \raw{outStrict} with an incorrect optimization in \raw{outStrictOptimized}]
-{-# LANGUAGE BangPatterns #-}
-loop, out, out' :: Int
-loop = loop;
-outStrict = let !x = loop in 42
-outStrictOptimized = 42 -- INCORRECT TRANSFORM
-main :: IO ()
-main = do print out; print out'
-\end{lstlisting}
-% \caption{An example of how strictness annotations break equational reasoning}
-
-\begin{lstlisting}[language=haskell, caption=Original haskell program]
-{-# LANGUAGE BangPatterns #-}
-out b = let large = unknown 10 in 
-  if False then large else 42
-out' b = let !large = unknown  10 in 
-  if False then large else 42
-main :: IO ()
-main = do print out; print out'
-\end{lstlisting}
-
-\begin{lstlisting}[language=Haskell, caption=Correct optimization that is incorrect with a bang pattern]
-{-# LANGUAGE BangPatterns #-}
-loop, out, out' :: Int
-loop = loop;
-out b= if b then 
-  let large = unknown 10 in large else 42
--- | incorrect. If `large` had been divergent, 
--- then `out` no longer
--- exhibits divergence
-out' b = if b then 
-  let !large = unknown 10 in large else 42
-
-main :: IO ()
-main = do print out; print out'
-\end{lstlisting}
-% \caption{An example of how strictness annotations disallow floating a \texttt{let}
-%          down into the use-site, inhibiting optimization}
-
-\section{Undefined Behaviour For Divergence?}
-
-As we have shown, the existence of bang patterns obviates equational reasoning.
-We are now in a bind: we can choose to forego bang patterns, thereby
-regaining equational reasoning with a loss of performance. This is not a 
-very reasonable suggestion, as bang patterns are critical for controlling
-evaluation ordering, and ensuring that we do not build up thunks.
-
-We first propose a simple, outrageous, solution:  Ban all divergence.
-We define divergence to be \emph{undefined behaviour}. In essence, this
-allows the compiler to assume that \emph{no value} is divergent, for if a
-value were divergent, then \emph{anything} can happen. We analyze the
-merits and deficiencies of this proposal, and then provide a more
-refined version of the above solution.
-
-% \begin{figure}[ht]
-% \begin{minted}{hs}
-% {-# LANGUAGE BangPatterns #-}
-% loop, out, out' :: Int
-% outStrict = let !x = loop in 42
-% -- ==If x were not divergent==, 
-% --   it is safe to set out' to 42.
-% -- ==If x were divergent==
-% --   It is undefined behaviour. Anything can
-% --   happen, including evaluating out' to 42.
-% outStrictOptimized = 42 
-% main :: IO ()
-% main = do print out; print out'
-% \end{minted}
-% \caption{An example of how defining divergence as UB allows continued
-%          equational reasoning with bang patterns.}
-% \label{fig:strict-loop-ub}
-% \end{figure}
-
-Consider the program in \autoref{fig:strict-loop}. Here, according
-to our semantics, we define \emph{divergence as undefined behaviour}.
-Interpreted according to these semantics, here's how we legally
-allow the transformation \raw{outStrictOptimized = 42}:
-
-\begin{itemize}
-    \item Since \raw{x} is forced, we must evaluate \raw{loop} into weak head normal form.
-    \item \raw{loop} does not have a weak head normal form. This is a divergent computation.
-    \item Since we have divergence, \emph{anything} can happen [we can refine $\bot$ into a
-    value that is greater than $\bot$].
-    \item We choose to refine $\bot$ into $42$ as $\bot \sqsubset 42$.
-\end{itemize}
-
-%\begin{figure}
-%\begin{minted}{hs}
-\begin{lstlisting}[language=haskell, caption=An example of how defining divergence as UB is too agressive]
-{-# LANGUAGE BangPatterns #-}
--- | never evaluates to a value.
-main = putStrLn "42"; main
-mainOptimized = return () 
- -- | legal if divergence is UB.
-\end{lstlisting}
-%\end{minted}
-%\label{fig:io-loop}
-%\caption{An example of how defining divergence as UB is too agressive}
-%\end{figure}
-
-On the other hand, this definition is \emph{far too agressive}. Consider
-the program in \autoref{fig:io-loop}. Here, main prints $42$ forever,
-never returning the final \raw{IO ()} that it is expected to. Thus,
-we can \emph{optimize this program} into \raw{mainOptimized = return ()}
-if divergence is undefined behaviour. This is not what we want! We wish
-to allow programs which \emph{make progess without halting}, while
-still being able to exploit undefined behaviour. Thus, we must change our
-semantics to adapt to this need.
-
-\section{Progressive operational semantics}
-
-Naively speaking, one maybe tempted to summarize our approach as
-"make divergence undefined". While this is appealing, this is both too
-strong, and too naive. For example, such a definition would allow us to
-optimize away programs that \emph{progress} without \emph{terminating}.
-A tiny example of such a program, which prints ``42'' forever is
-given in \autoref{fig:print-forever}
-
-% \begin{figure}[ht]
-% \begin{minted}{hs}
-\begin{lstlisting}[language=haskell, caption=A program that makes progress without terminating.  We wish to preserve such program semantics]
-action :: IO (); action = putStrLn "42";
-main :: IO (); main = action >> main
-\end{lstlisting}
-% \end{minted}
-% \label{fig:print-forever}
-% \caption{A program that makes progress without terminating.  We wish to preserve such program semantics}
-% \end{figure}
-
-% \begin{figure}[ht]
-
-%\begin{subfigure}[b]{0.48\textwidth}
-%\begin{minted}{hs}
-\begin{lstlisting}[language=haskell]
-{-# LANGUAGE BangPatterns #-}
-loop :: Int; loop = loop
-ex1 = let x = loop in 42
-ex2 = let !x = loop in 42
-ex3 = let !x = loop in x
--- | New program
-{-# LANGUAGE BangPatterns #-}
-loop :: Int; loop = loop
-ex1 = 42
-ex2 = 42
-ex3 = let !x = loop in x
-ex3' = loop -- equivalent semantics
-\end{lstlisting}
-% \end{minted}
-% \subcaption{Maxmially transformed programs with our proposed semantics}
-% \end{subfigure}
-% \caption{Programs that exhibit a range of behviours, clarifying what values we do and do not eliminate}
-% \end{figure}
-
-We call our semantics \emph{progressive}, which tries to make as much progress
-as possible before getting stuck at a stuck term.
-\section{Formal Semantics Of Bang Patterns in Haskell}
-
-There is currently no formal definition of bang patterns in the Haskell language.
-The semantics are defined based on how the Glasgow Haskell Compiler chooses
-to manipulate these strictness annotations. So wherever we have previously argued
-that it would be ``incorrect'' for GHC to perform these transforms, we have been
-arguing from the position of:
-
-\begin{itemize}
-\item This is what GHC does today, and that is the current state of definition for bang patterns.
-\item Were GHC to make the above transformation, we believe it would violate the
-      intuitions that many people have about how GHC \emph{ought} to transform code.
-      \sid{I asked bgamari who works on GHC full-time, he said that GHC does not perform the above transform most likely.
-           Similarly, other friends (carter, davean) who commit into GHC also said likely not.}
-\end{itemize}
-
-To begin our discussion, we first formalize the semantics of bang patterns.
 \textbf{TODO: fill in semantics here.}
@@
 % =========================
-\vspace{20em}
+\vspace{15em}
 
@@
 
-\vspace{10em}
-
+\vspace{15em}
+
+\begin{tabular}{ll}
 \begin{minipage}[t][1cm][b]{0.5\textwidth}
@@
 \end{minipage}
-
-\vspace{10em}
+&
 
@@
 \inference{
-FOO
+\texttt{data C = MkC(V) }\\
+\texttt{@f(\%inc: C)} \\
+\texttt{ case inc} \\
+\texttt{   [C inv -> } \\
+\texttt{     \%inv = extract(@MKC, \%inc) : V}\\
+\texttt{     \%w = ... : V; \%wc = construct(@MKC, w) : C} \\
+\texttt{     \%rec = apEager(@f, wc)]}\\
 }{
-BAR
+\texttt{@frec(\%inv: V)}\\
+\texttt{ \%inv = extract(@MKC, \%inc) : V}\\
+\texttt{ \%w = ... : V; \%wc = construct(@MKC, w) : C} \\
+\texttt{ \%rec = apEager(@f, wc)}\\
+\texttt{@f(\%inc: C)} \\
+\texttt{ \%inv = extract(\%inc) : V}\\
+\texttt{ apEager(@frec, inv)}\\
 }
 } % end footnotesize
-\subcaption{Inline a function call}
-\end{minipage}
-\caption{Local rewrites performed to eliminate laziness (1)}
-\end{figure*}
-
-\begin{figure*}
-
+\subcaption{Outline recursive call of constructor that is immediately unwrapped}
+\end{minipage}
+% OutlineRecursiveApEagerOfConstructorPattern
+\end{tabular}
+
+\vspace{15em}
+\begin{tabular}{ll}
 \begin{minipage}[t][1cm][b]{0.5\textwidth}
@@
 \inference{
-FOO
-}{
-BAR
-}
-} % end footnotesize
-\subcaption{Outline recursive call of constructor that is immediately unwrapped}
-\end{minipage}
-% OutlineRecursiveApEagerOfConstructorPattern
-
-\begin{minipage}[t][1cm][b]{0.5\textwidth}
-{\footnotesize
-\inference{
-FOO
 }{
@@
 \end{minipage}
+&
 
@@
 \inference{
-FOO
+\texttt{data C = MKC(V) }\\
+\texttt{@f(...)}\\
+\texttt{  ...}\\
+\texttt{  \%out = constructor(C, \%w)}\\
+\texttt{  lz.return (\%out)}\\
 }{
-BAR
+\texttt{data C = MKC(V) }\\
+\texttt{@finner(...)} \\
+\texttt{ ...} \\
+\texttt{ lz.return \%w} \\
+\texttt{ } \\
+\texttt{@f(...)}\\
+\texttt{ \%w = call \%f(...)}\\
+\texttt{ \%out = constructor(C, \%w)}\\
+\texttt{ lz.return (\%out)}
 }
@@
 \end{minipage}
-
-\begin{minipage}[t][1cm][b]{0.5\textwidth}
-{\footnotesize
-\inference{
-FOO
-}{
-BAR
-}
-} % end footnotesize
-\subcaption{Peel constructor from case}
-% peel constructor from case
-\end{minipage}
-
-%% % \begin{minted}{text}
-%% % Inlining
-%% % \end{minted}
+\end{tabular}
+%% 
+%% 
+%% \begin{minipage}[t][1cm][b]{0.5\textwidth}
+%% {\footnotesize
+%% \inference{
+%% FOO
+%% }{
+%% BAR
+%% }
+%% } % end footnotesize
+%% \subcaption{Peel constructor from case}
+%% % peel constructor from case
+%% \end{minipage}
 \caption{Local rewrites performed to eliminate laziness (2)}

### 2021-01-21 paper/paper.tex
remove more content

--- 2021-01-21 paper/thesis.tex
@@
 
-\section{Aiding the performance programmer with explicit forcing}
-
-\begin{lstlisting}[language=haskell, caption=Haskell program with two datatypes  \texttt{Lazy} and \texttt{Strict}
-with the same shape of data but different strictness annotations.]
-{-# LANGUAGE BangPatterns #-}
-{-# LANGUAGE UnboxedTuples #-}
-module Foo(boxLazyAndStrict) where
--- | A large datatype
-data Box = Box Int Int Int Int Int
-data Lazy = MkLazy Box
-data Strict = MkStrict !Box
-
-data BoxPair = MkBoxPair !Box !Box
-
-boxLazyAndStrict :: Lazy 
-  -> Strict 
-  -> BoxPair
-boxLazyAndStrict lz str =
-  case lz of
-   MkLazy lbox  ->
-       case str of
-         MkStrict strbox ->
-           MkBoxPair lbox strbox
-\end{lstlisting}
-
-\begin{lstlisting}[language=haskell,caption=\ghc \core corresponding to the  above Haskell program  where there is no indication
-         that \texttt{strbox} is a value while \texttt{lbox} is a thunk]
-boxLazyAndStrict :: Lazy -> Strict -> BoxPair
- ...
- Tmpl= \ (lz_aub :: Lazy)
-         (str_auc :: Strict) ->
-         case lz_aub of { MkLazy lbox_aud ->
-         case str_auc of { MkStrict strbox_aue ->
-         case lbox_aud of dt_Xve [Occ=Once]
-         { Box _ _ _ _ _  ->
-         Foo.MkBoxPair dt_Xve strbox_aue
-         }}}}]
-\end{lstlisting}
-
-In Haskell, Core and STG, all of GHC's intermediate representations, there is
-no explicit description of the act of forcing a thunk. The \texttt{case} construct
-is responsible for \emph{both} forcing and case analysis. This is a fine state
-of affairs when we are in a language where values are forced right before they
-are used --- that is, the language is \emph{truly lazy}. However, Haskell currently
-supports bang patterns in defining data constructors, which violates this
-assumption. We begin with an example of a simple program with two datatypes,
-called \texttt{Lazy} and \texttt{Strict}, where the both contain the lazy data
-of type \texttt{Box}. \texttt{Strict}, to its name, is actually strict
-in its \texttt{Box} field. However, this is completely opaque on reading the
-Core: We need to infer from the facts:
-\begin{itemize}
-\item because \raw{lbox_aud} is \texttt{case}'d on
-\item The result of the aforementioned case is called \raw{dt_Xve}.
-\item The case is followed by the pointless match \raw{(Box _ _ _ _ _)}.
-\item Finally, the \raw{(Foo.MkBoxPair dt_Xve ...)} is built.
-\item Thus, the only reason \raw{dt_Xve} exists is to represent the forced \raw{lbox_aud}.
-\item So, the value \raw{lbox_aud} is forced into the value \raw{dt_Xve}
-\end{itemize}
-
-These infelicities, while slight, add up rapidly when attempting to optimize
-more complicated Haskell programs. On the other hand, consider our encoding
-in \lz. Since we separate the concept of \emph{forcing} a value from \emph{pattern matching},
-we can represent the force necessary as a \texttt{lz.force}, without needlessly
-introducing a pattern match. This improves the readability of the source, and
-more directly expresses the semantics we are interested in.  While GHC Core
-is an amazing intermediate representation for analysis and transformation, it
-is sub-par for \emph{consumption by the end-user}. Though the source language has
-gotten more complex, \core has changed very little; This is a source of
-strength (as the optimizer needs to only reasons about a very minimal language).
-It is also a source of obfuscation, as many overlapping ideas are represented
-by the same construct: in this case, that of forcing, and that of pattern
-matching.
-
-\begin{lstlisting}[language=llvm, caption=\lz dialect encoding of the above program where there is a clear indication of the thunk being forced]
-lz.func(%lz: !lz.adt<@Lazy>, 
-        %str: !lz.adt<Strict>) {
-  %lboxt = lz.extractvalue(%lz, 0)
-     : !lz.thunk<!lz.adt<@Box>>
-  %strboxv = lz.extractvalue(%str, 1)
-     : !lz.adt<@Box>
-  %lboxv = lz.force(%lboxt) # FORCE
-    : !lz.adt<@Box>
-  %retval = lz.construct(@MkBoxPair, 
-                         %lboxv, %strboxv)
-    : !lz.adt<@BoxPair>
-  lz.return(%retval)
-}
-\end{lstlisting}
-
 \section{Worker wrapper by local rewrites: A first stab}

### 2021-01-21 paper/paper.tex
add explanation of the simple example

@@
 \newcommand{\controlflowarrow}{\xrightarrow{~\texttt{ctrl}~}}
+
+\lstset{
+language=C++,
+basicstyle=\small\ttfamily,
+escapeinside={<@}{@>},
+columns=fullflexible,
+keywords=[2]{old},
+keywords=[3]{new},
+keywordstyle={\color{blue!80!black}},
+keywordstyle=[2]{\color{red!80!black}},
+keywordstyle=[3]{\color{green!50!black}},
+}
+
+\colorlet{diffold}{red!80!black}
+\colorlet{diffnew}{green!50!black}
 
@@
         Thunk<SimpleInt> siprev_t = thunkify(siprev);
-        SimpleInt f_prev_v = apStrict(f, siprev_t);
+        SimpleInt f_prev_v = f(siprev_t);
         return f_prev_v;
@@
 }
-
 SimpleInt f(Thunk<SimpleInt> i) {
     SimpleInt icons = force(i);
-    SimpleInt ret = apStrict(f2, icons);
+    SimpleInt ret = f2(icons);
     return ret;
@@
 %\begin{minted}{cpp}
-\begin{lstlisting}[language=c++, caption=Step 2: Removing the strict application]
+\begin{lstlisting}[language=c++, caption=Step 2: Removing the un-necessary lazy recursive call]
 SimpleInt f2(SimpleInt icons) {
@@
         SimpleInt siprev = SimpleInt(prev);
-        // Thunk<SimpleInt> siprev_t = thunkify(siprev);
-        // SimpleInt f_prev_v = apStrict(f, siprev_t);
-        Thunk<SimpleInt> f_prev_v = apStrict(f2, siprev);
+        <@\textcolor{diffold}{Thunk<SimpleInt> siprev\_t = thunkify(siprev);}@>
+        <@\textcolor{diffold}{SimpleInt f\_prev\_v = f(siprev\_t);} @>
+       <@ \textcolor{diffnew}{Thunk<SimpleInt> f\_prev\_v = f2(siprev);} @>
         return f_prev_v;
@@
     SimpleInt icons = force(i);
-    SimpleInt ret = apStrict(f2, icons);
+    SimpleInt ret = f2(icons);
     return ret;
@@
         SimpleInt siprev = SimpleInt(prev);
-        Thunk<SimpleInt> f_prev_v = apStrict(f2, siprev);
+        Thunk<SimpleInt> f_prev_v = f2(siprev);
         return f_prev_v;
@@
     int ihash = casedefault(icons);
-    SimpleInt ret = apStrict(f3, ihash);
+    <@ \textcolor{diffnew}{SimpleInt ret = f3(ihash);} @>
+
     return ret;
 }
-
-SimpleInt f(Thunk<SimpleInt> i) {
-    SimpleInt icons = force(i);
-    SimpleInt ret = apStrict(f2, icons);
-    return ret;
-}
+...
 \end{lstlisting}
@@
         int prev = ihash - 1;
-        Thunk<SimpleInt> f_prev_v = apStrict(f3, prev);
+        <@ \textcolor{diffold}{SimpleInt siprev = SimpleInt(prev);} @>
+        <@ \textcolor{diffold}{Thunk<SimpleInt> f\_prev\_v = f2(siprev);} @>
+        <@ \textcolor{diffnew}{Thunk<SimpleInt> f\_prev\_v = f3(prev);} @>
         return f_prev_v;
@@
 }
+...
+\end{lstlisting}
+}
+
+%\caption{Step 4: replace call f2(SimpleInt(prev)) to f3(prev)}
+%\end{figure}
+
+{\footnotesize
+\begin{lstlisting}[language=c++, caption=taking stock]
+SimpleInt f3(int icons) {
+    if (ihash <= 0) {
+        return SimpleInt(42);
+    } else {
+        int prev = ihash - 1;
+        <@ \textcolor{diffnew}{Thunk<SimpleInt> f\_prev\_v = f3(prev);} @>
+        return f_prev_v;
+    }
+}
+
 SimpleInt f2(SimpleInt icons) {
     int ihash = casedefault(icons);
-    SimpleInt ret = apStrict(f3, ihash);
+    <@ \textcolor{diffnew}{SimpleInt ret = f3(ihash);} @>
     return ret;
@@
 }
-%\caption{Step 4: replace call f2(SimpleInt(prev)) to f3(prev)}
-%\end{figure}
+
+At this point, we have a clean program that has been worker/wrappered. We can
+see that the call chain looks as follows:
+
+$$
+f \xrightarrow{\texttt{force}} f_2 \xrightarrow{\texttt{unwrap}} f_3 \to f_3 \dots \xrightarrow{\texttt{wrap}} f_2
+\xrightarrow{\texttt{thunk}} f
+$$
 

### 2021-01-21 paper/paper.tex
blurb more
add fleshed out example 2

@@
 
+Of course, much remains to be discussed: what about sum types? what about
+non-tail recursion? Extensions to the same idea will handle the above problems,
+while retaining the pleasing simplicity of this outlining/inlining paradigm.
+
 %\begin{figure}[htb!]
@@
 
-Currently implementing the constructor simplification
+{\footnotesize
+\begin{lstlisting}[language=c++, caption=Step 0: the initial program]
+SimpleInt g(Thunk<SimpleInt> i) {
+  SimpleInt icons = force(i);
+  int ihash = casedefault(icons);
+  if (ihash <= 0) {
+    return SimpleInt(42);
+  } else {
+    int prev = ihash - 1;
+    SimpleInt siprev = SimpleInt(prev);
+    Thunk<SimpleInt> siprev_t = thunkify(siprev);
+    SimpleInt g_prev_v = g(siprev_t);
+    int g_prev_v_hash = casedefault(g_prev_v);
+    int rethash = g_prev_v_hash + 2;
+    SimpleInt ret = SimpleInt(rethash);
+    return ret;
+  }
+}
+\end{lstlisting}
+
+\begin{lstlisting}[language=c++, caption=Step 1: outline everything after \texttt{force}]
+SimpleInt g2(SimpleInt i) {
+  int ihash = casedefault(icons);
+  if (ihash <= 0) {
+    return SimpleInt(42);
+  } else {
+    int prev = ihash - 1;
+    SimpleInt siprev = SimpleInt(prev);
+    Thunk<SimpleInt> siprev_t = thunkify(siprev);
+    SimpleInt g_prev_v = g(siprev_t);
+    int g_prev_v_hash = casedefault(g_prev_v);
+    int rethash = g_prev_v_hash + 2;
+    SimpleInt ret = SimpleInt(rethash);
+    return ret;
+  }
+}
+SimpleInt g(Thunk<SimpleInt> i) {
+  SimpleInt icons = force(i);
+  g2(icons);
+}
+\end{lstlisting}
+
+\begin{lstlisting}[language=c++, caption=Step 2: outline everything after \texttt{force}]
+SimpleInt g2(SimpleInt i) {
+  int ihash = casedefault(icons);
+  if (ihash <= 0) {
+    return SimpleInt(42);
+  } else {
+    int prev = ihash - 1;
+    SimpleInt siprev = SimpleInt(prev);
+    Thunk<SimpleInt> siprev_t = thunkify(siprev);
+    SimpleInt g_prev_v = g(siprev_t);
+    int g_prev_v_hash = casedefault(g_prev_v);
+    int rethash = g_prev_v_hash + 2;
+    SimpleInt ret = SimpleInt(rethash);
+    return ret;
+  }
+}
+SimpleInt g(Thunk<SimpleInt> i) {
+  SimpleInt icons = force(i);
+  g2(icons);
+}
+\end{lstlisting}
+
+\begin{lstlisting}[language=c++, caption=Step 2: replace recursive call to \texttt{g}]
+SimpleInt g2(SimpleInt i) {
+  int ihash = casedefault(icons);
+  if (ihash <= 0) {
+    return SimpleInt(42);
+  } else {
+    int prev = ihash - 1;
+    SimpleInt siprev = SimpleInt(prev);
+    <@\textcolor{diffold}{ Thunk<SimpleInt> siprev\_t = thunkify(siprev)} @>;
+    <@\textcolor{diffold}{ SimpleInt g\_prev\_v = g(siprev\_t);} @>
+    <@\textcolor{diffnew}{ SimpleInt g\_prev\_v = g2(siprev);} @>
+    int g_prev_v_hash = casedefault(g_prev_v);
+    int rethash = g_prev_v_hash + 2;
+    SimpleInt ret = SimpleInt(rethash);
+    return ret;
+  }
+}
+SimpleInt g(Thunk<SimpleInt> i) {
+  SimpleInt icons = force(i);
+  g2(icons);
+}
+\end{lstlisting}
+
+\begin{lstlisting}[language=c++, caption=Step 3: outline block after \texttt{casedefault}]
+SimpleInt g3(int i) {
+  if (ihash <= 0) {
+    return SimpleInt(42);
+  } else {
+    int prev = ihash - 1;
+    SimpleInt siprev = SimpleInt(prev);
+    g_prev_v = g2(siprev);
+    int g_prev_v_hash = casedefault(g_prev_v);
+    int rethash = g_prev_v_hash + 2;
+    SimpleInt ret = SimpleInt(rethash);
+    return ret;
+  }
+}
+
+SimpleInt g2(SimpleInt i) {
+  int ihash = casedefault(icons);
+  return g3(ihash);
+}
+\end{lstlisting}
+
+\begin{lstlisting}[language=c++, caption=Step 4: replace \texttt{g2} with \texttt{g3}]
+SimpleInt g3(int i) {
+  if (ihash <= 0) {
+    return SimpleInt(42);
+  } else {
+    int prev = ihash - 1;
+    <@\textcolor{diffold}{SimpleInt siprev = SimpleInt(prev);} @>
+    <@\textcolor{diffold}{g\_prev\_v = g2(siprev);}@>
+    <@\textcolor{diffnew}{g\_prev\_v = g3(prev);}@>
+    int g_prev_v_hash = casedefault(g_prev_v);
+    int rethash = g_prev_v_hash + 2;
+    SimpleInt ret = SimpleInt(rethash);
+    return ret;
+  }
+}
+\end{lstlisting}
+
+\begin{lstlisting}[language=c++, caption=Step 5: float out \texttt{SimpleInt}]
+SimpleInt g3(int ihash) {
+  int out;
+  if (ihash <= 0) {
+    <@ \textcolor{diffold}{return SimpleInt(42)}; @>
+    <@ \textcolor{diffnew}{out = 42;} @>
+  } else {
+    int prev = ihash - 1;
+    SimpleInt g_prev_v = g3(prev);
+    int g_prev_v_hash = casedefault(g_prev_v);
+    int rethash = g_prev_v_hash + 2;
+    <@\textcolor{diffold}{SimpleInt ret = SimpleInt(rethash);} @>
+    <@\textcolor{diffold}{return ret;}@>
+    <@\textcolor{diffnew}{out = rethash;} @>
+  }
+  <@ \textcolor{diffnew}{return SimpleInt(out);} @>
+}
+\end{lstlisting}
+
+\begin{lstlisting}[language=c++, caption=Step 5: taking stock]
+SimpleInt g3(int ihash) {
+  int out;
+  if (ihash <= 0) {
+    out = 42;
+  } else {
+    int prev = ihash - 1;
+    SimpleInt g_prev_v = g3(siprev);
+    int g_prev_v_hash = casedefault(g_prev_v);
+    int rethash = g_prev_v_hash + 2;
+    out = rethash;
+  }
+  return SimpleInt(out);
+}
+SimpleInt g2(SimpleInt i) {
+  int ihash = casedefault(icons);
+  return g3(ihash);
+}
+
+SimpleInt g(Thunk<SimpleInt> i) {
+  SimpleInt icons = force(i);
+  g2(icons);
+}
+\end{lstlisting}
 
@@
 % =========================
-\vspace{15em}
+\vspace{18em}
 
@@
 {\footnotesize
-
 \inference{
@@
 
-\vspace{15em}
+\vspace{20em}
 
@@
 
-\vspace{15em}
+\vspace{18em}
 \begin{tabular}{ll}

### 2021-01-21 paper/paper.tex
add PDF
add factorial worker/wrapper example that becomes a constant value

@@
 \section{Worker wrapper by local rewrites: Sum types}
+
+The transformation of floating out common patterns works equally well
+for branches as it does for sum types, since a case analysis on a sum
+type is the same as a branch on the tag.
 
@@
 
+\subsection{A toy example: eliminating wrapper overhead}
+
+\begin{lstlisting}[language=llvm,caption="Original program with wrapper and laziness overhead"]
+module {
+// fact 0 = 1
+// fact n = n*fact(n-1)
+func @f (%i : !lz.thunk<i64>) -> i64 {
+  %ival = lz.force(%i):i64
+  %out = lz.caseint %ival 
+   [0 ->   { 
+    ^entry:
+    %c1 = constant 1 : i64
+    lz.return %c1 : i64
+   }]
+   [@default -> {
+    ^entry:
+    %c1 = constant 1 : i64
+    %idec = subi %ival, %c1 : i64
+    %idecthnk = lz.thunkify(%idec : i64) 
+    %f = constant @f : (!lz.thunk<i64>) -> i64
+    %f_idec_thnk = lz.ap(%f: (!lz.thunk<i64>) -> i64, 
+        %idecthnk)
+    %f_idec_val = lz.force(%f_idec_thnk) : i64
+    %prod = muli %ival, %f_idec_val : i64
+    lz.return %prod : i64
+   }]
+  return %out : i64
+}
+
+func @c8 () -> i64 {
+  %v = std.constant 8 : i64
+  return %v: i64
+}
+
+func @main() -> i64 {
+  %f = constant @f : (!lz.thunk<i64>) -> i64
+  %c8f = constant @c8 : () -> i64
+  %c8t = lz.ap(%c8f : () -> i64)
+  %outt = lz.ap(%f: (!lz.thunk<i64>) -> i64, %c8t)
+  %out = lz.force(%outt) : i64
+  return %out : i64
+}
+}
+
+\end{lstlisting}
+
+After worker/wrapper, we get the optimized program:
+
+\begin{lstlisting}[language=llvm,caption="factorial that has been worker/wrapper'd"]
+module  {
+  func @c8() -> i64 {
+    %c8_i64 = constant 8 : i64
+    return %c8_i64 : i64
+  }
+  func @main() -> i64 {
+    %c7_i64 = constant 7 : i64
+    %f = constant @frec_force_outline : (i64) -> i64
+    %c8_i64 = constant 8 : i64
+    %0 = "lz.apEager"(%f, %c7_i64) : ((i64) -> i64, i64) -> i64
+    %1 = muli %0, %c8_i64 : i64
+    return %1 : i64
+  }
+  func @frec_force_outline(%arg0: i64) -> i64 {
+    %c1_i64 = constant 1 : i64
+    %f = constant @frec_force_outline : (i64) -> i64
+    %0 = "lz.caseint"(%arg0) ( {
+      lz.return %c1_i64 : i64
+    },  {
+      %1 = subi %arg0, %c1_i64 : i64
+      %2 = "lz.apEager"(%f, %1) : ((i64) -> i64, i64) -> i64
+      %3 = muli %arg0, %2 : i64
+      lz.return %3 : i64
+    }) {alt0 = 0 : i64, alt1 = @default} : (i64) -> i64
+    return %0 : i64
+  }
+\end{lstlisting}
+
+\sid{Explain how this happens}
+We will explain this transformation step by step. Finally, when lowered to
+LLVM, LLVM's strong loop analyses is able to compute a closed form for 
+\texttt{frec\_force\_outline}, ending in a constant time computation.
+
+\subsection{A toy example: eliminating laziness and exposing loop optimisation}
+
 \section{Related Work}

### 2021-01-21 paper/paper.tex
add example of computing sum of first n numbers

@@
 
+\begin{lstlisting}[language=llvm, caption="example of laziness with tensor computations"]
+module {
+// sum up all values in the buffer
+func @sum(%buffert: !lz.thunk<memref<?xi64>>) -> i64 {
+  %buffer = lz.force(%buffert) :  memref<?xi64>
+  %c0 = constant 0 : index
+  %N = dim %buffer, %c0 : memref<?xi64>
+  %sum_0 = constant 0 : i64
+  %sum = affine.for %i = 0 to %N step 1
+  iter_args(%sum_iter = %sum_0) -> (i64) {
+    %t = affine.load %buffer[%i] : memref<?xi64>
+    %sum_next = std.addi %sum_iter, %t : i64
+    affine.yield %sum_next : i64
+  }
+  return %sum : i64
+}
+
+// create a sequence [0..upper_bound)
+func @seq(%upper_bound: i64) ->  memref<?xi64> {
+  %upper_bound_ix = std.index_cast %upper_bound : i64 to index
+  %buf = alloc(%upper_bound_ix) : memref<?xi64>
+  affine.for %i = 0 to %upper_bound_ix step 1 {
+    %ival = std.index_cast %i : index to i64
+    affine.store %ival, %buf[%i] : memref<?xi64>
+  }
+  return %buf : memref<?xi64>
+}
+
+func private @printInt(%i: i64)
+
+// computes sum of numbers upto 1023?
+func @main ()  -> i64 {
+  %seqf = constant @seq : (i64) ->  memref<?xi64>
+  %size = std.constant 1024 : i64
+  %seqt = lz.ap(%seqf: (i64) -> memref<?xi64>, %size)
+
+  %sumf = constant @sum : (!lz.thunk<memref<?xi64>>) -> i64
+  %outt = lz.ap(%sumf : (!lz.thunk<memref<?xi64>>) -> i64, %seqt)
+  %outv = lz.force(%outt): i64
+  call @printInt(%outv) : (i64) -> ()
+  %zero = constant 0 : i64
+  return %zero : i64
+}
+}
+\end{lstlisting}
+
+\sid{MLIR can work harder, but it does not. By the time we get to LLVM, we have a closed form}
+\begin{lstlisting}[language=llvm, caption="final program after eliminating laziness and loop fusion"]
+func @main() -> i64 {
+  %c0_i64 = constant 0 : i64
+  %0 = alloc() : memref<1xi64>
+  %1 = affine.for %arg0 = 0 to 1024 iter_args(%arg1 = %c0_i64) -> (i64) {
+    %2 = index_cast %arg0 : index to i64
+    affine.store %2, %0[0] : memref<1xi64>
+    %3 = affine.load %0[0] : memref<1xi64>
+    %4 = addi %arg1, %3 : i64
+    affine.yield %4 : i64
+  }
+  call @printInt(%1) : (i64) -> ()
+  return %c0_i64 : i64
+}
+\end{lstlisting}
+
+\begin{lstlisting}[language=llvm, caption="final LLVM program after lowering"]
+define i64 @main() local_unnamed_addr !dbg !35 {
+  tail call void @printInt(i64 523776), !dbg !36
+  ret i64 0, !dbg !38
+}
+\end{lstlisting}
+
 \section{Related Work}

### 2021-09-01 hask-opt/RgnDialect.cpp
add lowering for rgn

@@ -52,8 +52,7 @@ RgnDialect::RgnDialect(mlir::MLIRContext *context)
   // clang-format off
     // addOperations<RgnReturnOp, RgnSymOp, RgnValOp, RgnCallSymOp, RgnCallValOp, RgnJumpSymOp, RgnJumpValOp
     // >();
-    addOperations<RgnReturnOp, RgnSymOp, RgnValOp, RgnCallSymOp
-    , RgnCallValOp, RgnJumpSymOp, RgnJumpValOp, RgnEndOp, RgnSelectOp>();
+    addOperations<RgnReturnOp, RgnValOp, RgnJumpValOp, RgnEndOp, RgnSelectOp>();
 
     // addAttributes<DataConstructorAttr>();
     // addInterfaces<HaskInlinerInterface>();
@@ -146,58 +145,6 @@ void RgnValOp::getSuccessorRegions(
   return;
 }
 
-// RGN OPS::RgnSymOp
-// RGN OPS::RgnSymOp
-// RGN OPS::RgnSymOp
-// RGN OPS::RgnSymOp
-// RGN OPS::RgnSymOp
-
-mlir::ParseResult RgnSymOp::parse(mlir::OpAsmParser &parser,
-                                  mlir::OperationState &result) {
-  assert(false && "unimplemented");
-}
-
-void RgnSymOp::print(mlir::OpAsmPrinter &p) {
-  p.printGenericOp(this->getOperation());
-}
-
-// RGN OPS::RgnCallValOp
-// RGN OPS::RgnCallValOp
-// RGN OPS::RgnCallValOp
-// RGN OPS::RgnCallValOp
-// RGN OPS::RgnCallValOp
-// template <typename ValsT, typename TypesT>
-// void RgnCallValOp::build(mlir::OpBuilder &builder, mlir::OperationState &state,
-//                          mlir::Value rgn, ValsT args, TypesT resultTypes) {
-//   state.addOperands(rgn);
-//   state.addOperands(args);
-//   state.addTypes(resultTypes);
-// }
-
-mlir::ParseResult RgnCallValOp::parse(mlir::OpAsmParser &parser,
-                                      mlir::OperationState &result) {
-  assert(false && "unimplemented");
-}
-
-void RgnCallValOp::print(mlir::OpAsmPrinter &p) {
-  p.printGenericOp(this->getOperation());
-}
-
[…]

### 2021-09-01 test/lambdapure/compile/compile-rgn.sh

@@ -11,7 +11,7 @@ rm $1-exe.out || true
 # compile MLIR file
 lean $1 -m exe.mlir
 hask-opt exe.mlir | \
-  hask-opt  --convert-scf-to-std | hask-opt --lean-lower  | hask-opt --ptr-lower | \
+  hask-opt  --convert-scf-to-std | hask-opt --lean-lower-rgn  | hask-opt --convert-rgn-to-std | hask-opt --ptr-lower | \
   mlir-translate --mlir-to-llvmir -o exe.ll
 # | opt -S -O3 | llc -filetype=obj -o exe.o
 llvm-link exe.ll \

### 2021-09-01 hask-opt/RgnToStd.cpp
stupid linking error with switchop::build

@@ -17,6 +17,7 @@
 #include "mlir/IR/DialectImplementation.h"
 #include "mlir/IR/OpImplementation.h"
 #include "mlir/IR/Types.h"
+#include "mlir/IR/Value.h"
 #include "mlir/IR/Verifier.h"
 #include "mlir/Support/LLVM.h"
 #include "mlir/Support/LogicalResult.h"
@@ -49,6 +50,8 @@
 #include "mlir/Pass/Pass.h"
 #include "mlir/Transforms/DialectConversion.h"
 #include "mlir/Transforms/GreedyPatternRewriteDriver.h"
+#include "llvm/ADT/SmallVector.h"
+#include "llvm/IR/Instructions.h"
 
 #define DEBUG_TYPE "hask-ops"
 #include "llvm/Support/Debug.h"
@@ -84,7 +87,39 @@ struct RgnJumpValOpConversionPattern
   mlir::LogicalResult
   matchAndRewrite(RgnJumpValOp call,
                   mlir::PatternRewriter &rewriter) const override {
-    assert(false);
+
+    if (RgnValOp val = call.getFn().getDefiningOp<RgnValOp>()) {
+      rewriter.create<BranchOp>(call.getLoc(), &val.getRegion().getBlocks().front(), call.getFnArguments());
+      rewriter.inlineRegionBefore(val.getRegion(), call->getBlock());
+      rewriter.eraseOp(call);
+      return success();
+    }
+
+    if (RgnSelectOp select = call.getFn().getDefiningOp<RgnSelectOp>()) {
+      Block *defaultbb = nullptr;
+      SmallVector<int, 4> caseValues;
+      SmallVector<Block *, 4> caseDestinations;
+
+      for(int i = 0; i < select.getNumBranches(); ++i) {
+        std::pair<int, mlir::Value> branch = select.getBranch(i);
+        RgnValOp val = branch.second.getDefiningOp<RgnValOp>();
+        assert(val && "expected select of val"); 
+          
+        if (branch.first == -42) {
+          assert(defaultbb != nullptr && "can't have two defaults!");
+          defaultbb = &val.getRegion().getBlocks().front();
+        } else {
+          caseValues.push_back(branch.first);
+          caseDestinations.push_back(&val.getRegion().getBlocks().front());
+        }        
+      }
+
+      SmallVector<Value, 1> emptyArgs;
+      rewriter.replaceOpWithNewOp<SwitchOp>(call, select.getSwitcher(), defaultbb, emptyArgs,
[…]

### 2021-09-01 include/RgnDialect.h

@@ -12,6 +12,7 @@
 #include "mlir/Pass/Pass.h"
 #include "mlir/Support/LLVM.h"
 #include "llvm/BinaryFormat/Dwarf.h"
+#include <string>
 
 class RgnDialect : public mlir::Dialect {
 public:
@@ -29,6 +30,23 @@ public:
   static llvm::StringRef getOperationName() { return "rgn.select"; };
   static mlir::ParseResult parse(mlir::OpAsmParser &parser,
                                  mlir::OperationState &result);
+
+  mlir::Value getSwitcher() {
+    return this->getOperand(0);
+  }
+
+  int getNumBranches() {
+    int ans = this->getNumOperands() - 1; 
+    assert(ans > 0);
+    return ans;
+  }
+
+  std::pair<int, mlir::Value> getBranch(int i) {
+    assert(i + 1 < (int)this->getOperands().size());
+    mlir::IntegerAttr lhs = this->getOperation()->getAttrOfType<mlir::IntegerAttr>("case" + std::to_string(i));
+    mlir::Value rhs = this->getOperand(i + 1);
+    return { lhs.getInt(), rhs };
+  }
 
   static void build(mlir::OpBuilder &builder, mlir::OperationState &state,
                     mlir::Value switcher,

### 2021-09-02 hask-opt/RgnToStd.cpp
region rewriting crashes. yay.

@@ -89,34 +89,51 @@ struct RgnJumpValOpConversionPattern
                   mlir::PatternRewriter &rewriter) const override {
 
     if (RgnValOp val = call.getFn().getDefiningOp<RgnValOp>()) {
+      assert(false && "screwing up region");
       rewriter.create<BranchOp>(call.getLoc(), &val.getRegion().getBlocks().front(), call.getFnArguments());
       rewriter.inlineRegionBefore(val.getRegion(), call->getBlock());
       rewriter.eraseOp(call);
+      assert(false);
       return success();
     }
 
     if (RgnSelectOp select = call.getFn().getDefiningOp<RgnSelectOp>()) {
       Block *defaultbb = nullptr;
-      SmallVector<int, 4> caseValues;
-      SmallVector<Block *, 4> caseDestinations;
+      SmallVector<int32_t> caseValues;
+      SmallVector<Block *> caseDestinations;
+      SmallVector<RgnValOp> toErase;
 
       for(int i = 0; i < select.getNumBranches(); ++i) {
         std::pair<int, mlir::Value> branch = select.getBranch(i);
         RgnValOp val = branch.second.getDefiningOp<RgnValOp>();
-        assert(val && "expected select of val"); 
-          
+
         if (branch.first == -42) {
-          assert(defaultbb != nullptr && "can't have two defaults!");
+          assert(defaultbb == nullptr && "can't have two defaults!");
           defaultbb = &val.getRegion().getBlocks().front();
         } else {
           caseValues.push_back(branch.first);
           caseDestinations.push_back(&val.getRegion().getBlocks().front());
         }        
+        toErase.push_back(val);
+        rewriter.inlineRegionBefore(val.getRegion(), call->getBlock());
+
       }
+      if (!defaultbb) {
+        assert(caseDestinations.size());
+        defaultbb = caseDestinations[0];
+      }
+      assert(defaultbb);
 
       SmallVector<Value, 1> emptyArgs;
-      rewriter.replaceOpWithNewOp<SwitchOp>(call, select.getSwitcher(), defaultbb, emptyArgs,
+      assert(call->getUsers().empty());
+      rewriter.replaceOpWithNewOp<LLVM::SwitchOp>(call, select.getSwitcher(), defaultbb,
+        /*default operands*/emptyArgs,
         caseValues, caseDestinations);
+      rewriter.eraseOp(select);
+      for(RgnValOp v: toErase) { rewriter.eraseOp(v); }
[…]

### 2021-09-02 hask-opt/RgnToStd.cpp
get simple rgn lowering working

@@ -98,42 +98,50 @@ struct RgnJumpValOpConversionPattern
     }
 
     if (RgnSelectOp select = call.getFn().getDefiningOp<RgnSelectOp>()) {
-      Block *defaultbb = nullptr;
-      SmallVector<int32_t> caseValues;
-      SmallVector<Block *> caseDestinations;
-      SmallVector<RgnValOp> toErase;
+
+      // Block *defaultbb = nullptr;
+      SmallVector<std::pair<int, RgnValOp>> vals;
 
       for(int i = 0; i < select.getNumBranches(); ++i) {
         std::pair<int, mlir::Value> branch = select.getBranch(i);
         RgnValOp val = branch.second.getDefiningOp<RgnValOp>();
+        vals.push_back({branch.first, val});
+      };
 
-        if (branch.first == -42) {
-          assert(defaultbb == nullptr && "can't have two defaults!");
-          defaultbb = &val.getRegion().getBlocks().front();
-        } else {
-          caseValues.push_back(branch.first);
-          caseDestinations.push_back(&val.getRegion().getBlocks().front());
-        }        
-        toErase.push_back(val);
-        rewriter.inlineRegionBefore(val.getRegion(), call->getBlock());
+      Block *defaultBB =  [&]() {
+        // createBlock moves the insetion point x(
+        OpBuilder::InsertionGuard guard(rewriter);
+        Block *b = rewriter.createBlock(call->getParentRegion(), 
+          call->getParentRegion()->end(), {});
+        rewriter.setInsertionPointToStart(b);        
+        rewriter.create<ptr::PtrUnreachableOp>(rewriter.getUnknownLoc());
+        return b;
+      }();
 
-      }
-      if (!defaultbb) {
-        assert(caseDestinations.size());
-        defaultbb = caseDestinations[0];
-      }
-      assert(defaultbb);
 
-      SmallVector<Value, 1> emptyArgs;
-      assert(call->getUsers().empty());
-      rewriter.replaceOpWithNewOp<LLVM::SwitchOp>(call, select.getSwitcher(), defaultbb,
-        /*default operands*/emptyArgs,
-        caseValues, caseDestinations);
-      rewriter.eraseOp(select);
-      for(RgnValOp v: toErase) { rewriter.eraseOp(v); }
-        
-      // rewriter.create<RgnEndOp>(call.getLoc());
+      SmallVector<int32_t> caseValues;
[…]

### 2021-09-02 include/RgnDialect.h

@@ -106,7 +106,7 @@ public:
   static mlir::ParseResult parse(mlir::OpAsmParser &parser,
                                  mlir::OperationState &result);
 
-  static void build(mlir::OpBuilder &builder, mlir::OperationState &state);
+  static void build(mlir::OpBuilder &builder, mlir::OperationState &state) {}
   void print(mlir::OpAsmPrinter &p);
 };
 

### 2021-09-02 hask-opt/RgnToStd.cpp
Add rgn -> std conversion.
The lowering is janky, since I use *inlineRegionBefore* instead of using
*cloneRegonBefore* and then connecting the clones. What I should do
is to inline the rgn.val at the dominator region of all the rgn.jmp
users of a given rgn.val.

@@ -19,10 +19,13 @@
 #include "mlir/IR/Types.h"
 #include "mlir/IR/Value.h"
 #include "mlir/IR/Verifier.h"
+#include "mlir/IR/Visitors.h"
 #include "mlir/Support/LLVM.h"
 #include "mlir/Support/LogicalResult.h"
 
+#include <ios>
 #include <mlir/Parser.h>
+#include <set>
 #include <sstream>
 
 // Standard dialect
@@ -90,7 +93,9 @@ struct RgnJumpValOpConversionPattern
 
     if (RgnValOp val = call.getFn().getDefiningOp<RgnValOp>()) {
       assert(false && "screwing up region");
-      rewriter.create<BranchOp>(call.getLoc(), &val.getRegion().getBlocks().front(), call.getFnArguments());
+      rewriter.create<BranchOp>(call.getLoc(),
+                                &val.getRegion().getBlocks().front(),
+                                call.getFnArguments());
       rewriter.inlineRegionBefore(val.getRegion(), call->getBlock());
       rewriter.eraseOp(call);
       assert(false);
@@ -100,40 +105,55 @@ struct RgnJumpValOpConversionPattern
     if (RgnSelectOp select = call.getFn().getDefiningOp<RgnSelectOp>()) {
 
       // Block *defaultbb = nullptr;
-      SmallVector<std::pair<int, RgnValOp>> vals;
+      SmallVector<std::pair<int, RgnValOp>> branches;
 
-      for(int i = 0; i < select.getNumBranches(); ++i) {
+      for (int i = 0; i < select.getNumBranches(); ++i) {
         std::pair<int, mlir::Value> branch = select.getBranch(i);
         RgnValOp val = branch.second.getDefiningOp<RgnValOp>();
-        vals.push_back({branch.first, val});
+        branches.push_back({branch.first, val});
       };
 
-      Block *defaultBB =  [&]() {
+      Block *defaultBB = [&]() {
         // createBlock moves the insetion point x(
         OpBuilder::InsertionGuard guard(rewriter);
-        Block *b = rewriter.createBlock(call->getParentRegion(), 
-          call->getParentRegion()->end(), {});
-        rewriter.setInsertionPointToStart(b);        
+        Block *b = rewriter.createBlock(call->getParentRegion(),
+                                        call->getParentRegion()->end(), {});
+        rewriter.setInsertionPointToStart(b);
[…]

### 2021-09-02 hask-opt/RgnOptimization.cpp
(mis)compile jmp. Now fix it

@@ -1,0 +1,143 @@
+//===- HaskToLLVM.cpp - Hask dialect conversion ---------------*- C++ -*-===//
+//
+// This file is licensed under the Apache License v2.0 with LLVM Exceptions.
+// See https://llvm.org/LICENSE.txt for license information.
+// SPDX-License-Identifier: Apache-2.0 WITH LLVM-exception
+//
+//===----------------------------------------------------------------------===//
+
+#include "Hask/HaskDialect.h"
+#include "Hask/HaskOps.h"
+#include "Pointer/PointerDialect.h"
+#include "RgnDialect.h"
+#include "mlir/IR/Attributes.h"
+#include "mlir/IR/Block.h"
+#include "mlir/IR/BlockAndValueMapping.h"
+#include "mlir/IR/Builders.h"
+#include "mlir/IR/DialectImplementation.h"
+#include "mlir/IR/OpImplementation.h"
+#include "mlir/IR/Types.h"
+#include "mlir/IR/Verifier.h"
+#include "mlir/IR/Visitors.h"
+#include "mlir/Support/LLVM.h"
+#include "mlir/Support/LogicalResult.h"
+
+#include <mlir/Parser.h>
+#include <sstream>
+
+// Standard dialect
+#include "mlir/Dialect/Affine/IR/AffineOps.h"
+#include "mlir/Dialect/MemRef/IR/MemRef.h"
+#include "mlir/Dialect/SCF/SCF.h"
+#include "mlir/Dialect/StandardOps/IR/Ops.h"
+#include "mlir/IR/BuiltinTypes.h"
+
+// pattern matching
+#include "mlir/IR/Matchers.h"
+#include "mlir/IR/PatternMatch.h"
+
+// dilect lowering
+#include "mlir/Pass/PassRegistry.h"
+#include "mlir/Transforms/DialectConversion.h"
+// https://github.com/llvm/llvm-project/blob/80d7ac3bc7c04975fd444e9f2806e4db224f2416/mlir/examples/toy/Ch6/toyc.cpp
+#include "mlir/Conversion/AffineToStandard/AffineToStandard.h"
+#include "mlir/Conversion/SCFToStandard/SCFToStandard.h"
+#include "mlir/Conversion/StandardToLLVM/ConvertStandardToLLVM.h"
+#include "mlir/Conversion/StandardToLLVM/ConvertStandardToLLVMPass.h"
[…]

### 2021-09-02 hask-opt/RgnToStd.cpp

@@ -68,105 +68,6 @@
 namespace {
 using namespace mlir;
 
-class RgnTypeConverter : public mlir::TypeConverter {
-public:
-  // using LLVMTypeConverter::LLVMTypeConverter;
-  using TypeConverter::convertType;
-
-  RgnTypeConverter(MLIRContext *ctx) {
-
-    addConversion([](Type type) { return type; });
-  };
-};
-
-struct RgnJumpValOpConversionPattern
-    : public mlir::OpRewritePattern<RgnJumpValOp> {
-  /// We register this pattern to match every toy.transpose in the IR.
-  /// The "benefit" is used by the framework to order the patterns and process
-  /// them in order of profitability.
-  RgnJumpValOpConversionPattern(mlir::MLIRContext *context)
-      : OpRewritePattern<RgnJumpValOp>(context, /*benefit=*/1) {}
-
-  mlir::LogicalResult
-  matchAndRewrite(RgnJumpValOp call,
-                  mlir::PatternRewriter &rewriter) const override {
-
-    if (RgnValOp val = call.getFn().getDefiningOp<RgnValOp>()) {
-      assert(false && "screwing up region");
-      rewriter.create<BranchOp>(call.getLoc(),
-                                &val.getRegion().getBlocks().front(),
-                                call.getFnArguments());
-      rewriter.inlineRegionBefore(val.getRegion(), call->getBlock());
-      rewriter.eraseOp(call);
-      assert(false);
-      return success();
-    }
-
-    if (RgnSelectOp select = call.getFn().getDefiningOp<RgnSelectOp>()) {
-
-      // Block *defaultbb = nullptr;
-      SmallVector<std::pair<int, RgnValOp>> branches;
-
-      for (int i = 0; i < select.getNumBranches(); ++i) {
-        std::pair<int, mlir::Value> branch = select.getBranch(i);
-        RgnValOp val = branch.second.getDefiningOp<RgnValOp>();
-        branches.push_back({branch.first, val});
-      };
-
-      Block *defaultBB = [&]() {
-        // createBlock moves the insetion point x(
-        OpBuilder::InsertionGuard guard(rewriter);
-        Block *b = rewriter.createBlock(call->getParentRegion(),
-                                        call->getParentRegion()->end(), {});
-        rewriter.setInsertionPointToStart(b);
[…]

### 2021-09-02 hask-opt/hask-opt.cpp

@@ -93,6 +93,7 @@ int main(int argc, char **argv) {
   registerLzInterpretPass();
   registerLzLazifyPass();
   registerLowerRgnPass();
+  registerOptimizeRgnPass();
   mlir::standalone::registerHaskCanonicalizePass();
   mlir::standalone::registerWrapperWorkerPass();
 
@@ -100,7 +101,6 @@ int main(int argc, char **argv) {
   mlir::lambdapure::registerReferenceRewriterPattern();
   mlir::lambdapure::registerDestructiveUpdatePattern();
 
-  registerRgnOptPass();
   mlir::registerScfToRgnPass();  
 
   mlir::DialectRegistry registry;

### 2021-09-02 include/RgnDialect.h

@@ -193,6 +193,13 @@ public:
     state.addOperands(rgn);
   }
 
+    static void build(mlir::OpBuilder &builder, mlir::OperationState &state,
+                    mlir::Value rgn, mlir::ValueRange operands) {
+    state.addOperands(rgn);
+    state.addOperands(operands);
+  }
+
+
 
   void print(mlir::OpAsmPrinter &p);
 };
@@ -322,9 +329,8 @@ public:
 // create interface for call/jmp
 // https://mlir.llvm.org/docs/Interfaces/#attributeoperationtype-interfaces
 
-std::unique_ptr<mlir::Pass> createRgnOptPass();
-void registerRgnOptPass();
 
 std::unique_ptr<mlir::Pass> createRgnCSEPass();
 void registerRgnCSEPass();
 void registerLowerRgnPass();
+void registerOptimizeRgnPass();

### 2021-09-02 lib/Hask/LeanRgnLowering.cpp

@@ -789,9 +789,11 @@ public:
                   ConversionPatternRewriter &rewriter) const override {
 
     HaskJoinPointOp jpOp = cast<HaskJoinPointOp>(op);
-    assert(false && "lowering join point");
 
-    // TODO: replace only if name matches jump names.
+    rewriter.setInsertionPointAfter(jpOp);
+    RgnValOp rgnval = rewriter.create<RgnValOp>(jpOp.getLoc());
+    rewriter.inlineRegionBefore(jpOp.getLaterJumpedIntoRegion(),
+      rgnval.getRegion(), rgnval.getRegion().end());
 
     jpOp.getFirstRegionWithJmp().walk([&](HaskJumpOp jmp) {
       rewriter.setInsertionPoint(jmp);
@@ -799,51 +801,31 @@ public:
         return WalkResult::advance();
       }
 
-      // assert(false && "lowering join point");
-      Block *jumpTarget = &jpOp.getLaterJumpedIntoRegion().front();
-      // vvv this needs to go through type converter x(
+      // Block *jumpTarget = &jpOp.getLaterJumpedIntoRegion().front();
       rewriter.setInsertionPointAfter(jmp);
-      // rewriter.replaceOpWithNewOp<BranchOp>(jmp, jumpTarget,
-      // jmp.getOperand()); llvm::errs() << "creating a branch op: branch(" <<
-      // jumpTarget << ", " << jmp.getOperand() << ")\n"; getchar();
-      // rewriter.create<ptr::PtrBranchOp>(jmp.getLoc(),  jumpTarget,
-      // jmp.getOperand());
-      rewriter.create<BranchOp>(jmp.getLoc(), jumpTarget, jmp.getOperands());
+      rewriter.create<RgnJumpValOp>(jmp.getLoc(), rgnval, jmp.getOperands());
+      // rewriter.create<BranchOp>(jmp.getLoc(), jumpTarget, jmp.getOperands());
       rewriter.eraseOp(jmp);
 
       return WalkResult::advance();
     });
 
-    llvm::errs() << __FUNCTION__ << ":" << __LINE__ << "\n";
+        rewriter.inlineRegionBefore(jpOp.getLaterJumpedIntoRegion(),
+                                *jpOp->getParentRegion(),
+                                jpOp->getParentRegion()->end());
 
     // begin --> fstRegion
     rewriter.setInsertionPointToEnd(jpOp->getBlock());
[…]

### 2021-09-02 lib/RgnDialect.cpp

@@ -219,107 +219,107 @@ void RgnJumpSymOp::print(mlir::OpAsmPrinter &p) {
 // ==== REGION OPTIMIZATION ====
 // ==== REGION OPTIMIZATION ====
 
-struct PatternCallValKnownRegion : public mlir::OpRewritePattern<RgnCallValOp> {
-  /// We register this pattern to match every toy.transpose in the IR.
-  /// The "benefit" is used by the framework to order the patterns and process
-  /// them in order of profitability.
-  PatternCallValKnownRegion(mlir::MLIRContext *context)
-      : OpRewritePattern<RgnCallValOp>(context, /*benefit=*/1) {}
+// struct PatternCallValKnownRegion : public mlir::OpRewritePattern<RgnCallValOp> {
+//   /// We register this pattern to match every toy.transpose in the IR.
+//   /// The "benefit" is used by the framework to order the patterns and process
+//   /// them in order of profitability.
+//   PatternCallValKnownRegion(mlir::MLIRContext *context)
+//       : OpRewritePattern<RgnCallValOp>(context, /*benefit=*/1) {}
 
-  mlir::LogicalResult
-  matchAndRewrite(RgnCallValOp call,
-                  mlir::PatternRewriter &rewriter) const override {
-    RgnValOp rgnval = call.getFn().getDefiningOp<RgnValOp>();
-    if (!rgnval) {
-      return mlir::failure();
-    }
+//   mlir::LogicalResult
+//   matchAndRewrite(RgnCallValOp call,
+//                   mlir::PatternRewriter &rewriter) const override {
+//     RgnValOp rgnval = call.getFn().getDefiningOp<RgnValOp>();
+//     if (!rgnval) {
+//       return mlir::failure();
+//     }
 
-    // code from:
-    // https://github.com/llvm/llvm-project/blob/9a11c70c1856f4e801d0863c552c754f28110237/mlir/lib/Conversion/SCFToStandard/SCFToStandard.cpp#L411
-    auto loc = call.getLoc();
-    auto *condBlock = rewriter.getInsertionBlock();
-    auto opPosition = rewriter.getInsertionPoint();
-    auto *remainingOpsBlock = rewriter.splitBlock(condBlock, opPosition);
+//     // code from:
+//     // https://github.com/llvm/llvm-project/blob/9a11c70c1856f4e801d0863c552c754f28110237/mlir/lib/Conversion/SCFToStandard/SCFToStandard.cpp#L411
[…]

### 2021-09-02 test/lambdapure/compile/compile-lean.sh

@@ -21,13 +21,13 @@ llvm-link exe.ll \
   -S -o exe-linked.ll
 opt exe-linked.ll -passes=bitcast-call-converter -S -o exe-linked-nobitcast.ll
 # opt exe-linked.ll  -S -o exe-linked-nobitcast.ll
-opt -always-inline -O3  exe-linked-nobitcast.ll -S  -o exe-linked-o3.ll
-opt -always-inline -O3  exe-linked-o3.ll -S  -o exe-linked-o3-2.ll
+opt -always-inline -O2  exe-linked-nobitcast.ll -S  -o exe-linked-o3.ll
+opt -always-inline -O2  exe-linked-o3.ll -S  -o exe-linked-o3-2.ll
 mv exe-linked-o3-2.ll exe-linked-o3.ll
 opt -verify exe-linked-o3.ll
 echo "@@@@HACK: REMOVING TAIL ANNOTATIONS!"
 sed -i "s/musttail/tail/g" exe-linked-o3.ll
-llc --relocation-model=static -O3 -march=x86-64 -filetype=obj exe-linked-o3.ll -o exe.o
+llc --relocation-model=static -O1 -march=x86-64 -filetype=obj exe-linked-o3.ll -o exe.o
 
 # leancpp: undefined reference to lean_name_eq
 # `l_Lean_Syntax_isOfKind':

### 2021-09-02 test/lambdapure/compile/compile-rgn.sh

@@ -8,10 +8,11 @@ rm $1-exe.out || true
 # lean -c fails if relative path walks upward. eg. lean -c ../exe.c -o foo
 (lean $1 -c exe-ref.c && clang -I /home/bollu/work/lean4/build/stage0/include  -O2 -S  -emit-llvm exe-ref.c -o exe-lean-ref.ll) || true
 
+# hask-opt --convert-rgn-to-std | hask-opt --ptr-lower --convert-std-to-llvm
 # compile MLIR file
 lean $1 -m exe.mlir
 hask-opt exe.mlir | \
-  hask-opt  --convert-scf-to-std | hask-opt --lean-lower-rgn  | hask-opt --convert-rgn-to-std | hask-opt --ptr-lower | \
+  hask-opt  --convert-scf-to-std | hask-opt --lean-lower-rgn  | hask-opt --convert-rgn-to-std | hask-opt --convert-std-to-llvm --ptr-lower | \
   mlir-translate --mlir-to-llvmir -o exe.ll
 # | opt -S -O3 | llc -filetype=obj -o exe.o
 llvm-link exe.ll \
@@ -21,19 +22,19 @@ llvm-link exe.ll \
   -S -o exe-linked.ll
 opt exe-linked.ll -passes=bitcast-call-converter -S -o exe-linked-nobitcast.ll
 # opt exe-linked.ll  -S -o exe-linked-nobitcast.ll
-opt -always-inline -O3  exe-linked-nobitcast.ll -S  -o exe-linked-o3.ll
-opt -always-inline -O3  exe-linked-o3.ll -S  -o exe-linked-o3-2.ll
+opt -always-inline -O1  exe-linked-nobitcast.ll -S  -o exe-linked-o3.ll
+opt -always-inline -O1  exe-linked-o3.ll -S  -o exe-linked-o3-2.ll
 mv exe-linked-o3-2.ll exe-linked-o3.ll
 opt -verify exe-linked-o3.ll
 echo "@@@@HACK: REMOVING TAIL ANNOTATIONS!"
 sed -i "s/musttail/tail/g" exe-linked-o3.ll
-llc --relocation-model=static -O3 -march=x86-64 -filetype=obj exe-linked-o3.ll -o exe.o
+llc --relocation-model=static -O1 -march=x86-64 -filetype=obj exe-linked-o3.ll -o exe.o
 
 # leancpp: undefined reference to lean_name_eq
[…]

### 2021-09-02 test/lambdapure/compile/validate-rgn-lean.sh

@@ -1,0 +1,27 @@
+#!/usr/bin/env bash
+
+
+# https://stackoverflow.com/a/23161454/5305365
+SCRIPTDIR="${BASH_SOURCE[0]}";
+SCRIPTDIR=$(dirname $SCRIPTDIR)
+
+
+set -e
+set -o xtrace
+
+# For whatever reason this does not work with relative paths.
+lean $1 -c exe-ref.c
+leanc -O3 exe-ref.c -o exe-ref.out
+leanc -O3 exe-ref.c -S -emit-llvm -o exe-ref-leanc.ll
+clang++ -U LEAN_MULTI_THREAD -I/home/bollu/work/lean4/build/stage1/include \
+   -O3 -S  -emit-llvm  exe-ref.c -o exe-ref.ll
+rm ref.txt || true
+./exe-ref.out > ref.txt 
+
+rm ours.txt || true
+$SCRIPTDIR/compile-rgn.sh $1
+./exe.out > ours.txt 
+diff ref.txt ours.txt
+
+time ./exe-ref.out
+time ./exe.out

### 2021-09-02 hask-opt/RgnToStd.cpp
fix miscompile

@@ -123,7 +123,7 @@ struct LowerRgnPass : public Pass {
         Region *parent = jump->getParentRegion();
         llvm::errs() << __FILE__ << ":" << __LINE__ << "\n";
 
-        if (branch.first == -42) {
+        if (branch.first == 42) {
           assert(i == (int)select.getNumBranches() - 1);
           assert(!defaultBB && "can't have two default blocks");
           Region &r = v.getRegion();

### 2021-09-02 lib/Hask/LeanRgnLowering.cpp

@@ -810,9 +810,9 @@ public:
       return WalkResult::advance();
     });
 
-        rewriter.inlineRegionBefore(jpOp.getLaterJumpedIntoRegion(),
-                                *jpOp->getParentRegion(),
-                                jpOp->getParentRegion()->end());
+    rewriter.inlineRegionBefore(jpOp.getLaterJumpedIntoRegion(),
+                            *jpOp->getParentRegion(),
+                            jpOp->getParentRegion()->end());
 
     // begin --> fstRegion
     rewriter.setInsertionPointToEnd(jpOp->getBlock());
@@ -1559,7 +1559,8 @@ public:
     for (int i = 0; i < caseop.getNumAlts(); ++i) {
       rhss.push_back(
           genCaseAlt(caseop, rands[0], i, caseop.getNumAlts(), rewriter));
-      lhss.push_back(i);
+      Optional<int> lhs = caseop.getAltLHS(i);
+      lhss.push_back(lhs ? *lhs : 42);
     }
 
     assert(rhss.size() > 0);

### 2021-09-02 hask-opt/RgnToStd.cpp
jmp miscompile on qsort

@@ -68,7 +68,6 @@
 namespace {
 using namespace mlir;
 
-
 struct LowerRgnPass : public Pass {
   LowerRgnPass() : Pass(mlir::TypeID::get<LowerRgnPass>()){};
   StringRef getName() const override { return "LowerRgnPass"; }
@@ -81,41 +80,53 @@ struct LowerRgnPass : public Pass {
   }
 
   void
-  rewriteJump(RgnJumpValOp jump, mlir::IRRewriter &rewriter, 
+  rewriteJump(RgnJumpValOp jump, mlir::IRRewriter &rewriter,
               std::vector<std::pair<RgnValOp, Block *>> &inlineVals) const {
 
     static std::set<RgnValOp> seen;
+    static std::set<RgnJumpValOp> jumps;
+    assert(!jumps.count(jump));
+    jumps.insert(jump);
 
     if (RgnValOp val = jump.getFn().getDefiningOp<RgnValOp>()) {
+      llvm::errs() << "\n===jump(rgn)===\n";
+      llvm::errs() << "rgnval:\n\t" << val << "\n";
+      llvm::errs() << "parent:\n\t" << jump << "\n";
+      llvm::errs() << "\n===\n";
+
+      assert(!seen.count(val));
+      seen.insert(val);
+
+      llvm::errs() << __FILE__ << ":" << __LINE__ << "\n";
       rewriter.setInsertionPointAfter(jump);
+      assert(val.getRegion().getBlocks().size() > 0);
       rewriter.create<BranchOp>(jump.getLoc(),
                                 &val.getRegion().getBlocks().front(),
                                 jump.getFnArguments());
-      rewriter.inlineRegionBefore(val.getRegion(), 
-        *jump->getParentRegion(), jump->getParentRegion()->end());
+      rewriter.inlineRegionBefore(val.getRegion(), *jump->getParentRegion(),
+                                  jump->getParentRegion()->end());
       rewriter.eraseOp(jump);
+      llvm::errs() << __FILE__ << ":" << __LINE__ << "\n";
       return;
     }
 
     if (RgnSelectOp select = jump.getFn().getDefiningOp<RgnSelectOp>()) {
-
 
       Block *defaultBB = nullptr;
       SmallVector<int32_t> caseLhss;
       SmallVector<Block *> caseRhss;
       SmallVector<mlir::Value> defaultOperands;
 
-
-    llvm::errs() << __FILE__ << ":" << __LINE__ << "\n";
[…]

### 2021-09-02 include/RgnDialect.h

@@ -14,6 +14,8 @@
 #include "llvm/BinaryFormat/Dwarf.h"
 #include <string>
 
+static const int RGN_DIALECT_DEFAULT_CASE_MAGIC = 42;
+
 class RgnDialect : public mlir::Dialect {
 public:
   explicit RgnDialect(mlir::MLIRContext *ctx);

### 2021-09-02 lib/Hask/LeanRgnLowering.cpp

@@ -1434,7 +1434,7 @@ public:
       Optional<int> lhs = caseop.getAltLHS(i);
       lhss.push_back(
           lhs ? *lhs
-              : 42); // HACK HACK HACK: use "-42" to denote default case.
+              : RGN_DIALECT_DEFAULT_CASE_MAGIC); // HACK HACK HACK: use "-42" to denote default case.
     }
 
     assert(rhss.size() > 0);
@@ -1560,7 +1560,7 @@ public:
       rhss.push_back(
           genCaseAlt(caseop, rands[0], i, caseop.getNumAlts(), rewriter));
       Optional<int> lhs = caseop.getAltLHS(i);
-      lhss.push_back(lhs ? *lhs : 42);
+      lhss.push_back(lhs ? *lhs : RGN_DIALECT_DEFAULT_CASE_MAGIC);
     }
 
     assert(rhss.size() > 0);

### 2021-09-02 hask-opt/RgnToStd.cpp
use new clone with entry interface

@@ -85,7 +85,7 @@ struct LowerRgnPass : public Pass {
 
     static std::set<RgnValOp> seen;
     static std::set<RgnJumpValOp> jumps;
-    assert(!jumps.count(jump));
+    assert(true || !jumps.count(jump));
     jumps.insert(jump);
 
     if (RgnValOp val = jump.getFn().getDefiningOp<RgnValOp>()) {
@@ -94,17 +94,18 @@ struct LowerRgnPass : public Pass {
       llvm::errs() << "parent:\n\t" << jump << "\n";
       llvm::errs() << "\n===\n";
 
-      assert(!seen.count(val));
+      assert(true || !seen.count(val));
       seen.insert(val);
 
       llvm::errs() << __FILE__ << ":" << __LINE__ << "\n";
       rewriter.setInsertionPointAfter(jump);
       assert(val.getRegion().getBlocks().size() > 0);
-      rewriter.create<BranchOp>(jump.getLoc(),
-                                &val.getRegion().getBlocks().front(),
-                                jump.getFnArguments());
-      rewriter.inlineRegionBefore(val.getRegion(), *jump->getParentRegion(),
+      Block *newEntry = rewriter.cloneRegionBeforeAndReturnEntry(val.getRegion(), *jump->getParentRegion(),
                                   jump->getParentRegion()->end());
+
+      rewriter.create<BranchOp>(jump.getLoc(),
+                                newEntry,
+                                jump.getFnArguments());
       rewriter.eraseOp(jump);
       llvm::errs() << __FILE__ << ":" << __LINE__ << "\n";
       return;
@@ -128,7 +129,7 @@ struct LowerRgnPass : public Pass {
         llvm::errs() << "select:\n\t" << select << "\n";
         llvm::errs() << "parent:\n\t" << jump << "\n";
         llvm::errs() << "\n===\n";
-        assert(!seen.count(v));
+        assert(true || !seen.count(v));
         seen.insert(v);
 
         Region *parent = jump->getParentRegion();
@@ -139,8 +140,9 @@ struct LowerRgnPass : public Pass {
           assert(i == (int)select.getNumBranches() - 1);
           assert(!defaultBB && "can't have two default blocks");
           Region &r = v.getRegion();
-          defaultBB = &r.front();
-          rewriter.inlineRegionBefore(r, *parent, parent->end());
+          // defaultBB = &r.front();
[…]

### 2021-09-03 hask-opt/RgnToStd.cpp
add refactored version that uses the clone stuff

@@ -89,10 +89,10 @@ struct LowerRgnPass : public Pass {
     jumps.insert(jump);
 
     if (RgnValOp val = jump.getFn().getDefiningOp<RgnValOp>()) {
-      llvm::errs() << "\n===jump(rgn)===\n";
-      llvm::errs() << "rgnval:\n\t" << val << "\n";
-      llvm::errs() << "parent:\n\t" << jump << "\n";
-      llvm::errs() << "\n===\n";
+      // llvm::errs() << "\n===jump(rgn)===\n";
+      // llvm::errs() << "rgnval:\n\t" << val << "\n";
+      // llvm::errs() << "parent:\n\t" << jump << "\n";
+      // llvm::errs() << "\n===\n";
 
       assert(true || !seen.count(val));
       seen.insert(val);
@@ -124,11 +124,11 @@ struct LowerRgnPass : public Pass {
         RgnValOp v = branch.second.getDefiningOp<RgnValOp>();
         assert(v && "expected select(rgnval)");
 
-        llvm::errs() << "\n===jump(select(rgn))===\n";
-        llvm::errs() << "rgnval:\n\t" << v << "\n";
-        llvm::errs() << "select:\n\t" << select << "\n";
-        llvm::errs() << "parent:\n\t" << jump << "\n";
-        llvm::errs() << "\n===\n";
+        // llvm::errs() << "\n===jump(select(rgn))===\n";
+        // llvm::errs() << "rgnval:\n\t" << v << "\n";
+        // llvm::errs() << "select:\n\t" << select << "\n";
+        // llvm::errs() << "parent:\n\t" << jump << "\n";
+        // llvm::errs() << "\n===\n";
         assert(true || !seen.count(v));
         seen.insert(v);
 

### 2021-09-03 test/lambdapure/compile/bench/speedup-rgn.py

@@ -1,0 +1,258 @@
+#!/usr/bin/env python3
+
+import argparse
+import glob
+import os
+import json
+import sys
+import os
+import subprocess
+import os.path
+import matplotlib
+import matplotlib.pyplot as plt
+import numpy as np
+import json
+import sys
+import subprocess
+import numpy as np
+from scipy.stats import mstats
+
+
+PARSER = argparse.ArgumentParser(description='Speedup b/w MLIR and LEAN')
+PARSER.add_argument("--data", help="regenerate plotting data", action="store_true")
+PARSER.add_argument("--plot", help="make the plot", action="store_true")
+PARSER.add_argument('--out', metavar='o', type=str,
+                    help='path to dump output', default=os.path.basename(__file__).replace(".py", ".json"))
+PARSER.add_argument('--nruns', type=int,
+                    help='number of runs to average', default=10)
+ARGS = PARSER.parse_args()
+
+G_BASELINE = "../baseline-lean.sh"
+G_OURS = "../run-lean.sh"
+G_FPATHS = []
+G_FPATHS.append("binarytrees-int.lean")
+G_FPATHS.append("binarytrees.lean")
+G_FPATHS.append("const_fold.lean")
+G_FPATHS.append("deriv.lean")
+G_FPATHS.append("filter.lean")
+G_FPATHS.append("qsort.lean") # miscompile because of jmp!
+G_FPATHS.append("rbmap_checkpoint.lean")
+G_FPATHS.append("unionfind.lean")
+G_NFILES = len(G_FPATHS)
+
+
+# Color palette
+light_gray = "#cacaca"
+dark_gray = "#827b7b"
+light_blue = "#a6cee3"
+dark_blue = "#1f78b4"
+light_green = "#b2df8a"
+dark_green = "#33a02c"
+light_red = "#fb9a99"
+dark_red = "#e31a1c"
+
+
+def log(*ARGS, **kwargs):
+    print(*ARGS, file=sys.stderr, **kwargs)
+
+def os_system_synch(path):
+    return subprocess.check_output(path, shell=True)
+
+def sh(path):
+    proc = subprocess.Popen(path, shell=True, stderr=subprocess.PIPE, stdout=subprocess.PIPE)
+    output = proc.stdout.read()
+    err = proc.stderr.read()
+    return (output.decode(), err.decode())
+
+
+def run_data():
+    ds = []
+    for (i, fpath) in enumerate(G_FPATHS):
[…]

### 2021-09-03 test/lambdapure/compile/bench/speedup-rgn.py
disabling the LEAN optimization pipeline has no effect.,?

@@ -69,11 +69,11 @@ def run_data():
     ds = []
     for (i, fpath) in enumerate(G_FPATHS):
         datum = {"file": str(fpath) }
-        LEANPATH="/home/bollu/work/lean4-contrib/build/stage1/bin/lean"
-        LEANCPATH="/home/bollu/work/lean4-contrib/build/stage1/bin/leanc"
+        LEAN_OPTIMIZED_PATH="/home/bollu/work/lean4-contrib/build/stage1/bin/lean"
+        LEANC_OPTIMIZED_PATH="/home/bollu/work/lean4-contrib/build/stage1/bin/leanc"
         os_system_synch(f"rm exe-ref.out || true")
-        os_system_synch(f"{LEANPATH} {fpath} -c exe-ref.c")
-        os_system_synch(f"{LEANCPATH} exe-ref.c -O3 -o exe-ref.out")
+        os_system_synch(f"{LEAN_OPTIMIZED_PATH} {fpath} -c exe-ref.c")
+        os_system_synch(f"{LEANC_OPTIMIZED_PATH} exe-ref.c -O3 -o exe-ref.out")
         # x = run_with_output("perf stat ./exe-ref.out 1>/dev/null")
         # proc = subprocess.Popen("perf stat ./exe-ref.out", shell=True, stderr=subprocess.PIPE, stdout=subprocess.PIPE)
         datum["theirs-out"] = []
@@ -87,7 +87,8 @@ def run_data():
         os_system_synch(f"rm exe.ll  || true")
         os_system_synch(f"rm exe-linked.ll  || true")
         os_system_synch(f"rm exe.o  || true")
-        os_system_synch(f"lean {fpath} -m exe.mlir")
+        LEAN_NOOPTIMIZE_PATH="/home/bollu/work/lean4/build/stage1/bin/lean"
+        os_system_synch(f"{LEAN_NOOPTIMIZE_PATH} {fpath} -m exe.mlir")
         os_system_synch("hask-opt exe.mlir --convert-scf-to-std --lean-lower-rgn --convert-rgn-to-std --convert-std-to-llvm --ptr-lower | \
                 mlir-translate --mlir-to-llvmir -o exe.ll")
         os_system_synch("llvm-link " + 

### 2021-09-03 test/lambdapure/compile/bench/speedup-rgn.py
add region plots -- have a day to make it better :)

@@ -224,7 +224,7 @@ def plot():
   # Y-Axis Label
   #
   # Use a horizontal label for improved readability.
-  ax.set_ylabel('Speedup over leanc', rotation='horizontal', position = (1, 1.1),
+  ax.set_ylabel('Speedup over leanc (via regions for optimization)', rotation='horizontal', position = (1, 1.1),
       horizontalalignment='left', verticalalignment='bottom', fontsize=8)
 
   # Add some text for labels, title and custom x-axis tick labels, etc.

### 2021-09-04 test/lambdapure/compile/bench/speedup-rgn.tex
commit before paper

% This file was created with tikzplotlib v0.9.12.
\begin{tikzpicture}

\definecolor{color0}{rgb}{0.847058823529412,0.105882352941176,0.376470588235294}
\definecolor{color1}{rgb}{0.984313725490196,0.603921568627451,0.6}

\begin{axis}[
legend cell align={left},
legend columns=100,
legend style={fill opacity=0.8, draw opacity=1, text opacity=1, at={(1,1.3)}, draw=none},
tick align=outside,
tick pos=left,
x grid style={white!69.0196078431373!black},
xmin=-0.72, xmax=8.52,
xtick style={color=black},
xtick={0,1,2,3,4,5,6,7,8},
xticklabel style={rotate=15.0},
xticklabels={
  binarytrees-int,
  binarytrees,
  const\_fold,
  deriv,
  filter,
  qsort,
  rbmap\_checkpoint,
  unionfind,
  geomean
},
y grid style={white!69.0196078431373!black},
ylabel style={rotate=-90.0},
ylabel={Speedup over λpure simplifier},
ymin=0, ymax=1.2075,
ytick style={color=black}
]
\draw[draw=none,fill=color0] (axis cs:-0.3,0) rectangle (axis cs:-0.1,1.15);
\addlegendimage{ybar,ybar legend,draw=none,fill=color0}
\addlegendentry{rgn simplifier}

\draw[draw=none,fill=color0] (axis cs:0.7,0) rectangle (axis cs:0.9,1);
\draw[draw=none,fill=color0] (axis cs:1.7,0) rectangle (axis cs:1.9,0.92);
\draw[draw=none,fill=color0] (axis cs:2.7,0) rectangle (axis cs:2.9,0.86);
\draw[draw=none,fill=color0] (axis cs:3.7,0) rectangle (axis cs:3.9,0.99);
\draw[draw=none,fill=color0] (axis cs:4.7,0) rectangle (axis cs:4.9,1);
\draw[draw=none,fill=color0] (axis cs:5.7,0) rectangle (axis cs:5.9,0.99);
\draw[draw=none,fill=color0] (axis cs:6.7,0) rectangle (axis cs:6.9,1.01);
\draw[draw=color1,fill=color1] (axis cs:7.7,0) rectangle (axis cs:7.9,0.99);
\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:-0.1,0) rectangle (axis cs:0.1,1.12);
\addlegendimage{ybar,ybar legend,draw=none,fill=white!61.9607843137255!black}
\addlegendentry{none}

\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:0.9,0) rectangle (axis cs:1.1,1);
\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:1.9,0) rectangle (axis cs:2.1,0.92);
\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:2.9,0) rectangle (axis cs:3.1,0.98);
\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:3.9,0) rectangle (axis cs:4.1,1.03);
\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:4.9,0) rectangle (axis cs:5.1,1.02);
\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:5.9,0) rectangle (axis cs:6.1,0.99);
\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:6.9,0) rectangle (axis cs:7.1,1);
\draw[draw=white!79.2156862745098!black,fill=white!79.2156862745098!black] (axis cs:7.9,0) rectangle (axis cs:8.1,1.01);
\draw (axis cs:-0.3,1.15) ++(0pt,1pt) node[
  scale=0.4165,
  anchor=south,
  text=black,
  rotate=0.0
]{1.15};
\draw (axis cs:0.7,1) ++(0pt,1pt) node[
  scale=0.4165,
  anchor=south,
  text=black,
  rotate=0.0
]{1.0};
\draw (axis cs:1.7,0.92) ++(0pt,1pt) node[
  scale=0.4165,
  anchor=south,
  text=black,
  rotate=0.0
]{0.92};
\draw (axis cs:2.7,0.86) ++(0pt,1pt) node[
  scale=0.4165,
  anchor=south,
  text=black,
  rotate=0.0
]{0.86};
\draw (axis cs:3.7,0.99) ++(0pt,1pt) node[
  scale=0.4165,
  anchor=south,
  text=black,
  rotate=0.0
]{0.99};
\draw (axis cs:4.7,1) ++(0pt,1pt) node[
  scale=0.4165,
  anchor=south,
  text=black,
  rotate=0.0
]{1.0};
\draw (axis cs:5.7,0.99) ++(0pt,1pt) node[
  scale=0.4165,
  anchor=south,
  text=black,
  rotate=0.0
]{0.99};
\draw (axis cs:6.7,1.01) ++(0pt,1pt) node[
  scale=0.4165,
  anchor=south,
  text=black,
  rotate=0.0
]{1.01};
\draw (axis cs:7.7,0.99) ++(0pt,1pt) node[
  scale=0.4165,
  anchor=south,
  text=black,
  rotate=0.0
]{0.99};
\end{axis}

\end{tikzpicture}

### 2021-09-04 hask-opt/LeanOptimization.cpp

@@ -64,80 +64,100 @@ namespace standalone {
 
 namespace {
 
-
-struct OptimizeRgnPass : public Pass {
-  OptimizeRgnPass() : Pass(mlir::TypeID::get<OptimizeRgnPass>()){};
-  StringRef getName() const override { return "OptimizeRgnPass"; }
+struct OptimizeLeanPass : public Pass {
+  OptimizeLeanPass() : Pass(mlir::TypeID::get<OptimizeLeanPass>()){};
+  StringRef getName() const override { return "OptimizeLeanPass"; }
 
   std::unique_ptr<Pass> clonePass() const override {
-    auto newInst = std::make_unique<OptimizeRgnPass>(
-        *static_cast<const OptimizeRgnPass *>(this));
+    auto newInst = std::make_unique<OptimizeLeanPass>(
+        *static_cast<const OptimizeLeanPass *>(this));
     newInst->copyOptionValuesFrom(this);
     return newInst;
   }
 
-  void runPatterns(ConversionTarget &target,
-                   mlir::OwningRewritePatternList &patterns) {
-
-    // ::llvm::DebugFlag = true;
-
-    if (failed(mlir::applyPartialConversion(getOperation(), target,
-                                            std::move(patterns)))) {
-      llvm::errs() << "===Hask lowering failed at Conversion===\n";
-      // getOperation()->print(llvm::errs(),
-      // mlir::OpPrintingFlags().printGenericOpForm());
-      llvm::errs() << "\n===\n";
-      signalPassFailure();
-      ::llvm::DebugFlag = false;
-      return;
-    };
-
-    if (failed(mlir::verify(getOperation()))) {
-      llvm::errs() << "===Hask lowering failed at Verification===\n";
-      // getOperation()->print(llvm::errs());
-      llvm::errs() << "\n===\n";
-      signalPassFailure();
-      ::llvm::DebugFlag = false;
-      return;
-    }
-
-    ::llvm::DebugFlag = false;
-  }
-
   void runOnOperation() override {
+    
+    mlir::IRRewriter rewriter(getOperation()->getContext());
     int known_constructors = 0;
     getOperation()->walk([&](ProjectionOp proj) {
-        HaskConstructOp cons = proj.getOperand().getDefiningOp<HaskConstructOp>();
-        if (!cons) { return WalkResult::advance(); }
-        llvm::errs() << "#optimizable constructs |" << ++known_constructors << "|\n";
-        assert(false);
[…]

### 2021-09-04 hask-opt/LeanPipeline.cpp

@@ -1,0 +1,1308 @@
+//===- HaskToLLVM.cpp - Hask dialect conversion ---------------*- C++ -*-===//
+//
+// This file is licensed under the Apache License v2.0 with LLVM Exceptions.
+// See https://llvm.org/LICENSE.txt for license information.
+// SPDX-License-Identifier: Apache-2.0 WITH LLVM-exception
+//
+//===----------------------------------------------------------------------===//
+
+#include "Hask/HaskDialect.h"
+#include "Hask/HaskOps.h"
+#include "Pointer/PointerDialect.h"
+#include "RgnDialect.h"
+#include "mlir/IR/Attributes.h"
+#include "mlir/IR/Block.h"
+#include "mlir/IR/BlockAndValueMapping.h"
+#include "mlir/IR/Builders.h"
+#include "mlir/IR/DialectImplementation.h"
+#include "mlir/IR/OpImplementation.h"
+#include "mlir/IR/Types.h"
+#include "mlir/IR/Verifier.h"
+#include "mlir/IR/Visitors.h"
+#include "mlir/Support/LLVM.h"
+#include "mlir/Support/LogicalResult.h"
+
+#include <mlir/Parser.h>
+#include <sstream>
+
+// Standard dialect
+#include "mlir/Dialect/Affine/IR/AffineOps.h"
+#include "mlir/Dialect/MemRef/IR/MemRef.h"
+#include "mlir/Dialect/SCF/SCF.h"
+#include "mlir/Dialect/StandardOps/IR/Ops.h"
+#include "mlir/IR/BuiltinTypes.h"
+
+// pattern matching
+#include "mlir/IR/Matchers.h"
+#include "mlir/IR/PatternMatch.h"
+
+// dilect lowering
+#include "mlir/Pass/PassRegistry.h"
+#include "mlir/Transforms/DialectConversion.h"
+// https://github.com/llvm/llvm-project/blob/80d7ac3bc7c04975fd444e9f2806e4db224f2416/mlir/examples/toy/Ch6/toyc.cpp
+#include "mlir/Conversion/AffineToStandard/AffineToStandard.h"
+#include "mlir/Conversion/SCFToStandard/SCFToStandard.h"
+#include "mlir/Conversion/StandardToLLVM/ConvertStandardToLLVM.h"
+#include "mlir/Conversion/StandardToLLVM/ConvertStandardToLLVMPass.h"
[…]

### 2021-09-04 hask-opt/RgnCSEPass.cpp

@@ -70,6 +70,7 @@ struct SimpleOperationInfo : public llvm::DenseMapInfo<Operation *> {
         rhs == getTombstoneKey() || rhs == getEmptyKey())
       return false;
     if (isa<RgnValOp>(lhs) && isa<RgnValOp>(rhs)) {
+      return false;
       assert(false  && "comparing regions");
     }
     return OperationEquivalence::isEquivalentTo(const_cast<Operation *>(lhsC),
@@ -163,10 +164,14 @@ llvm::hash_code rgnValHashOp(RgnCSE::ScopedMapTy &knownValues, Operation *op) {
 bool RgnCSE::isRgnValOpEqual(ScopedMapTy &knownValues, RgnValOp us, RgnValOp other) {
   ScopedMapTy::ScopeTy scope(knownValues);
 
-  llvm::errs() << "comparing region vals\n";
+  // llvm::errs() << "comparing region vals\n";
 
   Region &r = us.getRegion();
   Region &s = other.getRegion();
+  if (!(r.getBlocks().size() == 1 && s.getBlocks().size() == 1)) {
+    llvm::errs() << "TODO: investigate this!\n";
+    return false;
+  }
   assert(r.getBlocks().size() == 1);
   assert(s.getBlocks().size() == 1);
   Block &b = r.getBlocks().front();
@@ -236,8 +241,10 @@ LogicalResult RgnCSE::simplifyOperation(std::set<RgnValOp >&seen,
 
   if (RgnValOp rgnval = dyn_cast<RgnValOp>(op)) {
     Region &rgn = rgnval.getRegion();
+
     if (rgn.getBlocks().size() == 0 || rgn.getBlocks().size() > 2) {
-      assert(false && "found illegal rgn val op");
+      // assert(false && "found illegal rgn val op");
+      llvm::errs() << "TODO: investigate!\n";
       return failure();
     }
 

### 2021-09-04 hask-opt/hask-opt.cpp

@@ -93,7 +93,8 @@ int main(int argc, char **argv) {
   registerLzInterpretPass();
   registerLzLazifyPass();
   registerLowerRgnPass();
-  registerOptimizeRgnPass();
+  registerOptimizeLeanPass();
+  registerLeanPipelinePass();
   mlir::standalone::registerHaskCanonicalizePass();
   mlir::standalone::registerWrapperWorkerPass();
 

### 2021-09-04 lib/Hask/HaskDialect.cpp

@@ -7,7 +7,7 @@
 
 #include "Hask/HaskDialect.h"
 #include "Hask/HaskOps.h"
-
+#include "RgnDialect.h"
 // includes
 #include "mlir/IR/Attributes.h"
 #include "mlir/IR/Builders.h"
@@ -213,7 +213,12 @@ void HaskInlinerInterface::handleTerminator(
     valuesToRepl[0].replaceAllUsesWith(returnOp.getOperand());
     return;
   }
+  if (auto returnOp = mlir::dyn_cast<RgnReturnOp>(op)) {
+    valuesToRepl[0].replaceAllUsesWith(returnOp.getOperand());
+    return;
+  }
 
+  llvm::errs() << "UNKNOWN RETURN: |" << *op << "|\n";
   assert(false && "unknown return operation");
 }
 

### 2021-09-04 test/lambdapure/compile/bench/speedup-rgn.py

@@ -16,7 +16,7 @@ import sys
 import subprocess
 import numpy as np
 from scipy.stats import mstats
-
+import tikzplotlib
 
 PARSER = argparse.ArgumentParser(description='Speedup b/w MLIR and LEAN')
 PARSER.add_argument("--data", help="regenerate plotting data", action="store_true")
@@ -43,14 +43,15 @@ G_NFILES = len(G_FPATHS)
 
 # Color palette
 light_gray = "#cacaca"
-dark_gray = "#827b7b"
+# dark_gray = "#827b7b"
+dark_gray = "#9E9E9E"
 light_blue = "#a6cee3"
 dark_blue = "#1f78b4"
 light_green = "#b2df8a"
 dark_green = "#33a02c"
 light_red = "#fb9a99"
-dark_red = "#e31a1c"
-
+# dark_red = "#e31a1c"
+dark_red = "#D81B60"
 
 def log(*ARGS, **kwargs):
     print(*ARGS, file=sys.stderr, **kwargs)
@@ -65,17 +66,33 @@ def sh(path):
     return (output.decode(), err.decode())
 
 
+
 def run_data():
     ds = []
     for (i, fpath) in enumerate(G_FPATHS):
+        # with simpcase: 850fd84e43407ed647837652b6442e143199abb0
         datum = {"file": str(fpath) }
-        LEAN_OPTIMIZED_PATH="/home/bollu/work/lean4-contrib/build/stage1/bin/lean"
-        LEANC_OPTIMIZED_PATH="/home/bollu/work/lean4-contrib/build/stage1/bin/leanc"
-        os_system_synch(f"rm exe-ref.out || true")
-        os_system_synch(f"{LEAN_OPTIMIZED_PATH} {fpath} -c exe-ref.c")
-        os_system_synch(f"{LEANC_OPTIMIZED_PATH} exe-ref.c -O3 -o exe-ref.out")
-        # x = run_with_output("perf stat ./exe-ref.out 1>/dev/null")
-        # proc = subprocess.Popen("perf stat ./exe-ref.out", shell=True, stderr=subprocess.PIPE, stdout=subprocess.PIPE)
+        LEAN_ENABLE_SIMPCASE_PATH="/home/bollu/work/lean4-contrib/build/stage1/bin/lean"
+        os_system_synch(f"{LEAN_ENABLE_SIMPCASE_PATH} {fpath} -m exe.mlir")
[…]

### 2021-10-06 test/lambdapure/compile/bench/speedup-rgn.tex
checkpoint rgn stuff

@@
 \definecolor{color0}{rgb}{0.847058823529412,0.105882352941176,0.376470588235294}
-\definecolor{color1}{rgb}{0.984313725490196,0.603921568627451,0.6}
 
@@
 x grid style={white!69.0196078431373!black},
-xmin=-0.72, xmax=8.52,
+xmin=-0.32, xmax=0.12,
 xtick style={color=black},
-xtick={0,1,2,3,4,5,6,7,8},
+xtick={0},
 xticklabel style={rotate=15.0},
-xticklabels={
-  binarytrees-int,
-  binarytrees,
-  const\_fold,
-  deriv,
-  filter,
-  qsort,
-  rbmap\_checkpoint,
-  unionfind,
-  geomean
-},
+xticklabels={binarytrees-int},
 y grid style={white!69.0196078431373!black},
 ylabel style={rotate=-90.0},
-ylabel={Speedup over λpure simplifier},
-ymin=0, ymax=1.2075,
+ylabel={Speedup over no simplifier},
+ymin=0, ymax=1.0395,
 ytick style={color=black}
 ]
-\draw[draw=none,fill=color0] (axis cs:-0.3,0) rectangle (axis cs:-0.1,1.15);
+\draw[draw=none,fill=color0] (axis cs:-0.3,0) rectangle (axis cs:-0.1,0.87);
 \addlegendimage{ybar,ybar legend,draw=none,fill=color0}
-\addlegendentry{rgn simplifier}
+\addlegendentry{λpure over none}
 
-\draw[draw=none,fill=color0] (axis cs:0.7,0) rectangle (axis cs:0.9,1);
-\draw[draw=none,fill=color0] (axis cs:1.7,0) rectangle (axis cs:1.9,0.92);
-\draw[draw=none,fill=color0] (axis cs:2.7,0) rectangle (axis cs:2.9,0.86);
-\draw[draw=none,fill=color0] (axis cs:3.7,0) rectangle (axis cs:3.9,0.99);
-\draw[draw=none,fill=color0] (axis cs:4.7,0) rectangle (axis cs:4.9,1);
-\draw[draw=none,fill=color0] (axis cs:5.7,0) rectangle (axis cs:5.9,0.99);
-\draw[draw=none,fill=color0] (axis cs:6.7,0) rectangle (axis cs:6.9,1.01);
-\draw[draw=color1,fill=color1] (axis cs:7.7,0) rectangle (axis cs:7.9,0.99);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:-0.1,0) rectangle (axis cs:0.1,1.12);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:-0.1,0) rectangle (axis cs:0.1,0.99);
 \addlegendimage{ybar,ybar legend,draw=none,fill=white!61.9607843137255!black}
-\addlegendentry{none}
+\addlegendentry{rgn over none}
 
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:0.9,0) rectangle (axis cs:1.1,1);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:1.9,0) rectangle (axis cs:2.1,0.92);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:2.9,0) rectangle (axis cs:3.1,0.98);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:3.9,0) rectangle (axis cs:4.1,1.03);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:4.9,0) rectangle (axis cs:5.1,1.02);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:5.9,0) rectangle (axis cs:6.1,0.99);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:6.9,0) rectangle (axis cs:7.1,1);
-\draw[draw=white!79.2156862745098!black,fill=white!79.2156862745098!black] (axis cs:7.9,0) rectangle (axis cs:8.1,1.01);
-\draw (axis cs:-0.3,1.15) ++(0pt,1pt) node[
-  scale=0.4165,
+\draw (axis cs:-0.3,0.87) ++(0pt,1pt) node[
+  scale=0.347,
   anchor=south,
@@
   rotate=0.0
-]{1.15};
-\draw (axis cs:0.7,1) ++(0pt,1pt) node[
-  scale=0.4165,
-  anchor=south,
-  text=black,
-  rotate=0.0
-]{1.0};
-\draw (axis cs:1.7,0.92) ++(0pt,1pt) node[
-  scale=0.4165,
-  anchor=south,
-  text=black,
-  rotate=0.0
-]{0.92};
-\draw (axis cs:2.7,0.86) ++(0pt,1pt) node[
-  scale=0.4165,
-  anchor=south,
-  text=black,
-  rotate=0.0
-]{0.86};
-\draw (axis cs:3.7,0.99) ++(0pt,1pt) node[
-  scale=0.4165,
-  anchor=south,
-  text=black,
-  rotate=0.0
-]{0.99};
-\draw (axis cs:4.7,1) ++(0pt,1pt) node[
-  scale=0.4165,
-  anchor=south,
-  text=black,
-  rotate=0.0
-]{1.0};
-\draw (axis cs:5.7,0.99) ++(0pt,1pt) node[
-  scale=0.4165,
-  anchor=south,
-  text=black,
-  rotate=0.0
-]{0.99};
-\draw (axis cs:6.7,1.01) ++(0pt,1pt) node[
-  scale=0.4165,
-  anchor=south,
-  text=black,
-  rotate=0.0
-]{1.01};
-\draw (axis cs:7.7,0.99) ++(0pt,1pt) node[
-  scale=0.4165,
-  anchor=south,
-  text=black,
-  rotate=0.0
-]{0.99};
+]{0.87};
 \end{axis}

### 2021-10-06 include/Hask/HaskDialect.h

@@ -282,4 +282,7 @@ void registerFuseRefcountPass();
 } // namespace standalone
 } // namespace mlir
 
+void registerOptimizeLeanPass();
+void registerLeanPipelinePass();
+
 #endif // STANDALONE_STANDALONEDIALECT_H

### 2021-10-06 include/RgnDialect.h

@@ -335,4 +335,3 @@ public:
 std::unique_ptr<mlir::Pass> createRgnCSEPass();
 void registerRgnCSEPass();
 void registerLowerRgnPass();
-void registerOptimizeRgnPass();

### 2021-10-06 test/lambdapure/compile/bench/speedup-rgn.py

@@ -24,20 +24,21 @@ PARSER.add_argument("--plot", help="make the plot", action="store_true")
 PARSER.add_argument('--out', metavar='o', type=str,
                     help='path to dump output', default=os.path.basename(__file__).replace(".py", ".json"))
 PARSER.add_argument('--nruns', type=int,
-                    help='number of runs to average', default=10)
+                    # help='number of runs to average', default=10)
+                    help='number of runs to average', default=1)
 ARGS = PARSER.parse_args()
 
 G_BASELINE = "../baseline-lean.sh"
 G_OURS = "../run-lean.sh"
 G_FPATHS = []
 G_FPATHS.append("binarytrees-int.lean")
-G_FPATHS.append("binarytrees.lean")
-G_FPATHS.append("const_fold.lean")
-G_FPATHS.append("deriv.lean")
-G_FPATHS.append("filter.lean")
-G_FPATHS.append("qsort.lean") # miscompile because of jmp!
-G_FPATHS.append("rbmap_checkpoint.lean")
-G_FPATHS.append("unionfind.lean")
+# G_FPATHS.append("binarytrees.lean")
+# G_FPATHS.append("const_fold.lean")
+# G_FPATHS.append("deriv.lean")
+# G_FPATHS.append("filter.lean")
+# G_FPATHS.append("qsort.lean") # miscompile because of jmp!
+# G_FPATHS.append("rbmap_checkpoint.lean")
+# G_FPATHS.append("unionfind.lean")
 G_NFILES = len(G_FPATHS)
 
 
@@ -65,113 +66,160 @@ def sh(path):
     err = proc.stderr.read()
     return (output.decode(), err.decode())
 
+# returns datum
+def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fpath, out_index, perf_index):
+    # simpcase ENABLED: 850fd84e43407ed647837652b6442e143199abb0
+    LEAN_PATH="/home/bollu/work/lean4/build/stage1/bin/lean"
+    if case_simpl_enabled:
+        lean_opts = "-Dcompiler.caseSimpl=true"
+    else:
+        lean_opts = "-Dcompiler.caseSimpl=false"
+
+    os_system_synch(f"{LEAN_PATH} {fpath} {lean_opts}  -m exe.mlir")
+
+    if False:
[…]

### 2021-10-06 test/lambdapure/compile/bench/speedup.py

@@ -210,8 +210,8 @@ def plot():
 
   fig, ax = plt.subplots()
   # rects1 = ax.bar(x - width/2, baselines, width, label='Baseline', color = light_blue)
-  rects2 = ax.bar(x + width/2, optims, width, label='Optimised', color = dark_blue)
-  rects2[-1].set_color(light_blue) # color geomean separately.
+  rects2 = ax.bar(x + width/2, optims, width, label='Optimised', color = light_blue)
+  rects2[-1].set_color(dark_blue) # color geomean separately.
 
 
 

### 2021-10-06 test/lambdapure/compile/compile-optimize-rgn.sh

@@ -1,0 +1,49 @@
+#!/usr/bin/env bash
+
+set -e
+set -o xtrace
+
+rm $1-exe.out || true
+
+# lean -c fails if relative path walks upward. eg. lean -c ../exe.c -o foo
+(lean $1 -c exe-ref.c && clang -I /home/bollu/work/lean4/build/stage0/include  -O2 -S  -emit-llvm exe-ref.c -o exe-lean-ref.ll) || true
+
+# hask-opt --convert-rgn-to-std | hask-opt --ptr-lower --convert-std-to-llvm
+# compile MLIR file
+lean $1 -m exe.mlir
+hask-opt exe.mlir | \
+  hask-opt  --convert-scf-to-std | hask-opt --lean-lower-rgn  | hask-opt --lean-opt | \
+  hask-opt --convert-rgn-to-std | hask-opt --convert-std-to-llvm --ptr-lower | \
+  mlir-translate --mlir-to-llvmir -o exe.ll
+# | opt -S -O3 | llc -filetype=obj -o exe.o
+llvm-link exe.ll \
+  /home/bollu/work/lz/lean-linking-incantations/lib-includes/library.ll \
+  /home/bollu/work/lz/lean-linking-incantations/lib-runtime/runtime.ll \
+  /home/bollu/work/lz/lean-linking-incantations/lean-shell.ll \
+  -S -o exe-linked.ll
+opt exe-linked.ll -passes=bitcast-call-converter -S -o exe-linked-nobitcast.ll
+# opt exe-linked.ll  -S -o exe-linked-nobitcast.ll
+opt -always-inline -O1  exe-linked-nobitcast.ll -S  -o exe-linked-o3.ll
+opt -always-inline -O1  exe-linked-o3.ll -S  -o exe-linked-o3-2.ll
+mv exe-linked-o3-2.ll exe-linked-o3.ll
+opt -verify exe-linked-o3.ll
+echo "@@@@HACK: REMOVING TAIL ANNOTATIONS!"
+sed -i "s/musttail/tail/g" exe-linked-o3.ll
+llc --relocation-model=static -O1 -march=x86-64 -filetype=obj exe-linked-o3.ll -o exe.o
+
+# leancpp: undefined reference to lean_name_eq
+# `l_Lean_Syntax_isOfKind':
+
+# Lean: lean_name_hash 
+c++ -O1 -U LEAN_MULTI_THREAD -I/home/bollu/work/lean4/build/stage1/include \
+    exe.o \
[…]

### 2021-10-08 test/lambdapure/compile/bench/speedup-rgn.tex
still no perf difference

@@
 x grid style={white!69.0196078431373!black},
-xmin=-0.32, xmax=0.12,
+xmin=-0.57, xmax=5.37,
 xtick style={color=black},
-xtick={0},
+xtick={0,1,2,3,4,5},
 xticklabel style={rotate=15.0},
-xticklabels={binarytrees-int},
+xticklabels={binarytrees-int,binarytrees,filter,qsort,rbmap\_checkpoint,unionfind},
 y grid style={white!69.0196078431373!black},
@@
 ylabel={Speedup over no simplifier},
-ymin=0, ymax=1.0395,
+ymin=0, ymax=1.071,
 ytick style={color=black}
 ]
-\draw[draw=none,fill=color0] (axis cs:-0.3,0) rectangle (axis cs:-0.1,0.87);
+\draw[draw=none,fill=color0] (axis cs:-0.3,0) rectangle (axis cs:-0.1,0.99);
 \addlegendimage{ybar,ybar legend,draw=none,fill=color0}
@@
 
+\draw[draw=none,fill=color0] (axis cs:0.7,0) rectangle (axis cs:0.9,1);
+\draw[draw=none,fill=color0] (axis cs:1.7,0) rectangle (axis cs:1.9,1.02);
+\draw[draw=none,fill=color0] (axis cs:2.7,0) rectangle (axis cs:2.9,0.98);
+\draw[draw=none,fill=color0] (axis cs:3.7,0) rectangle (axis cs:3.9,0.98);
+\draw[draw=none,fill=color0] (axis cs:4.7,0) rectangle (axis cs:4.9,0.97);
 \draw[draw=none,fill=white!61.9607843137255!black] (axis cs:-0.1,0) rectangle (axis cs:0.1,0.99);
@@
 
-\draw (axis cs:-0.3,0.87) ++(0pt,1pt) node[
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:0.9,0) rectangle (axis cs:1.1,1);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:1.9,0) rectangle (axis cs:2.1,1.02);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:2.9,0) rectangle (axis cs:3.1,1);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:3.9,0) rectangle (axis cs:4.1,1);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:4.9,0) rectangle (axis cs:5.1,1);
+\draw (axis cs:-0.3,0.99) ++(0pt,1pt) node[
   scale=0.347,
@@
   rotate=0.0
-]{0.87};
+]{0.99};
+\draw (axis cs:0.7,1) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{1.0};
+\draw (axis cs:1.7,1.02) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{1.02};
+\draw (axis cs:2.7,0.98) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{0.98};
+\draw (axis cs:3.7,0.98) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{0.98};
+\draw (axis cs:4.7,0.97) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{0.97};
 \end{axis}

### 2021-10-08 test/lambdapure/compile/bench/speedup-rgn.py

@@ -25,20 +25,20 @@ PARSER.add_argument('--out', metavar='o', type=str,
                     help='path to dump output', default=os.path.basename(__file__).replace(".py", ".json"))
 PARSER.add_argument('--nruns', type=int,
                     # help='number of runs to average', default=10)
-                    help='number of runs to average', default=1)
+                    help='number of runs to average', default=10)
 ARGS = PARSER.parse_args()
 
 G_BASELINE = "../baseline-lean.sh"
 G_OURS = "../run-lean.sh"
 G_FPATHS = []
 G_FPATHS.append("binarytrees-int.lean")
-# G_FPATHS.append("binarytrees.lean")
+G_FPATHS.append("binarytrees.lean")
 # G_FPATHS.append("const_fold.lean")
 # G_FPATHS.append("deriv.lean")
-# G_FPATHS.append("filter.lean")
-# G_FPATHS.append("qsort.lean") # miscompile because of jmp!
-# G_FPATHS.append("rbmap_checkpoint.lean")
-# G_FPATHS.append("unionfind.lean")
+G_FPATHS.append("filter.lean")
+G_FPATHS.append("qsort.lean") # miscompile because of jmp!
+G_FPATHS.append("rbmap_checkpoint.lean")
+G_FPATHS.append("unionfind.lean")
 G_NFILES = len(G_FPATHS)
 
 
@@ -68,6 +68,7 @@ def sh(path):
 
 # returns datum
 def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fpath, out_index, perf_index):
+    print(f"running f{fpath} case_simpl_enabled={case_simpl_enabled} rgn_optimization_enabled={rgn_optimization_enabled}")
     # simpcase ENABLED: 850fd84e43407ed647837652b6442e143199abb0
     LEAN_PATH="/home/bollu/work/lean4/build/stage1/bin/lean"
     if case_simpl_enabled:

### 2021-10-08 test/lambdapure/compile/bench/speedup-rgn.py
still no perf delta

@@ -58,6 +58,7 @@ def log(*ARGS, **kwargs):
     print(*ARGS, file=sys.stderr, **kwargs)
 
 def os_system_synch(path):
+    print(f"\t$ {path}")
     return subprocess.check_output(path, shell=True)
 
 def sh(path):

### 2021-10-08 test/lambdapure/compile/bench/speedup-rgn.tex
okay, there is SOME speedup at O1

--- 2021-10-06 test/lambdapure/compile/bench/speedup-rgn.tex
@@
 x grid style={white!69.0196078431373!black},
-xmin=-0.32, xmax=0.12,
+xmin=-0.62, xmax=6.42,
 xtick style={color=black},
-xtick={0},
+xtick={0,1,2,3,4,5,6},
 xticklabel style={rotate=15.0},
-xticklabels={binarytrees-int},
+xticklabels={binarytrees-int,const\_fold,deriv,filter,qsort,rbmap\_checkpoint,unionfind},
 y grid style={white!69.0196078431373!black},
@@
 ylabel={Speedup over no simplifier},
-ymin=0, ymax=1.0395,
+ymin=0, ymax=1.3335,
 ytick style={color=black}
 ]
-\draw[draw=none,fill=color0] (axis cs:-0.3,0) rectangle (axis cs:-0.1,0.87);
+\draw[draw=none,fill=color0] (axis cs:-0.3,0) rectangle (axis cs:-0.1,0.98);
 \addlegendimage{ybar,ybar legend,draw=none,fill=color0}
@@
 
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:-0.1,0) rectangle (axis cs:0.1,0.99);
+\draw[draw=none,fill=color0] (axis cs:0.7,0) rectangle (axis cs:0.9,1.01);
+\draw[draw=none,fill=color0] (axis cs:1.7,0) rectangle (axis cs:1.9,1.27);
+\draw[draw=none,fill=color0] (axis cs:2.7,0) rectangle (axis cs:2.9,1.1);
+\draw[draw=none,fill=color0] (axis cs:3.7,0) rectangle (axis cs:3.9,1);
+\draw[draw=none,fill=color0] (axis cs:4.7,0) rectangle (axis cs:4.9,0.98);
+\draw[draw=none,fill=color0] (axis cs:5.7,0) rectangle (axis cs:5.9,0.96);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:-0.1,0) rectangle (axis cs:0.1,1);
 \addlegendimage{ybar,ybar legend,draw=none,fill=white!61.9607843137255!black}
@@
 
-\draw (axis cs:-0.3,0.87) ++(0pt,1pt) node[
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:0.9,0) rectangle (axis cs:1.1,1.06);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:1.9,0) rectangle (axis cs:2.1,0.99);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:2.9,0) rectangle (axis cs:3.1,1.14);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:3.9,0) rectangle (axis cs:4.1,0.97);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:4.9,0) rectangle (axis cs:5.1,0.99);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:5.9,0) rectangle (axis cs:6.1,0.96);
+\draw (axis cs:-0.3,0.98) ++(0pt,1pt) node[
   scale=0.347,
@@
   rotate=0.0
-]{0.87};
+]{0.98};
+\draw (axis cs:0.7,1.01) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{1.01};
+\draw (axis cs:1.7,1.27) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{1.27};
+\draw (axis cs:2.7,1.1) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{1.1};
+\draw (axis cs:3.7,1) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{1.0};
+\draw (axis cs:4.7,0.98) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{0.98};
+\draw (axis cs:5.7,0.96) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{0.96};
 \end{axis}

### 2021-10-08 test/lambdapure/compile/bench/speedup-rgn.py

@@ -32,9 +32,9 @@ G_BASELINE = "../baseline-lean.sh"
 G_OURS = "../run-lean.sh"
 G_FPATHS = []
 G_FPATHS.append("binarytrees-int.lean")
-G_FPATHS.append("binarytrees.lean")
-# G_FPATHS.append("const_fold.lean")
-# G_FPATHS.append("deriv.lean")
+# G_FPATHS.append("binarytrees.lean")
+G_FPATHS.append("const_fold.lean")
+G_FPATHS.append("deriv.lean")
 G_FPATHS.append("filter.lean")
 G_FPATHS.append("qsort.lean") # miscompile because of jmp!
 G_FPATHS.append("rbmap_checkpoint.lean")
@@ -90,14 +90,14 @@ def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fp
               "exe.ll " + 
               "/home/bollu/work/lz/lean-linking-incantations/lib-includes/library.ll " + 
               "/home/bollu/work/lz/lean-linking-incantations/lib-runtime/runtime.ll " +
-              "| opt -passes=bitcast-call-converter  | opt --always-inline -O3 -S -o exe-linked.ll")
+              "| opt -passes=bitcast-call-converter  | opt --always-inline -O1 -S -o exe-linked.ll")
     os_system_synch("sed -i s/musttail/tail/g exe-linked.ll")
-    os_system_synch("opt -O3 exe-linked.ll -o exe-linked-o3.ll")
+    os_system_synch("opt -O1 exe-linked.ll -o exe-linked-o3.ll")
     os_system_synch("mv exe-linked-o3.ll exe-linked.ll")
     print("@@@ HACK: converting muttail to tail because of llc miscompile@@@")
     os_system_synch("sed -i s/musttail/tail/g exe-linked.ll")
-    os_system_synch("llc -O2 -march=x86-64 -filetype=obj exe-linked.ll -o exe.o")
-    os_system_synch(f"c++ -O3 -D LEAN_MULTI_THREAD -I/home/bollu/work/lean4/build/stage1/include \
+    os_system_synch("llc -O1 -march=x86-64 -filetype=obj exe-linked.ll -o exe.o")
+    os_system_synch(f"c++ -O1 -D LEAN_MULTI_THREAD -I/home/bollu/work/lean4/build/stage1/include \
         exe.o \
         /home/bollu/work/lz/lean-linking-incantations/lean-shell.o \
[…]

### 2021-10-08 test/lambdapure/compile/bench/speedup-rgn.tex
nope, no speedup

--- 2021-10-06 test/lambdapure/compile/bench/speedup-rgn.tex
@@
 ylabel={Speedup over no simplifier},
-ymin=0, ymax=1.0395,
+ymin=0, ymax=1.071,
 ytick style={color=black}
 ]
-\draw[draw=none,fill=color0] (axis cs:-0.3,0) rectangle (axis cs:-0.1,0.87);
+\draw[draw=none,fill=color0] (axis cs:-0.3,0) rectangle (axis cs:-0.1,0.96);
 \addlegendimage{ybar,ybar legend,draw=none,fill=color0}
@@
 
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:-0.1,0) rectangle (axis cs:0.1,0.99);
+\draw[draw=none,fill=color0] (axis cs:0.7,0) rectangle (axis cs:0.9,1);
+\draw[draw=none,fill=color0] (axis cs:1.7,0) rectangle (axis cs:1.9,1);
+\draw[draw=none,fill=color0] (axis cs:2.7,0) rectangle (axis cs:2.9,0.99);
+\draw[draw=none,fill=color0] (axis cs:3.7,0) rectangle (axis cs:3.9,1);
+\draw[draw=none,fill=color0] (axis cs:4.7,0) rectangle (axis cs:4.9,0.99);
+\draw[draw=none,fill=color0] (axis cs:5.7,0) rectangle (axis cs:5.9,1.01);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:-0.1,0) rectangle (axis cs:0.1,0.98);
 \addlegendimage{ybar,ybar legend,draw=none,fill=white!61.9607843137255!black}
@@
 
-\draw (axis cs:-0.3,0.87) ++(0pt,1pt) node[
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:0.9,0) rectangle (axis cs:1.1,1.02);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:1.9,0) rectangle (axis cs:2.1,0.98);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:2.9,0) rectangle (axis cs:3.1,0.99);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:3.9,0) rectangle (axis cs:4.1,0.99);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:4.9,0) rectangle (axis cs:5.1,1);
+\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:5.9,0) rectangle (axis cs:6.1,1.01);
+\draw (axis cs:-0.3,0.96) ++(0pt,1pt) node[
   scale=0.347,
@@
   rotate=0.0
-]{0.87};
+]{0.96};
+\draw (axis cs:0.7,1) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{1.0};
+\draw (axis cs:1.7,1) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{1.0};
+\draw (axis cs:2.7,0.99) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{0.99};
+\draw (axis cs:3.7,1) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{1.0};
+\draw (axis cs:4.7,0.99) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{0.99};
+\draw (axis cs:5.7,1.01) ++(0pt,1pt) node[
+  scale=0.347,
+  anchor=south,
+  text=black,
+  rotate=0.0
+]{1.01};
 \end{axis}

### 2021-10-08 test/lambdapure/compile/bench/speedup-rgn.py
give up, no perf delta even on O0
New plan of attack: take small examples where I expect a delta and see
what's going wrong there.

@@ -31,14 +31,14 @@ ARGS = PARSER.parse_args()
 G_BASELINE = "../baseline-lean.sh"
 G_OURS = "../run-lean.sh"
 G_FPATHS = []
-G_FPATHS.append("binarytrees-int.lean")
-# G_FPATHS.append("binarytrees.lean")
-G_FPATHS.append("const_fold.lean")
-G_FPATHS.append("deriv.lean")
-G_FPATHS.append("filter.lean")
-G_FPATHS.append("qsort.lean") # miscompile because of jmp!
-G_FPATHS.append("rbmap_checkpoint.lean")
-G_FPATHS.append("unionfind.lean")
+# G_FPATHS.append("binarytrees-int.lean")
+G_FPATHS.append("binarytrees.lean")
+# G_FPATHS.append("const_fold.lean")
+# G_FPATHS.append("deriv.lean")
+# G_FPATHS.append("filter.lean")
+# G_FPATHS.append("qsort.lean") # miscompile because of jmp!
+# G_FPATHS.append("rbmap_checkpoint.lean")
+# G_FPATHS.append("unionfind.lean")
 G_NFILES = len(G_FPATHS)
 
 
@@ -59,6 +59,7 @@ def log(*ARGS, **kwargs):
 
 def os_system_synch(path):
     print(f"\t$ {path}")
+    # input("continue>")
     return subprocess.check_output(path, shell=True)
 
 def sh(path):
@@ -90,14 +91,14 @@ def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fp
               "exe.ll " + 
               "/home/bollu/work/lz/lean-linking-incantations/lib-includes/library.ll " + 
               "/home/bollu/work/lz/lean-linking-incantations/lib-runtime/runtime.ll " +
-              "| opt -passes=bitcast-call-converter  | opt --always-inline -O1 -S -o exe-linked.ll")
+              "| opt -passes=bitcast-call-converter  | opt --always-inline -O0 -S -o exe-linked.ll")
     os_system_synch("sed -i s/musttail/tail/g exe-linked.ll")
-    os_system_synch("opt -O1 exe-linked.ll -o exe-linked-o3.ll")
+    os_system_synch("opt -O0 exe-linked.ll -o exe-linked-o3.ll")
     os_system_synch("mv exe-linked-o3.ll exe-linked.ll")
[…]

### 2021-10-16 test/lambdapure/compile/1-debug.sh
have example that shows different IR

@@ -1,0 +1,13 @@
+#!/usr/bin/env bash
+make -C ~/work/lean4/build/release -j
+echo ""
+echo "NO-CASE-SIMPL"
+~/work/lean4/build/release/stage1/bin/lean case-branches.lean -Dcompiler.caseSimpl=false -m case-simpl-NO.mlir
+~/work/lean4/build/release/stage1/bin/lean case-branches.lean -Dcompiler.caseSimpl=false -c case-simpl-NO.c
+echo ""
+echo "YES-CASE-SIMPL"
+~/work/lean4/build/release/stage1/bin/lean case-branches.lean -Dcompiler.caseSimpl=true -m case-simpl-YES.mlir
+~/work/lean4/build/release/stage1/bin/lean case-branches.lean -Dcompiler.caseSimpl=false -c case-simpl-YES.c
+diff case-simpl-NO.mlir case-simpl-YES.mlir
+
+

### 2021-10-16 test/lambdapure/compile/bench/speedup-rgn.py
redid plots to be based on program with variable problem sizes

@@ -31,11 +31,11 @@ ARGS = PARSER.parse_args()
 G_BASELINE = "../baseline-lean.sh"
 G_OURS = "../run-lean.sh"
 G_FPATHS = []
-# G_FPATHS.append("binarytrees-int.lean")
-G_FPATHS.append("binarytrees.lean")
+G_FPATHS.append(("binarytrees-int.lean", 20))
+# G_FPATHS.append(("binarytrees.lean", 20))
 # G_FPATHS.append("const_fold.lean")
 # G_FPATHS.append("deriv.lean")
-# G_FPATHS.append("filter.lean")
+# G_FPATHS.append(("filter.lean", 50000))
 # G_FPATHS.append("qsort.lean") # miscompile because of jmp!
 # G_FPATHS.append("rbmap_checkpoint.lean")
 # G_FPATHS.append("unionfind.lean")
@@ -69,10 +69,10 @@ def sh(path):
     return (output.decode(), err.decode())
 
 # returns datum
-def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fpath, out_index, perf_index):
+def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fpath, run_args, out_index, perf_index):
     print(f"running f{fpath} case_simpl_enabled={case_simpl_enabled} rgn_optimization_enabled={rgn_optimization_enabled}")
     # simpcase ENABLED: 850fd84e43407ed647837652b6442e143199abb0
-    LEAN_PATH="/home/bollu/work/lean4/build/stage1/bin/lean"
+    LEAN_PATH="/home/bollu/work/lean4/build/release/stage1/bin/lean"
     if case_simpl_enabled:
         lean_opts = "-Dcompiler.caseSimpl=true"
     else:
@@ -98,17 +98,20 @@ def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fp
     print("@@@ HACK: converting muttail to tail because of llc miscompile@@@")
     os_system_synch("sed -i s/musttail/tail/g exe-linked.ll")
     os_system_synch("llc -O0 -march=x86-64 -filetype=obj exe-linked.ll -o exe.o")
+    os_system_synch("rm exe-ref.out")
     os_system_synch(f"c++ -O0 -D LEAN_MULTI_THREAD -I/home/bollu/work/lean4/build/stage1/include \
         exe.o \
[…]

### 2021-10-16 test/lambdapure/compile/bench/speedup.py

@@ -30,14 +30,14 @@ ARGS = PARSER.parse_args()
 G_BASELINE = "../baseline-lean.sh"
 G_OURS = "../run-lean.sh"
 G_FPATHS = []
-G_FPATHS.append("binarytrees-int.lean")
-G_FPATHS.append("binarytrees.lean")
-G_FPATHS.append("const_fold.lean")
-G_FPATHS.append("deriv.lean")
-G_FPATHS.append("filter.lean")
-G_FPATHS.append("qsort.lean")
-G_FPATHS.append("rbmap_checkpoint.lean")
-G_FPATHS.append("unionfind.lean")
+G_FPATHS.append(("binarytrees-int.lean", 20))
+G_FPATHS.append(("binarytrees.lean", 20))
+# G_FPATHS.append("const_fold.lean")
+# G_FPATHS.append("deriv.lean")
+G_FPATHS.append(("filter.lean", 50000))
+# G_FPATHS.append("qsort.lean")
+# G_FPATHS.append("rbmap_checkpoint.lean")
+# G_FPATHS.append("unionfind.lean")
 G_NFILES = len(G_FPATHS)
 
 
@@ -56,6 +56,7 @@ def log(*ARGS, **kwargs):
     print(*ARGS, file=sys.stderr, **kwargs)
 
 def os_system_synch(path):
+    print(f"\t$ {path}")
     return subprocess.check_output(path, shell=True)
 
 def sh(path):
@@ -67,26 +68,28 @@ def sh(path):
 
 def run_data():
     ds = []
-    for (i, fpath) in enumerate(G_FPATHS):
+    LEAN_PATH="/home/bollu/work/lean4/build/release/stage1/bin/lean"
+    LEANC_PATH="/home/bollu/work/lean4/build/release/stage1/bin/leanc"
+    for (i, (fpath, problem_size)) in enumerate(G_FPATHS):
         datum = {"file": str(fpath) }
-        os_system_synch(f"rm exe-ref.out")
-        os_system_synch(f"lean {fpath} -c exe-ref.c")
-        os_system_synch(f"leanc exe-ref.c -O3 -o exe-ref.out")
+        os_system_synch(f"rm exe-ref.out || true")
+        os_system_synch(f"{LEAN_PATH} {fpath} -c exe-ref.c")
+        os_system_synch(f"{LEANC_PATH} exe-ref.c -O3 -o exe-ref.out")
[…]

### 2021-10-17 test/lambdapure/compile/bench/speedup-rgn.tex
add speedup plotting code

--- 2021-10-08 test/lambdapure/compile/bench/speedup-rgn.tex
@@
 
-\definecolor{color0}{rgb}{0.847058823529412,0.105882352941176,0.376470588235294}
+\definecolor{color0}{rgb}{0.12156862745098,0.470588235294118,0.705882352941177}
+\definecolor{color1}{rgb}{0.847058823529412,0.105882352941176,0.376470588235294}
 
@@
 x grid style={white!69.0196078431373!black},
-xmin=-0.57, xmax=5.37,
+xmin=-0.82, xmax=6.22,
 xtick style={color=black},
-xtick={0,1,2,3,4,5},
+xtick={0,1,2,3,4,5,6},
 xticklabel style={rotate=15.0},
-xticklabels={binarytrees-int,binarytrees,filter,qsort,rbmap\_checkpoint,unionfind},
+xticklabels={binarytrees-int,binarytrees,const\_fold,deriv,filter,rbmap\_checkpoint,unionfind},
 y grid style={white!69.0196078431373!black},
 ylabel style={rotate=-90.0},
-ylabel={Speedup over no simplifier},
-ymin=0, ymax=1.071,
+ylabel={Runtimes},
+ymin=0, ymax=1.491,
 ytick style={color=black}
 ]
-\draw[draw=none,fill=color0] (axis cs:-0.3,0) rectangle (axis cs:-0.1,0.99);
+\draw[draw=none,fill=color0] (axis cs:-0.3,0) rectangle (axis cs:-0.1,1);
 \addlegendimage{ybar,ybar legend,draw=none,fill=color0}
-\addlegendentry{λpure over none}
+\addlegendentry{λpure simplifier}
 
-\draw[draw=none,fill=color0] (axis cs:0.7,0) rectangle (axis cs:0.9,1);
-\draw[draw=none,fill=color0] (axis cs:1.7,0) rectangle (axis cs:1.9,1.02);
-\draw[draw=none,fill=color0] (axis cs:2.7,0) rectangle (axis cs:2.9,0.98);
-\draw[draw=none,fill=color0] (axis cs:3.7,0) rectangle (axis cs:3.9,0.98);
-\draw[draw=none,fill=color0] (axis cs:4.7,0) rectangle (axis cs:4.9,0.97);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:-0.1,0) rectangle (axis cs:0.1,0.99);
-\addlegendimage{ybar,ybar legend,draw=none,fill=white!61.9607843137255!black}
-\addlegendentry{rgn over none}
+\draw[draw=none,fill=color0] (axis cs:0.7,0) rectangle (axis cs:0.9,1.01);
+\draw[draw=none,fill=color0] (axis cs:1.7,0) rectangle (axis cs:1.9,1.3);
+\draw[draw=none,fill=color0] (axis cs:2.7,0) rectangle (axis cs:2.9,0.78);
+\draw[draw=none,fill=color0] (axis cs:3.7,0) rectangle (axis cs:3.9,0.97);
+\draw[draw=none,fill=color0] (axis cs:4.7,0) rectangle (axis cs:4.9,1.15);
+\draw[draw=none,fill=color0] (axis cs:5.7,0) rectangle (axis cs:5.9,0.85);
+\draw[draw=none,fill=color1] (axis cs:-0.5,0) rectangle (axis cs:-0.3,1);
+\addlegendimage{ybar,ybar legend,draw=none,fill=color1}
+\addlegendentry{rgn simplifier}
 
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:0.9,0) rectangle (axis cs:1.1,1);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:1.9,0) rectangle (axis cs:2.1,1.02);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:2.9,0) rectangle (axis cs:3.1,1);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:3.9,0) rectangle (axis cs:4.1,1);
-\draw[draw=none,fill=white!61.9607843137255!black] (axis cs:4.9,0) rectangle (axis cs:5.1,1);
-\draw (axis cs:-0.3,0.99) ++(0pt,1pt) node[
-  scale=0.347,
-  anchor=south,
-  text=black,
-  rotate=0.0
-]{0.99};
-\draw (axis cs:0.7,1) ++(0pt,1pt) node[
-  scale=0.347,
-  anchor=south,
+\draw[draw=none,fill=color1] (axis cs:0.5,0) rectangle (axis cs:0.7,1);
+\draw[draw=none,fill=color1] (axis cs:1.5,0) rectangle (axis cs:1.7,1.42);
+\draw[draw=none,fill=color1] (axis cs:2.5,0) rectangle (axis cs:2.7,0.96);
+\draw[draw=none,fill=color1] (axis cs:3.5,0) rectangle (axis cs:3.7,1.12);
+\draw[draw=none,fill=color1] (axis cs:4.5,0) rectangle (axis cs:4.7,1.01);
+\draw[draw=none,fill=color1] (axis cs:5.5,0) rectangle (axis cs:5.7,1);
+\draw (axis cs:-0.3,1) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
   text=black,
@@
 ]{1.0};
-\draw (axis cs:1.7,1.02) ++(0pt,1pt) node[
-  scale=0.347,
-  anchor=south,
+\draw (axis cs:0.7,1.01) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
   text=black,
   rotate=0.0
-]{1.02};
-\draw (axis cs:2.7,0.98) ++(0pt,1pt) node[
-  scale=0.347,
-  anchor=south,
+]{1.01};
+\draw (axis cs:1.7,1.3) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
   text=black,
   rotate=0.0
-]{0.98};
-\draw (axis cs:3.7,0.98) ++(0pt,1pt) node[
-  scale=0.347,
-  anchor=south,
+]{1.3};
+\draw (axis cs:2.7,0.78) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
   text=black,
   rotate=0.0
-]{0.98};
-\draw (axis cs:4.7,0.97) ++(0pt,1pt) node[
-  scale=0.347,
-  anchor=south,
+]{0.78};
+\draw (axis cs:3.7,0.97) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
   text=black,
@@
 ]{0.97};
+\draw (axis cs:4.7,1.15) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{1.15};
+\draw (axis cs:5.7,0.85) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{0.85};
+\draw (axis cs:-0.5,1) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{1.0};
+\draw (axis cs:0.5,1) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{1.0};
+\draw (axis cs:1.5,1.42) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{1.42};
+\draw (axis cs:2.5,0.96) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{0.96};
+\draw (axis cs:3.5,1.12) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{1.12};
+\draw (axis cs:4.5,1.01) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{1.01};
+\draw (axis cs:5.5,1) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{1.0};
 \end{axis}

### 2021-10-17 test/lambdapure/compile/bench/speedup-rgn.py

@@ -32,13 +32,14 @@ G_BASELINE = "../baseline-lean.sh"
 G_OURS = "../run-lean.sh"
 G_FPATHS = []
 G_FPATHS.append(("binarytrees-int.lean", 20))
-# G_FPATHS.append(("binarytrees.lean", 20))
-# G_FPATHS.append("const_fold.lean")
-# G_FPATHS.append("deriv.lean")
-# G_FPATHS.append(("filter.lean", 50000))
-# G_FPATHS.append("qsort.lean") # miscompile because of jmp!
-# G_FPATHS.append("rbmap_checkpoint.lean")
-# G_FPATHS.append("unionfind.lean")
+G_FPATHS.append(("binarytrees.lean", 20))
+# G_FPATHS.append(("const_fold.lean", 10))
+G_FPATHS.append(("const_fold.lean", 9)) # 10 miscompiles!
+G_FPATHS.append(("deriv.lean", 4))
+G_FPATHS.append(("filter.lean", 50000))
+# G_FPATHS.append(("qsort.lean", 100)) # miscompile because of jmp!
+G_FPATHS.append(("rbmap_checkpoint.lean", 400000))
+G_FPATHS.append(("unionfind.lean", 100000))
 G_NFILES = len(G_FPATHS)
 
 
@@ -111,6 +112,8 @@ def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fp
     datum[out_index] = []
     datum[perf_index] = []
     for _ in range(ARGS.nruns):
+      print(f"\t$ # {fpath}")
+      os_system_synch(f"./exe-ref.out {run_args}") # run once for correctness / check that stuff doesn't crash.
       out, perf = sh(f"perf stat ./exe-ref.out {run_args}")
       datum[out_index].append(out)
       datum[perf_index].append(perf)
@@ -282,13 +285,14 @@ def perf_stat_to_time(file, o):
 def autolabel(ax, rects):
     """Attach a text label above each bar in *rects*, displaying its height."""
     for rect in rects:
+        x0, y0 = rect.xy # bottom left
         height = rect.get_height()
         ax.annotate('{}'.format(height),
-                    xy=(rect.get_x(), height),
-                    xytext=(0, 1),  # 1 points vertical offset
+                    xy=(rect.get_x(), y0 + height),
+                    xytext=(0, 5),  # 1 points vertical offset
                     textcoords="offset points",
-                    fontsize="x-small",
-                    ha='center', va='bottom')
[…]

### 2021-10-18 test/lambdapure/compile/bench/speedup-rgn.tex
show that programs look different, but performance is the same

@@
 ylabel={Runtimes},
-ymin=0, ymax=1.491,
+ymin=0, ymax=1.1865,
 ytick style={color=black}
@@
 
-\draw[draw=none,fill=color0] (axis cs:0.7,0) rectangle (axis cs:0.9,1.01);
-\draw[draw=none,fill=color0] (axis cs:1.7,0) rectangle (axis cs:1.9,1.3);
-\draw[draw=none,fill=color0] (axis cs:2.7,0) rectangle (axis cs:2.9,0.78);
-\draw[draw=none,fill=color0] (axis cs:3.7,0) rectangle (axis cs:3.9,0.97);
-\draw[draw=none,fill=color0] (axis cs:4.7,0) rectangle (axis cs:4.9,1.15);
-\draw[draw=none,fill=color0] (axis cs:5.7,0) rectangle (axis cs:5.9,0.85);
+\draw[draw=none,fill=color0] (axis cs:0.7,0) rectangle (axis cs:0.9,1);
+\draw[draw=none,fill=color0] (axis cs:1.7,0) rectangle (axis cs:1.9,1.01);
+\draw[draw=none,fill=color0] (axis cs:2.7,0) rectangle (axis cs:2.9,1);
+\draw[draw=none,fill=color0] (axis cs:3.7,0) rectangle (axis cs:3.9,1.03);
+\draw[draw=none,fill=color0] (axis cs:4.7,0) rectangle (axis cs:4.9,1.13);
+\draw[draw=none,fill=color0] (axis cs:5.7,0) rectangle (axis cs:5.9,0.86);
 \draw[draw=none,fill=color1] (axis cs:-0.5,0) rectangle (axis cs:-0.3,1);
@@
 \draw[draw=none,fill=color1] (axis cs:0.5,0) rectangle (axis cs:0.7,1);
-\draw[draw=none,fill=color1] (axis cs:1.5,0) rectangle (axis cs:1.7,1.42);
-\draw[draw=none,fill=color1] (axis cs:2.5,0) rectangle (axis cs:2.7,0.96);
-\draw[draw=none,fill=color1] (axis cs:3.5,0) rectangle (axis cs:3.7,1.12);
-\draw[draw=none,fill=color1] (axis cs:4.5,0) rectangle (axis cs:4.7,1.01);
-\draw[draw=none,fill=color1] (axis cs:5.5,0) rectangle (axis cs:5.7,1);
+\draw[draw=none,fill=color1] (axis cs:1.5,0) rectangle (axis cs:1.7,1);
+\draw[draw=none,fill=color1] (axis cs:2.5,0) rectangle (axis cs:2.7,1);
+\draw[draw=none,fill=color1] (axis cs:3.5,0) rectangle (axis cs:3.7,1.01);
+\draw[draw=none,fill=color1] (axis cs:4.5,0) rectangle (axis cs:4.7,0.98);
+\draw[draw=none,fill=color1] (axis cs:5.5,0) rectangle (axis cs:5.7,0.99);
 \draw (axis cs:-0.3,1) ++(0pt,5pt) node[
@@
 ]{1.0};
-\draw (axis cs:0.7,1.01) ++(0pt,5pt) node[
+\draw (axis cs:0.7,1) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{1.0};
+\draw (axis cs:1.7,1.01) ++(0pt,5pt) node[
   scale=0.2895,
@@
 ]{1.01};
-\draw (axis cs:1.7,1.3) ++(0pt,5pt) node[
+\draw (axis cs:2.7,1) ++(0pt,5pt) node[
   scale=0.2895,
@@
   rotate=0.0
-]{1.3};
-\draw (axis cs:2.7,0.78) ++(0pt,5pt) node[
+]{1.0};
+\draw (axis cs:3.7,1.03) ++(0pt,5pt) node[
   scale=0.2895,
@@
   rotate=0.0
-]{0.78};
-\draw (axis cs:3.7,0.97) ++(0pt,5pt) node[
+]{1.03};
+\draw (axis cs:4.7,1.13) ++(0pt,5pt) node[
   scale=0.2895,
@@
   rotate=0.0
-]{0.97};
-\draw (axis cs:4.7,1.15) ++(0pt,5pt) node[
+]{1.13};
+\draw (axis cs:5.7,0.86) ++(0pt,5pt) node[
   scale=0.2895,
@@
   rotate=0.0
-]{1.15};
-\draw (axis cs:5.7,0.85) ++(0pt,5pt) node[
-  scale=0.2895,
-  anchor=north west,
-  text=black,
-  rotate=0.0
-]{0.85};
+]{0.86};
 \draw (axis cs:-0.5,1) ++(0pt,5pt) node[
@@
 ]{1.0};
-\draw (axis cs:1.5,1.42) ++(0pt,5pt) node[
+\draw (axis cs:1.5,1) ++(0pt,5pt) node[
   scale=0.2895,
@@
   rotate=0.0
-]{1.42};
-\draw (axis cs:2.5,0.96) ++(0pt,5pt) node[
+]{1.0};
+\draw (axis cs:2.5,1) ++(0pt,5pt) node[
   scale=0.2895,
@@
   rotate=0.0
-]{0.96};
-\draw (axis cs:3.5,1.12) ++(0pt,5pt) node[
-  scale=0.2895,
-  anchor=north west,
-  text=black,
-  rotate=0.0
-]{1.12};
-\draw (axis cs:4.5,1.01) ++(0pt,5pt) node[
+]{1.0};
+\draw (axis cs:3.5,1.01) ++(0pt,5pt) node[
   scale=0.2895,
@@
 ]{1.01};
-\draw (axis cs:5.5,1) ++(0pt,5pt) node[
+\draw (axis cs:4.5,0.98) ++(0pt,5pt) node[
   scale=0.2895,
@@
   rotate=0.0
-]{1.0};
+]{0.98};
+\draw (axis cs:5.5,0.99) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{0.99};
 \end{axis}

### 2021-10-18 test/lambdapure/compile/bench/speedup-rgn.py

@@ -33,11 +33,9 @@ G_OURS = "../run-lean.sh"
 G_FPATHS = []
 G_FPATHS.append(("binarytrees-int.lean", 20))
 G_FPATHS.append(("binarytrees.lean", 20))
-# G_FPATHS.append(("const_fold.lean", 10))
 G_FPATHS.append(("const_fold.lean", 9)) # 10 miscompiles!
 G_FPATHS.append(("deriv.lean", 4))
 G_FPATHS.append(("filter.lean", 50000))
-# G_FPATHS.append(("qsort.lean", 100)) # miscompile because of jmp!
 G_FPATHS.append(("rbmap_checkpoint.lean", 400000))
 G_FPATHS.append(("unionfind.lean", 100000))
 G_NFILES = len(G_FPATHS)
@@ -170,6 +168,10 @@ def run_data():
             fpath=fpath, 
             run_args=str(problemsize),
             out_index="none-out", perf_index="none-perf"))
+        ds.append(datum)
+        
+        assert datum["ours-out"] == datum["theirs-out"]
+        assert datum["none-out"] == datum["theirs-out"]
 
         # disabled simpcase + MLIR.rgn optimization passes: 55a63f500b23b8c0c180e43108c5f844839a693f
         # NO simpcase (DISABLED): 850fd84e43407ed647837652b6442e143199abb0
@@ -242,7 +244,6 @@ def run_data():
         #    out, perf = sh("perf stat ./exe-mlir.out")
         #    datum["none-out"].append(out)
         #    datum["none-perf"].append(perf)
-        ds.append(datum)
     with open(ARGS.out, "w") as f:
         json.dump(ds, f, indent=2)
     return ds

### 2021-10-18 test/lambdapure/compile/bench/speedup-rgn.tex
add speedup with delta to see WTF is happening

@@
 ylabel={Runtimes},
-ymin=0, ymax=1.1865,
+ymin=0, ymax=1.197,
 ytick style={color=black}
@@
 \draw[draw=none,fill=color0] (axis cs:0.7,0) rectangle (axis cs:0.9,1);
-\draw[draw=none,fill=color0] (axis cs:1.7,0) rectangle (axis cs:1.9,1.01);
-\draw[draw=none,fill=color0] (axis cs:2.7,0) rectangle (axis cs:2.9,1);
-\draw[draw=none,fill=color0] (axis cs:3.7,0) rectangle (axis cs:3.9,1.03);
-\draw[draw=none,fill=color0] (axis cs:4.7,0) rectangle (axis cs:4.9,1.13);
+\draw[draw=none,fill=color0] (axis cs:1.7,0) rectangle (axis cs:1.9,0.88);
+\draw[draw=none,fill=color0] (axis cs:2.7,0) rectangle (axis cs:2.9,0.77);
+\draw[draw=none,fill=color0] (axis cs:3.7,0) rectangle (axis cs:3.9,1.07);
+\draw[draw=none,fill=color0] (axis cs:4.7,0) rectangle (axis cs:4.9,1.14);
 \draw[draw=none,fill=color0] (axis cs:5.7,0) rectangle (axis cs:5.9,0.86);
@@
 \draw[draw=none,fill=color1] (axis cs:0.5,0) rectangle (axis cs:0.7,1);
-\draw[draw=none,fill=color1] (axis cs:1.5,0) rectangle (axis cs:1.7,1);
-\draw[draw=none,fill=color1] (axis cs:2.5,0) rectangle (axis cs:2.7,1);
-\draw[draw=none,fill=color1] (axis cs:3.5,0) rectangle (axis cs:3.7,1.01);
-\draw[draw=none,fill=color1] (axis cs:4.5,0) rectangle (axis cs:4.7,0.98);
-\draw[draw=none,fill=color1] (axis cs:5.5,0) rectangle (axis cs:5.7,0.99);
+\draw[draw=none,fill=color1] (axis cs:1.5,0) rectangle (axis cs:1.7,0.96);
+\draw[draw=none,fill=color1] (axis cs:2.5,0) rectangle (axis cs:2.7,0.93);
+\draw[draw=none,fill=color1] (axis cs:3.5,0) rectangle (axis cs:3.7,0.89);
+\draw[draw=none,fill=color1] (axis cs:4.5,0) rectangle (axis cs:4.7,1);
+\draw[draw=none,fill=color1] (axis cs:5.5,0) rectangle (axis cs:5.7,1);
 \draw (axis cs:-0.3,1) ++(0pt,5pt) node[
@@
 ]{1.0};
-\draw (axis cs:1.7,1.01) ++(0pt,5pt) node[
+\draw (axis cs:1.7,0.88) ++(0pt,5pt) node[
   scale=0.2895,
@@
   rotate=0.0
-]{1.01};
-\draw (axis cs:2.7,1) ++(0pt,5pt) node[
+]{0.88};
+\draw (axis cs:2.7,0.77) ++(0pt,5pt) node[
   scale=0.2895,
@@
   rotate=0.0
-]{1.0};
-\draw (axis cs:3.7,1.03) ++(0pt,5pt) node[
+]{0.77};
+\draw (axis cs:3.7,1.07) ++(0pt,5pt) node[
   scale=0.2895,
@@
   rotate=0.0
-]{1.03};
-\draw (axis cs:4.7,1.13) ++(0pt,5pt) node[
+]{1.07};
+\draw (axis cs:4.7,1.14) ++(0pt,5pt) node[
   scale=0.2895,
@@
   rotate=0.0
-]{1.13};
+]{1.14};
 \draw (axis cs:5.7,0.86) ++(0pt,5pt) node[
@@
 ]{1.0};
-\draw (axis cs:1.5,1) ++(0pt,5pt) node[
+\draw (axis cs:1.5,0.96) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{0.96};
+\draw (axis cs:2.5,0.93) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{0.93};
+\draw (axis cs:3.5,0.89) ++(0pt,5pt) node[
+  scale=0.2895,
+  anchor=north west,
+  text=black,
+  rotate=0.0
+]{0.89};
+\draw (axis cs:4.5,1) ++(0pt,5pt) node[
   scale=0.2895,
@@
 ]{1.0};
-\draw (axis cs:2.5,1) ++(0pt,5pt) node[
+\draw (axis cs:5.5,1) ++(0pt,5pt) node[
   scale=0.2895,
@@
 ]{1.0};
-\draw (axis cs:3.5,1.01) ++(0pt,5pt) node[
-  scale=0.2895,
-  anchor=north west,
-  text=black,
-  rotate=0.0
-]{1.01};
-\draw (axis cs:4.5,0.98) ++(0pt,5pt) node[
-  scale=0.2895,
-  anchor=north west,
-  text=black,
-  rotate=0.0
-]{0.98};
-\draw (axis cs:5.5,0.99) ++(0pt,5pt) node[
-  scale=0.2895,
-  anchor=north west,
-  text=black,
-  rotate=0.0
-]{0.99};
 \end{axis}

### 2021-10-18 test/lambdapure/compile/bench/speedup-rgn.py

@@ -14,6 +14,7 @@ import numpy as np
 import json
 import sys
 import subprocess
+import difflib
 import numpy as np
 from scipy.stats import mstats
 import tikzplotlib
@@ -68,7 +69,11 @@ def sh(path):
     return (output.decode(), err.decode())
 
 # returns datum
-def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fpath, run_args, out_index, perf_index):
+def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fpath, run_args, ll_index, mlir_index, out_index, perf_index):
+    datum = {}
+    datum["file"] = fpath
+    datum[out_index] = []
+    datum[perf_index] = []
     print(f"running f{fpath} case_simpl_enabled={case_simpl_enabled} rgn_optimization_enabled={rgn_optimization_enabled}")
     # simpcase ENABLED: 850fd84e43407ed647837652b6442e143199abb0
     LEAN_PATH="/home/bollu/work/lean4/build/release/stage1/bin/lean"
@@ -77,8 +82,13 @@ def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fp
     else:
         lean_opts = "-Dcompiler.caseSimpl=false"
 
-    os_system_synch(f"{LEAN_PATH} {fpath} {lean_opts}  -m exe.mlir")
+    os_system_synch("rm exe.mlir || true")
+    os_system_synch(f"{LEAN_PATH} {fpath} {lean_opts}  -m exe-raw.mlir")
+    os_system_synch(f"hask-opt exe-raw.mlir -o exe.mlir")
+    with open("exe.mlir", "r") as f:
+        datum[mlir_index] = f.readlines()
 
+    os_system_synch("rm exe.ll || true")
     if False:
         os_system_synch("hask-opt exe.mlir --convert-scf-to-std --lean-lower-rgn  --rgn-cse --cse --convert-rgn-to-std --convert-std-to-llvm --ptr-lower | \
                mlir-translate --mlir-to-llvmir -o exe.ll")
@@ -86,6 +96,10 @@ def compile_and_run_with_option(case_simpl_enabled, rgn_optimization_enabled, fp
         os_system_synch("hask-opt exe.mlir --convert-scf-to-std --lean-lower-rgn --convert-rgn-to-std --convert-std-to-llvm --ptr-lower | \
[…]

### 2021-10-20 include/Hask/HaskOps.h
fixed bug

@@ -597,10 +597,12 @@ public:
                     mlir::Value v, int count);
 
   bool isCheckRef() {
+    assert(this->getOperation()->hasAttr("checkref"));
     return this->getOperation()->getAttrOfType<BoolAttr>("checkref").getValue();
   }
 
   int getIncCount() {
+    assert(this->getOperation()->hasAttr("value"));
     return this->getOperation()->getAttrOfType<IntegerAttr>("value").getInt();
   }
 };

### 2021-10-20 lib/Hask/LeanLowering.cpp

@@ -2261,13 +2261,17 @@ public:
       llvm::SmallVector<Value, 1> args = {rands[0]};
       rewriter.replaceOpWithNewOp<CallOp>(inc, f, args);
     } else {
-      assert(false && "unhandled lean_inc with n > 1");
+      // assert(false && "unhandled lean_inc with n > 1");
       // n > 1
       FuncOp f = inc.isCheckRef() ? getOrCreateLeanIncN(rewriter, mod)
                                   : getOrCreateLeanIncRefN(rewriter, mod);
       const int WIDTH = 32;
+
+      llvm::errs() << "found createLeanIncN: "; 
+      llvm::errs() << inc << "\n";
       ConstantIntOp n = rewriter.create<ConstantIntOp>(
           inc.getLoc(), inc.getIncCount(), WIDTH);
+      llvm::errs() << "inc constant int: " << n << "\n";
       llvm::SmallVector<Value, 2> args = {rands[0], n};
       rewriter.replaceOpWithNewOp<CallOp>(inc, f, args);
     }

### 2021-10-20 lean-linking-incantations/lean-shell.c
fixup bug in expr.lean; call lean_initialize to initialize the debug printing system

@@ -27,6 +27,7 @@ lean_object *initialize_Init(lean_object *); // forward declaration
 lean_object *init_lean_custom_entrypoint_hack(lean_object*);
 lean_object *main_lean_custom_entrypoint_hack(lean_object *argv, lean_object *world);
 void lean_initialize_runtime_module();
+void lean_initialize();
 
 #if defined(WIN32) || defined(_WIN32)
 #include <windows.h>
@@ -38,6 +39,7 @@ int main(int argc, char **argv) {
 #endif
   lean_object *in;
   lean_object *res;
+  lean_initialize();
   lean_initialize_runtime_module();
   // lean_init_task_manager();
 
